\section{Positional encoding}
\label{sec:positional-encoding}

Self-attention is permutation-equivariant: with no positional signal the
output for $(x_1, \ldots, x_S)$ is invariant under any permutation of
positions, so the model treats ``dog bites man'' and ``man bites dog''
identically~\citep{vaswani2017}. Positional encoding is the mechanism
that breaks this symmetry, and the family of choices is one of the
small number of axes along which the modern Transformer is
parameterized. The thesis of this section: between 2017 and 2026 the
field walked sinusoidal $\to$ learned $\to$ T5 relative bias $\to$
ALiBi $\to$ RoPE, RoPE became dominant for decoder-only LLMs, and the
practical question of long-context extension is now \emph{primarily a
positional question} (NTK/YaRN/LongRoPE), not (only) an attention-
implementation question. We close with NoPE and iRoPE, where the field
asks whether explicit positional encoding is needed at all once depth
and causal masking are in place.

\subsection{The 2017 sinusoidal baseline}
\label{sec:pos-sinusoidal}

\citet{vaswani2017} adds a deterministic positional vector
$\boldsymbol{p}_m \in \R^{D}$ to the token embedding at the input of
the encoder/decoder stack, before any attention layer:
\begin{equation}
f_{\{q,k\}}(\boldsymbol{x}_m, m) = \boldsymbol{W}_{\{q,k\}}\,
   (\boldsymbol{x}_m + \boldsymbol{p}_m),
\label{eq:pe-additive}
\end{equation}
with the sinusoidal choice~\citep[\S 3.5]{vaswani2017}%
\footnote{KB excerpt: \texttt{kb/excerpts/vaswani2017.md\#sec-3-5}.}
\begin{equation}
\begin{aligned}
PE_{(pos,\,2i)}     &= \sin\!\bigl(pos\,/\,10000^{2i/d_{\text{model}}}\bigr), \\
PE_{(pos,\,2i+1)}   &= \cos\!\bigl(pos\,/\,10000^{2i/d_{\text{model}}}\bigr),
\end{aligned}
\label{eq:pe-sinusoidal}
\end{equation}
where $pos$ is the token position and $i$ indexes dimension. The
wavelengths form a geometric progression from $2\pi$ to
$10000 \cdot 2\pi$. \citet{vaswani2017} motivate the closed form on two
grounds: a fixed offset $k$ admits a linear map taking $PE_{pos}$ to
$PE_{pos+k}$, ``[allowing] the model to easily learn to attend by
relative positions''; and the closed form ``may allow the model to
extrapolate to sequence lengths longer than the ones encountered during
training''~\citep[\S 3.5]{vaswani2017}.

The same paper notes that learned absolute embeddings produce
``nearly identical results'' on WMT~\citep[Table 3 row (E)]{vaswani2017}
--- so at the original $d_{\text{model}}=512$, $S \le 100$ scale,
fixed-vs-learned was a non-issue. The fixed/learned dichotomy only
becomes load-bearing once $S$ grows past the training horizon, which is
precisely the regime where every successor encoding (T5 bias, ALiBi,
RoPE) is designed to behave gracefully.

\begin{intuition}
The sinusoidal formula is the closest to a ``coordinate system'' the
input can carry. Each pair of dimensions $(2i, 2i{+}1)$ traces a
unit-circle hand at angular rate $\theta_i = 10000^{-2i/d_{\text{model}}}$.
High-frequency hands ($i \to 0$) discriminate local order; low-frequency
hands ($i \to d/2$) carry coarse global position. Returning to the
canonical form: equation~\eqref{eq:pe-sinusoidal} is exactly the
$(\sin, \cos)$ projection of those rotating hands at position $pos$.
\end{intuition}

\subsection{Learned absolute, T5 relative bias, and ALiBi}
\label{sec:pos-learned-and-bias}

Three intermediate families followed the sinusoidal baseline.

\paragraph{Learned absolute (BERT, GPT-2).} Replace the fixed
$\boldsymbol{p}_m$ in~\eqref{eq:pe-additive} with a learned embedding
indexed by $m \in \{0, \ldots, S_{\text{train}}-1\}$~\citep{devlin2019-bert,
radford2019-gpt2}. The cost: a hard cutoff at the training context
$S_{\text{train}}$. Inference at $m \ge S_{\text{train}}$ has no defined
embedding, and there is no closed-form way to extend the table.

\paragraph{T5 relative bias.} \citet{raffel2019} replace the additive
input-side encoding with a per-layer scalar bias on the attention
logit, binned by signed relative distance:
\begin{equation}
\boldsymbol{q}_m^\top \boldsymbol{k}_n
= \boldsymbol{x}_m^\top \boldsymbol{W}_q^\top \boldsymbol{W}_k \boldsymbol{x}_n
  + b_{m-n}^{(h)},
\label{eq:t5-bias}
\end{equation}
with one $b_{\Delta}^{(h)}$ table per head $h$. The bias is added inside
each attention layer rather than once at the input, which avoids the
``diluted-by-depth'' degradation of additive input
encoding~\citep{raffel2019}.

\paragraph{ALiBi.} \citet[ALiBi]{press-alibi} make the bias
parameter-free and \emph{linear in distance}:
$b_{m-n}^{(h)} = -|m-n| \cdot s_h$ with a fixed per-head slope $s_h$,
producing a built-in linear penalty for distance and --- the load-
bearing claim --- length extrapolation past $S_{\text{train}}$ without
fine-tuning.\deepencite{press-alibi §3 (excerpt not yet in KB; cite
``Train Short, Test Long: Attention with Linear Biases Enables Input
Length Extrapolation'', Press et al.\ 2022)}

All three lost ground to RoPE for decoder-only LLMs by 2022--2023.
Learned absolute lost on extrapolation; T5 bias's binned table did not
generalize cleanly across context lengths; ALiBi's linear slope is the
right shape for some workloads but not for the long-range structure
attention is asked to handle in modern decoder-only training.

\subsection{RoPE: rotation in the QK inner product}
\label{sec:pos-rope}

\citet{su2021} replace the additive encoding with a multiplicative
rotation applied to $Q$ and $K$ inside attention.\footnote{KB excerpt:
\texttt{kb/excerpts/su2021.md\#sec-3-2-2}.} For a residual-stream
activation $\boldsymbol{x}_m \in \R^{d_h}$ at position $m$ (per head,
with $d_h$ even):
\begin{equation}
f_{\{q,k\}}(\boldsymbol{x}_m, m)
= \boldsymbol{R}^{d_h}_{\Theta, m}\, \boldsymbol{W}_{\{q,k\}}\, \boldsymbol{x}_m,
\label{eq:rope}
\end{equation}
where $\boldsymbol{R}^{d_h}_{\Theta, m}$ is the block-diagonal
orthogonal rotation matrix with $d_h/2$ blocks of the form
\begin{equation}
\begin{pmatrix}\cos m\theta_i & -\sin m\theta_i \\
               \sin m\theta_i & \phantom{-}\cos m\theta_i\end{pmatrix},
\quad i = 1, \ldots, d_h/2,
\label{eq:rope-block}
\end{equation}
and the frequencies form a geometric progression with base $b$:
\begin{equation}
\Theta = \bigl\{\theta_i = b^{-2(i-1)/d_h}\bigr\}_{i=1}^{d_h/2},
\quad b = 10000\ \text{(default)}
\label{eq:rope-theta}
\end{equation}
\citep[\S 3.2.2 Eq.\ 14--15]{su2021}. The critical property follows
from orthogonality: substituting~\eqref{eq:rope}
into the attention dot product gives
\begin{equation}
\boldsymbol{q}_m^\top \boldsymbol{k}_n
= \boldsymbol{x}_m^\top \boldsymbol{W}_q^\top
  \boldsymbol{R}^{d_h}_{\Theta, n - m} \boldsymbol{W}_k
  \boldsymbol{x}_n,
\label{eq:rope-relative}
\end{equation}
\citep[\S 3.2.2 Eq.\ 16]{su2021}. Position information enters the
attention score \emph{only as the relative offset $n-m$} --- when
$m = n$ the rotation is the identity and the inner product is content-
only; otherwise it depends on $(n-m)$ but not on $m$ or $n$ separately.

In practice the rotation is computed elementwise rather than as a
matmul, exploiting the sparsity of $\boldsymbol{R}^{d_h}_{\Theta, m}$:
\begin{equation}
\boldsymbol{R}^{d_h}_{\Theta, m}\, \boldsymbol{x}
= \boldsymbol{x} \otimes \cos(m\boldsymbol{\theta})
  + \mathrm{rotate\_half}(\boldsymbol{x}) \otimes \sin(m\boldsymbol{\theta}),
\label{eq:rope-pair}
\end{equation}
where $\boldsymbol{\theta}$ duplicates each $\theta_i$ to length $d_h$
and $\mathrm{rotate\_half}$ swaps each pair
$(x_{2i-1}, x_{2i}) \mapsto (-x_{2i}, x_{2i-1})$~\citep[\S 3.4.2 Eq.\ 34]{su2021}.
This is the form behind the
\texttt{apply\_rotary\_pos\_emb(q, k, cos, sin)} helper in HF
transformers and the \texttt{ggml\_rope\_*} family in \texttt{llama.cpp};
two element-wise multiplies and one add per query/key, negligible
overhead.

Three properties make RoPE the dominant choice in 2022--2026 frontier
decoder-only LLMs (LLaMA 1--3, Mistral, Qwen, DeepSeek, Gemma, OLMo).
First, the relative-only dependence~\eqref{eq:rope-relative} is
arithmetic, not learned --- $\boldsymbol{R}^{d_h}_{\Theta, m}$ is
defined for any $m \in \mathbb{N}$ without indexing a learned table.
Second, applying the rotation \emph{at every layer on $Q, K$ only}
preserves the residual stream's content basis. Third, the
long-term-decay property~\citep[\S 3.3]{su2021}: at fixed $\Theta$, the
inner product $\langle f_q(\cdot, m), f_k(\cdot, n) \rangle$ tends to
decay as $|m-n|$ grows because rotations at different frequencies
dephase, matching the prior that distant tokens should interact less.

\begin{analogy}
RoPE is a clock with $d_h/2$ hands, each ticking at a different rate.
At position $m$, hand $i$ points to angle $m\theta_i$. Query $\boldsymbol{q}_m$
and key $\boldsymbol{k}_n$ ``agree'' on dimension-pair $i$ when their
hands align modulo $2\pi$ --- when $(m-n)\theta_i \equiv 0$. Returning
to canonical form: hand alignment is exactly the cosine factor
$\cos((m-n)\theta_i)$ that emerges from expanding
$\boldsymbol{x}_m^\top \boldsymbol{R}^{d_h}_{\Theta, n-m} \boldsymbol{x}_n$
in the canonical $2{\times}2$ block basis of~\eqref{eq:rope-block}.
\end{analogy}

\subsection{Long-context extension: NTK-aware, YaRN, LongRoPE}
\label{sec:pos-long-context}

The pre-trained context $S_{\text{train}}$ is set by training data and
compute. Deploying at $S_{\text{infer}} > S_{\text{train}}$ requires
some adjustment because positions $m > S_{\text{train}}$ produce
rotations $m\theta_i$ that are out-of-distribution: the model has
never observed these phases on $Q, K$ during training, and attention
features may not generalize. The extension methods are the practical
positional question for the 128K-and-up regime.

\paragraph{Position interpolation (PI).} Chen et al.\ 2023 propose,
for ratio $s = S_{\text{infer}} / S_{\text{train}} > 1$, replacing
$m\theta_i$ with $(m/s)\theta_i$ in~\eqref{eq:rope-block}, mapping the
new range into the training-seen
range.\deepencite{chen2023-position-interpolation §3 (excerpt not yet
in KB; cite ``Extending Context Window of Large Language Models via
Positional Interpolation'', Chen et al.\ 2023, arXiv 2306.15595)} Cheap
but degrades short-context quality because it compresses the
high-frequency dimensions.

\paragraph{NTK-aware scaling.} The insight: high-frequency dimensions
($\theta_i$ small $i$) execute many full rotations within $S_{\text{train}}$
and should not be interpolated --- compressing them harms local
fidelity. Low-frequency dimensions execute less than one rotation in
$S_{\text{train}}$ and benefit from interpolation. The NTK-aware fix
scales the \emph{base} $b$ instead of the positions:
\begin{equation}
b' = b \cdot s^{d_h / (d_h - 2)},
\label{eq:ntk-base}
\end{equation}
which compresses primarily the low-frequency dimensions while leaving
high-frequency ones nearly intact~\citep[\S 3.2,
\texttt{kb/notes/architecture/position-encoding.md\#sec-3-2}]{su2021}.

\paragraph{YaRN.} \deepencite{peng2023-yarn (excerpt not yet in KB;
cite ``YaRN: Efficient Context Window Extension of Large Language
Models'', Peng et al.\ 2023, arXiv 2309.00071)} YaRN combines
NTK-aware base-scaling with a per-frequency interpolation
schedule: dimensions with wavelength shorter than the training context
use no scaling, dimensions longer than the extended context use full
PI, intermediate dimensions use a smooth ramp. YaRN additionally
rescales attention logits by $1/\sqrt{t}$ where $t$ depends on the
extension factor $s$:
\begin{equation}
\softmax\!\Bigl(\frac{Q K^\top}{\sqrt{d_h}\,\sqrt{t}}\Bigr) V,
\quad \sqrt{t} = 0.1 \ln(s) + 1,
\label{eq:yarn-temp}
\end{equation}
preserving the softmax temperature regime as the effective sequence
length grows. YaRN reports roughly $10\times$ fewer fine-tuning tokens
than plain PI to extend LLaMA-2 to 128K (note: equations
\eqref{eq:ntk-base},~\eqref{eq:yarn-temp} are reconstructed from the
LLaMA-3 implementation and secondary references; see the deepen-cite
above for primary verification).

\paragraph{LongRoPE / LongRoPE2.} LongRoPE introduces evolutionary
search over per-frequency scaling factors --- a non-uniform rescaling
that no closed-form method can produce, reaching 2M-token context.
\citet{longrope2-2025} extend this to a per-position search and report
near-lossless extension (98.5\% of short-context performance preserved)
at $80\times$ fewer tokens than Meta's earlier LLaMA-3 long-context
fine-tuning approach.\deepencite{longrope2-2025 §3 (KB excerpt
abstract-only as of 2026-05-05; equations and the per-position search
schedule still need transcription)}

The pattern across PI $\to$ NTK $\to$ YaRN $\to$ LongRoPE is one of
\emph{increasingly fine-grained per-frequency control over $\theta_i$}.
Each step preserves the multiplicative structure
of~\eqref{eq:rope}; only the rule for which $\theta_i$ get rescaled
and by how much changes. Long-context extension is, in this lineage, a
positional question --- a choice of frequency schedule --- not a
choice of attention implementation. The attention kernel
(FlashAttention, NSA; \S 6 of this paper) is unchanged.

\subsection{NoPE and iRoPE}
\label{sec:pos-nope}

\deepencite{kazemnejad2023-nope (excerpt not yet in KB; cite ``The
Impact of Positional Encoding on Length Generalization in Transformers'',
Kazemnejad et al.\ 2023, NeurIPS)} The NoPE result (Kazemnejad et al.\
2023) is that decoder-only Transformers trained \emph{without any
positional encoding} can generalize as well or better than RoPE/ALiBi
to longer sequences on synthetic length-generalization tasks. The
mechanism is depth-induced: causal masking gives every position a
unique attention pattern (position $m$ has $m+1$ attendable tokens, so
position 0 is uniquely identifiable as the only one with empty
attention context), and stacked layers can recover positional
information from this content-mask interaction without an explicit
embedding.

\citet{meta-llama4} apply this in production with iRoPE: alternate
RoPE-equipped attention layers with NoPE attention layers in a 3:1
ratio, with a temperature-scaling factor in the NoPE layers. Trained
natively at $S_{\text{train}} = 256\text{K}$, LLaMA 4 Scout reports a
claimed 10M-token inference context --- a qualitative regime change,
not just better interpolation. The pattern resembles a fallback: RoPE
layers provide explicit local-order constraints; NoPE layers are
forced to derive position from content alone, providing a channel that
does not depend on the rotation when phases at $m \gg S_{\text{train}}$
go out-of-distribution.

\begin{contradiction}
Long-context extrapolation method comparison is contested as of
2026-05. Three competing positions: (a) RoPE-only with NTK-aware base
scaling is sufficient for the $\le 4\times$ extension regime (e.g.\
LLaMA-3's bump from 8K to 128K); (b) YaRN with a per-frequency
schedule is required for the $\sim 10\times$ regime
(\deepencite{peng2023-yarn}); (c) per-position search (LongRoPE2) and
mixed RoPE/NoPE layers (iRoPE) are required for the $\ge 40\times$
regime~\citep{longrope2-2025, meta-llama4}. The benchmarks backing
each are largely single-source: Meta reports iRoPE at 10M, Microsoft
reports LongRoPE2 at 2M, with no third-party head-to-head at frontier
scale. NIAH-style retrieval saturates near 100\% at all of these
context lengths; harder evaluations (RULER, HELMET) show steeper and
methodologically inconsistent degradation patterns past 32K, making
``which extension method is best'' contingent on which long-context
benchmark is treated as ground truth (see
\texttt{kb/notes/architecture/long-context.md\#sec-5}).
\end{contradiction}

\subsection{Summary and forward link}
\label{sec:pos-summary}

The 2017--2026 lineage along the positional-encoding axis is
sinusoidal $\to$ learned absolute $\to$ T5 relative bias $\to$ ALiBi
$\to$ RoPE $\to$ \{YaRN, LongRoPE, iRoPE, NoPE-hybrids\}, with RoPE as
the convergent baseline since 2022. The decisive change is the move
from \emph{additive at input} to \emph{multiplicative on $Q, K$ at
every layer}: the multiplicative form
of~\eqref{eq:rope}--\eqref{eq:rope-relative} makes the positional
information arithmetic-extensible, which is the property every
post-2023 long-context extension method exploits.

The next axis (\S 5, attention KV-sharing) is largely
\emph{independent} of this one: a model can choose any of
\{MHA, MQA, GQA, MLA\} on the head-sharing axis and any of \{sinusoidal,
RoPE, RoPE+YaRN, iRoPE\} on the position axis without strong cross-
constraints. The one notable interaction is DeepSeek-V2's
\emph{decoupled-RoPE} construction in MLA --- where applying the
rotation to a low-rank latent $K$ projection breaks the cache-absorbed
forward, and a separate small RoPE-only key channel is split out to
preserve it~\citep{deepseek-v2}. We treat that interaction in detail
in the next section.
