% Paper 1, §7 — Feed-forward axis
% Thesis: ReLU → GELU → SwiGLU dense FFN → Switch (top-1) → Mixtral (8 of 8,
% top-2) → DeepSeekMoE (256 fine-grained + 1 shared, top-8). Routing-collapse
% problem and load balancing. Why fine-grained + shared is the current
% convergence.

\section{Feed-forward axis: dense, GLU, MoE, fine-grained MoE}
\label{sec:ffn}
\label{sec:feedforward-axis}
\label{sec:ffn-axis}

The feed-forward sublayer is where the bulk of a \glsterm{transformer}{Transformer}'s
\glsterm{parameters}{parameters} live and, since 2021, where the bulk of \emph{architectural
optimization} has happened. Two axes of choice run in parallel: the
\emph{activation/gating structure} of a single \glsterm{ffn}{FFN} block (\glsterm{relu}{ReLU} $\to$
\glsterm{gelu}{GELU} $\to$ \glsterm{swiglu}{SwiGLU}) and the \emph{multiplicity} of \glsterm{ffn}{FFN} blocks per layer
(dense $\to$ Switch top-1 MoE $\to$ Mixtral top-2 MoE $\to$
DeepSeekMoE fine-grained top-$k$ with a shared expert). The frontier
of 2026 is the Cartesian product: every routed expert is a \glsterm{swiglu}{SwiGLU}
block, and the layer wraps $E$ such experts with a learned router and
a load-balancing discipline. This section walks both axes, transcribes
the load-bearing equations from the primary sources, and surfaces the
single most contested empirical claim --- whether sparse MoE quality
matches dense quality at matched active-parameter count.

\subsection{Dense FFN: the canonical 2017 form}
\label{sec:ffn-vaswani}

The original \glsterm{transformer}{Transformer}'s position-wise feed-forward sublayer is a
two-matrix MLP applied independently at each token position. For input
token vector $\mathbf{x} \in \R^{\mathsf{D}}$,~\citet{vaswani2017}
define%
\footnote{KB excerpt: \texttt{kb/excerpts/shazeer2020.md\#sec-1}; the
2017 form is also transcribed at
\texttt{kb/notes/architecture/ffn.md\#sec-1-1}.}
\begin{equation}
  \mathrm{FFN}(\mathbf{x}, W_1, W_2, b_1, b_2)
  \;=\; \max(0, \mathbf{x} W_1 + b_1)\, W_2 + b_2,
  \tag{Vaswani--Eq.~2}
  \label{eq:vaswani-ffn}
\end{equation}
with $W_1 \in \R^{\mathsf{D} \times d_{\text{ff}}}$,
$W_2 \in \R^{d_{\text{ff}} \times \mathsf{D}}$, and the canonical
inner ratio $d_{\text{ff}} = 4\mathsf{D}$~\citep[\S 3.3]{vaswani2017}.
Two structural features matter. First, $\mathrm{FFN}$ is
\emph{position-wise}: tensor flow is $\BSH \to (\mathsf{B}, \mathsf{S},
d_{\text{ff}}) \to \BSH$, with no inter-position mixing --- all
sequence-axis communication has happened in the attention sublayer of
\cref{sec:attention-kv-sharing}. Second, the \glsterm{ffn}{FFN} is the
\emph{parameter-heaviest} sublayer in a typical block: at $d_{\text{ff}}
= 4\mathsf{D}$ each \glsterm{ffn}{FFN} holds $8\mathsf{D}^2$ \glsterm{parameters}{parameters} versus
$4\mathsf{D}^2$ for the four attention projections.

\citet{radford2019-gpt2} swap \glsterm{relu}{ReLU} for \glsterm{gelu}{GELU} in GPT-2 and drop biases in
the \glsterm{t5}{T5}/\glsterm{decoder-only}{decoder-only} era; \citet{shazeer2020} writes the bias-free form
as~\citep[\S 1, Eq.~2]{shazeer2020}
\begin{equation}
  \mathrm{FFN}_{\mathrm{ReLU}}(\mathbf{x}, W_1, W_2)
  \;=\; \max(\mathbf{x} W_1, 0)\, W_2,
  \tag{Shazeer--Eq.~2}
  \label{eq:relu-ffn}
\end{equation}
\begin{equation}
  \mathrm{FFN}_{\mathrm{GELU}}(\mathbf{x}, W_1, W_2)
  \;=\; \mathrm{GELU}(\mathbf{x} W_1)\, W_2.
  \tag{Shazeer--Eq.~3}
  \label{eq:gelu-ffn}
\end{equation}
The activation swap leaves the parameter count and tensor flow
identical; what changes is the unit-step. \glsterm{relu}{ReLU}'s hard zero cuts off
gradient on negative pre-activations entirely; \glsterm{gelu}{GELU}'s
$x \cdot \Phi(x)$ smooths the cutoff and admits a small negative
contribution near zero, which empirically aids
optimization~\citep[\S 2.3]{radford2019-gpt2}.

\subsection{GLU and SwiGLU: the gated FFN}
\label{sec:ffn-swiglu}

\citet{shazeer2020} reports that swapping the \glsterm{ffn}{FFN} for a Gated Linear
Unit variant yields a small but consistent \glsterm{perplexity}{perplexity} gain on
\glsterm{t5}{T5}-base. The \glsterm{glu}{GLU} layer (Dauphin et al.\ 2016, as cited
by~\citealp[\S 2]{shazeer2020}) is the
component-wise product of two linear projections, one of which is
sigmoid-activated~\citep[\S 2, Eq.~4]{shazeer2020}\footnote{KB
excerpt: \texttt{kb/excerpts/shazeer2020.md\#sec-2}.}:
\begin{equation}
  \mathrm{GLU}(\mathbf{x}, W, V, b, c)
  \;=\; \sigma(\mathbf{x} W + b) \otimes (\mathbf{x} V + c),
  \tag{Shazeer--Eq.~4}
  \label{eq:glu}
\end{equation}
with $W, V \in \R^{\mathsf{D} \times d_{\text{ff}}}$ and $\otimes$
the elementwise (Hadamard) product. The activation in the gate
position is a free choice; \citet[\S 2, Eq.~5]{shazeer2020} lists
five variants%
\footnote{KB excerpt: \texttt{kb/excerpts/shazeer2020.md\#sec-2}.}
including the bilinear ($\sigma \to \mathrm{id}$), the ReGLU
($\sigma \to \mathrm{ReLU}$), the \glsterm{geglu}{GEGLU} ($\sigma \to \mathrm{GELU}$),
and the \glsterm{swiglu}{SwiGLU} ($\sigma \to \mathrm{Swish}_\beta$, with
$\mathrm{Swish}_\beta(z) = z\,\sigma(\beta z)$).

Bundling the gating with the down-projection yields the bias-free
3-matrix \glsterm{ffn}{FFN} that every modern frontier \glsterm{llm}{LLM}
ships~\citep[\S 2, Eq.~6]{shazeer2020}\footnote{KB excerpt:
\texttt{kb/excerpts/shazeer2020.md\#sec-2-eq6}.}:
\begin{equation}
  \mathrm{FFN}_{\mathrm{SwiGLU}}(\mathbf{x}, W, V, W_2)
  \;=\; \bigl(\mathrm{Swish}_1(\mathbf{x} W) \otimes (\mathbf{x} V)\bigr)\, W_2,
  \tag{Shazeer--Eq.~6}
  \label{eq:swiglu}
\end{equation}
with $\mathrm{Swish}_1(z) = z\,\sigma(z)$ (also called SiLU). Per token
the tensor flow now uses three projections of width $d_{\text{ff}}$
rather than two, so to compare against
Eq.~\eqref{eq:vaswani-ffn} at matched parameter count Shazeer reduces
the inner width by a factor of $\tfrac{2}{3}$~\citep[\S 2]{shazeer2020}:
\begin{quote}
``All of these layers have three weight matrices, as opposed to two
for the original \glsterm{ffn}{FFN}. To keep the number of \glsterm{parameters}{parameters} and the amount
of computation constant, we reduce the number of hidden units
$d_{\text{ff}}$ \dots\ by a factor of $\tfrac{2}{3}$ when comparing
these layers to the original two-matrix
version.''~\citep[\S 2]{shazeer2020}
\end{quote}
This is the origin of the modern $d_{\text{ff}} \approx
\tfrac{8}{3}\mathsf{D}$ rule. For LLaMA-1 7B, $\mathsf{D} = 4096$ gives
target $d_{\text{ff}} = 10923$, rounded to 11008 for hardware
alignment~\citep{touvron2023}.

The empirical motivation is~\citet[\S 3.2,
Tab.~1]{shazeer2020}\footnote{KB excerpt:
\texttt{kb/excerpts/shazeer2020.md\#sec-3-2}.}: at 524k \glsterm{t5}{T5}-base steps
the heldout-set log-perplexities are 1.677 (\glsterm{relu}{ReLU}), 1.679 (\glsterm{gelu}{GELU}), 1.683
(Swish), 1.633 (\glsterm{geglu}{GEGLU}), and 1.636 (\glsterm{swiglu}{SwiGLU}). \glsterm{geglu}{GEGLU} and \glsterm{swiglu}{SwiGLU} edge each
other within $0.003$; the field selected \glsterm{swiglu}{SwiGLU} on the inertial
strength of LLaMA-1's choice.

\begin{intuition}
Vaswani's \glsterm{ffn}{FFN} (Eq.~\eqref{eq:vaswani-ffn}) computes a sum of features
$\mathrm{ReLU}(\mathbf{x} W_1) W_2 = \sum_k \mathrm{ReLU}(\mathbf{x}\!\cdot\!W_1[:,k])\,W_2[k,:]$
--- a sparse weighted sum of ``\glsterm{value}{value} rows'' selected by which
``\glsterm{key}{key} rows'' the input activates. \glsterm{swiglu}{SwiGLU}
(Eq.~\eqref{eq:swiglu}) computes a \emph{product} of features:
$\mathrm{Swish}_1(\mathbf{x} W) \otimes (\mathbf{x} V)$. The gating
factor $\mathrm{Swish}_1(\mathbf{x} W)$ multiplies the \glsterm{value}{value}
$\mathbf{x} V$ component-wise, which expresses ``this \glsterm{feature}{feature}, only
when that \glsterm{feature}{feature} is also active'' patterns the additive form cannot.
\citet[\S 4]{shazeer2020} closes with the well-known disclaimer ---
``We offer no explanation as to why these architectures seem to
work; we attribute their success, as all else, to divine benevolence''
--- so the canonical form Eq.~\eqref{eq:swiglu} is the only honest
endpoint of the intuition.
\end{intuition}

\subsection{Mixture of Experts: Switch and Mixtral}
\label{sec:ffn-moe-coarse}

The dense \glsterm{ffn}{FFN}'s parameter cost grows with $\mathsf{D}^2$ at fixed
depth. Mixture of Experts (MoE) breaks the coupling: replace the
single \glsterm{ffn}{FFN} at depth $\ell$ with $E$ parallel \glsterm{ffn}{FFN}
\emph{experts}, learn a router that picks $k \ll E$ experts per
token, and pay compute for $k$ \glsterm{ffn}{FFN} evaluations while holding $E$
times the \glsterm{parameters}{parameters}. Per-token compute scales with $k$;
parameter capacity scales with $E$.

Let $\mathbf{x}_t \in \R^{\mathsf{D}}$ be the token's residual-stream
slice and $\{\mathrm{FFN}_e\}_{e=1}^{E}$ a family of \glsterm{swiglu}{SwiGLU} blocks
(Eq.~\eqref{eq:swiglu}). The router is a learned linear projection
$W_g \in \R^{\mathsf{D} \times E}$ producing per-expert \glsterm{logits}{logits}, with
top-$k$ selection and a \glsterm{softmax}{softmax}-renormalized gating
weight~\citep[\S 2.1]{fedus2021-switch}\footnote{KB note:
\texttt{kb/notes/architecture/moe.md\#sec-1}; the equation form is
verified against \citet{fedus2021-switch} for top-1 routing and
\citet{mixtral2024} for top-2 routing. \deepencite{fedus2021-switch
\S 2.1 (router-and-gate equations) and mixtral2024 \S 2.2 (top-2
form) --- KB excerpt files not yet ground-truthed against the PDFs.}}:
\begin{align}
  \mathbf{g}_t       &= W_g\, \mathbf{x}_t \quad\in\R^{E}, \label{eq:moe-router} \\
  \mathcal{T}_t      &= \mathrm{TopK}_k(\mathbf{g}_t) \subset \{1,\dots,E\}, \label{eq:moe-topk} \\
  p_{t,e}            &= \frac{\exp(g_{t,e})}{\sum_{e' \in \mathcal{T}_t} \exp(g_{t,e'})}, \quad e \in \mathcal{T}_t, \label{eq:moe-gate} \\
  \mathrm{MoE}(\mathbf{x}_t) &= \sum_{e \in \mathcal{T}_t} p_{t,e}\,\mathrm{FFN}_e(\mathbf{x}_t). \label{eq:moe-output}
\end{align}
Switch \glsterm{transformer}{Transformer}~\citep{fedus2021-switch} fixes $k = 1$: the router
sends each token to exactly one expert, $|\mathcal{T}_t| = 1$, and
$p_{t,e} = 1$. The choice maximally simplifies routing arithmetic
and minimizes communication --- there is exactly one all-to-all
dispatch per layer. Switch-XXL has $E \approx 64$ experts and
$\sim\!395\mathrm{B}$ total \glsterm{parameters}{parameters} at $\sim\!12\mathrm{B}$ active.

Mixtral~\citep{mixtral2024} fixes $k = 2$, $E = 8$: each token's
output is the gated combination of \emph{two} expert FFNs.
Mixtral-8$\times$7B reports $46.7\mathrm{B}$ total \glsterm{parameters}{parameters},
$12.9\mathrm{B}$ active. Both designs leave the dense
\glsterm{swiglu}{SwiGLU} \glsterm{ffn}{FFN} unchanged \emph{within} each expert; the router and the
combination Eq.~\eqref{eq:moe-output} are the additions.

\begin{analogy}
The MoE layer is a soft committee of specialists. The router inspects
each token and dispatches to the few experts most relevant; the chosen
experts produce candidate updates $\mathrm{FFN}_e(\mathbf{x}_t)$, and
the gate weights $p_{t,e}$ blend them. The committee is ``soft''
because the weights in Eq.~\eqref{eq:moe-gate} are continuous; the
dispatch is ``hard'' because the top-$k$ selection
Eq.~\eqref{eq:moe-topk} is discrete. Returning to math: the layer's
output is exactly the gated weighted sum
$\sum_{e \in \mathcal{T}_t} p_{t,e}\,\mathrm{FFN}_e(\mathbf{x}_t)$
of Eq.~\eqref{eq:moe-output}; the analogy survives because each
expert is a full \glsterm{swiglu}{SwiGLU} \glsterm{ffn}{FFN} of the form Eq.~\eqref{eq:swiglu}, not a
parameter-shared head.
\end{analogy}

\subsection{Routing collapse and the load-balancing loss}
\label{sec:ffn-load-balance}

The router in Eq.~\eqref{eq:moe-router}--\eqref{eq:moe-gate} is
differentiable through $p_{t,e}$ on selected experts but
\emph{discrete} through Eq.~\eqref{eq:moe-topk}. Two failure modes
follow.

\paragraph{Routing collapse.}
Without intervention, the router learns to send most tokens to a small
set of ``favored'' experts; under-used experts receive vanishing
gradient and never improve, while over-used experts get richer
updates. The dynamic is self-reinforcing.

\paragraph{Load imbalance.}
Even without total collapse, GPUs hosting cold experts sit idle while
GPUs hosting hot experts oversubscribe. Under expert parallelism
(one expert per device), throughput is bounded by the slowest device.

The standard mitigation is an auxiliary load-balancing loss
$\mathcal{L}_{\text{aux}}$, added to the language-modeling loss with
coefficient $\alpha$. For a batch of $T$ tokens, define the empirical
fraction $f_e$ of tokens routed to expert $e$ and the mean router
probability $P_e$ for expert
$e$~\citep[\S 2.2]{fedus2021-switch}\footnote{KB note:
\texttt{kb/notes/architecture/moe.md\#sec-2-1}. \deepencite{fedus2021-switch
\S 2.2 --- the auxiliary-loss form below is verified against
secondary sources; the original PDF section number should be
re-checked before final.}}:
\begin{align}
  f_e &= \frac{1}{T} \sum_{t=1}^T \mathbb{1}\!\left[e \in \mathcal{T}_t\right], \label{eq:moe-fe} \\
  P_e &= \frac{1}{T} \sum_{t=1}^T \frac{\exp(g_{t,e})}{\sum_{e'=1}^E \exp(g_{t,e'})}, \label{eq:moe-pe} \\
  \mathcal{L}_{\text{aux}} &= \alpha \cdot E \sum_{e=1}^{E} f_e\, P_e. \label{eq:moe-aux}
\end{align}
The product $f_e P_e$ is differentiable through $P_e$ even though
$f_e$ depends on the discrete selection
Eq.~\eqref{eq:moe-topk}, and is minimized at uniform load
$f_e = P_e = 1/E$. Mixtral and Switch both use the form
Eq.~\eqref{eq:moe-aux}; production stacks pair it with a
\emph{capacity factor} $C \in [1, 2]$ that caps each expert at
$\lceil C\,T k / E \rceil$ tokens per batch and drops (residual-only
passes) tokens routed past the
cap~\citep[\S 2.3]{fedus2021-switch}.

\citet[\S 2]{deepseek-v3} report that high-$\alpha$ auxiliary losses
hurt LM quality at scale, and propose an
\emph{auxiliary-loss-free} alternative: maintain a per-expert
\emph{bias} $b_e$ that is added to the router \glsterm{logits}{logits} at selection
time but \emph{not} to the gate weight, and update the bias online
based on observed
load~\citep{deepseek-v3}\footnote{KB note:
\texttt{kb/notes/architecture/moe.md\#sec-2-3}. \deepencite{deepseek-v3
\S 2 (auxiliary-loss-free balancing scheme) --- claim ground-truthed
against the V3 tech report's training-side excerpt at
\texttt{kb/excerpts/deepseek-v3-training.md\#abstract}, but the
exact equation form requires the architecture-side excerpt that is
not yet built.}}:
\begin{equation}
  \mathbf{g}_t' \;=\; W_g\,\mathbf{x}_t + \mathbf{b},
  \qquad
  b_e \;\leftarrow\; b_e + \gamma\,(\bar{f} - f_e),
  \quad \bar{f} = 1/E,
  \label{eq:moe-aux-free}
\end{equation}
with $\gamma$ a small step size. Overloaded experts get $b_e$
decreased so future tokens go elsewhere; underloaded experts get
$b_e$ increased. Since $\mathbf{b}$ shifts the \emph{selection}
\glsterm{logits}{logits} but not the renormalized gate $p_{t,e}$ in
Eq.~\eqref{eq:moe-gate}, the contribution magnitudes stay
clean. \glsterm{precision-pinning}{DeepSeek-V3} reports this strategy recovers the LM quality
that high-$\alpha$ auxiliary loss would
sacrifice~\citep[abstract]{deepseek-v3}.

\subsection{DeepSeekMoE: fine-grained experts and a shared expert}
\label{sec:ffn-fine-grained}

\citet{deepseekmoe2024} push two changes against the
Switch/Mixtral design point: \emph{fine-grained experts} (many narrow
experts in place of few wide ones) and a \emph{shared expert} that
processes every
token~\citep{deepseekmoe2024}\footnote{KB note:
\texttt{kb/notes/architecture/moe.md\#sec-3-1,\#sec-3-2}. \deepencite{deepseekmoe2024
\S 3 (fine-grained segmentation: split each expert into $m$
narrower experts of width $d_{\text{ff}}/m$ and select $mk$ rather
than $k$) --- excerpt file not yet built; arithmetic of
combinatorial set count $\binom{E}{k}$ verified against the
DeepSeek-V2/V3 reported $E$, $k$.}}.

\paragraph{Fine-grained segmentation.}
A coarse Mixtral expert at $d_{\text{ff}} = \tfrac{8}{3}\mathsf{D}$ is
replaced by $m$ narrower experts of width $d_{\text{ff}}/m$; the
top-$k$ selection becomes top-$mk$. Active per-token compute is held
fixed (the $m$ scaling cancels), but the number of distinct activated
combinations $\binom{E}{mk}$ grows combinatorially. Concretely:
Mixtral has $\binom{8}{2} = 28$ active sets; DeepSeek-V2 with
$E = 160$, $k = 6$ has $\binom{160}{6} \approx 2 \times 10^{10}$;
\glsterm{precision-pinning}{DeepSeek-V3} with $E = 256$, $k = 8$ has
$\binom{256}{8} \approx 4 \times 10^{14}$. The argument is that
combinatorial expressiveness of the active set --- which functions
the layer can dynamically assemble from its experts --- scales with
this count, so finer experts at fixed active compute give more
\emph{flexible specialization} per token.

\paragraph{Shared expert as residual.}
Layered alongside the $E$ routed experts is a small set $E_s$ of
\emph{shared} experts that process \emph{every} token unconditionally.
With $E_s = 1$ (the \glsterm{precision-pinning}{DeepSeek-V3} choice) the layer becomes
\begin{equation}
  \mathrm{MoE}^{\text{DS}}(\mathbf{x}_t)
  \;=\;
  \mathrm{FFN}_{\text{shared}}(\mathbf{x}_t)
  \;+\; \sum_{e \in \mathcal{T}_t} p_{t,e}\,\mathrm{FFN}_e(\mathbf{x}_t),
  \label{eq:deepseek-moe}
\end{equation}
with $|\mathcal{T}_t| = k$ routed experts selected by
Eq.~\eqref{eq:moe-router}--\eqref{eq:moe-gate} and the shared expert
$\mathrm{FFN}_{\text{shared}}$ contributing unconditionally. The
shared term is structurally a \emph{residual within the \glsterm{ffn}{FFN}
sublayer}: it provides the always-on common-knowledge component that
the routed experts no longer have to redundantly learn. The motivation
is that without a shared expert, each routed expert is forced to
re-learn token-level features that every token needs (basic syntax,
general \glsterm{vocabulary}{vocabulary}), wasting capacity that could be spent on
specialization. Eq.~\eqref{eq:deepseek-moe} factorizes that into
``shared $+$ specialized.''

\paragraph{Frontier-2026 instances.}
\glsterm{precision-pinning}{DeepSeek-V3} ships $E = 256$ routed, $E_s = 1$ shared, $k = 8$ active,
totaling $9$ active experts per token from $257$ total per
layer~\citep[abstract,
\S 2]{deepseek-v3}\footnote{KB excerpt:
\texttt{kb/excerpts/\glsterm{precision-pinning}{deepseek-v3}-training.md\#abstract}.}. With $671\mathrm{B}$
total \glsterm{parameters}{parameters} and $37\mathrm{B}$ activated per token, the
parameter-to-active ratio is $\approx 18\!:\!1$ ---
the largest such ratio shipped at frontier as of
2026-Q2. \citet{qwen3} ship $E = 128$, $E_s = 0$, $k = 8$ in
Qwen3-235B with $22\mathrm{B}$ active, deliberately dropping shared
experts and replacing them with a global-batch load-balancing
loss~\citep{qwen3}.

\subsection{Frontier-2026 choice landscape}
\label{sec:ffn-frontier}

The dense-vs-MoE split in 2026 is no longer a question of
\emph{whether} to do MoE at frontier scale --- the answer is yes ---
but of \emph{how}. Three concrete design points coexist
in publicly-disclosed frontier
models~\citep{meta-llama3,meta-llama4,deepseek-v3,qwen3,mixtral2024}:
\begin{itemize}
\item \textbf{Dense at moderate scale.} LLaMA-3 8B and 70B remain
    fully dense \glsterm{swiglu}{SwiGLU} FFNs at every layer~\citep{meta-llama3}; the
    open-weights ecosystem retains a single dense codepath through
    the small/medium tier.
\item \textbf{Coarse MoE.} Mixtral 8$\times$7B and 8$\times$22B keep
    $E = 8$, $k = 2$, no shared expert, classical
    auxiliary-loss balancing~\citep{mixtral2024}; every layer is MoE.
\item \textbf{Fine-grained MoE with shared expert.} \glsterm{precision-pinning}{DeepSeek-V3} with
    $E = 256$, $E_s = 1$, $k = 8$, auxiliary-loss-free balancing,
    every \glsterm{ffn}{FFN} sublayer except the first is
    MoE~\citep{deepseek-v3}; the V2/V3 lineage is the most
    aggressive parameter-to-active ratio shipped.
\item \textbf{Fine-grained MoE without shared expert.} Qwen3 with
    $E = 128$, $E_s = 0$, $k = 8$, global-batch load
    balancing~\citep{qwen3}.
\item \textbf{LLaMA-4 hybrid.} Scout and Maverick alternate dense and
    MoE \glsterm{ffn}{FFN} sublayers, with $E = 16$ (Scout) or $E = 128$ (Maverick),
    $E_s = 1$, $k = 1$ ~\citep{meta-llama4}.
\end{itemize}
The two structural primitives common to every MoE in this list ---
\glsterm{swiglu}{SwiGLU} experts and a learned \glsterm{softmax}{softmax} router with top-$k$ ---
are the same primitives Switch fixed in 2021. What evolved is the
\emph{granularity} ($E$ from 8 to 256), the \emph{shared-expert}
position, and the \emph{load-balancing} discipline (aux-loss $\to$
aux-loss-free $\to$ global-batch).

\begin{contradiction}
The empirical claim that fine-grained MoE matches dense
quality at \emph{matched active-parameter count} is the most
load-bearing and least-replicated assertion in this section. \citet{deepseek-v3}
report \glsterm{precision-pinning}{DeepSeek-V3} ($37\mathrm{B}$ active) tracking or exceeding
LLaMA-3 405B (dense) on benchmark
suites~\citep[abstract]{deepseek-v3}; \citet{deepseekmoe2024}
make a similar matched-active-FLOPs claim at smaller scale.
Independent third-party reproductions at frontier scale are sparse as
of 2026-Q2: most public head-to-head numbers come from the
DeepSeek/Qwen labs themselves or from secondary aggregators. \citet{qwen3}
explicitly disagree with \citet{deepseek-v3} on whether shared
experts are net beneficial, dropping $E_s$ to zero and reporting
comparable quality with global-batch balancing. Treat the
``MoE matches dense at matched active params'' claim as
\emph{strongly supported by primary sources, weakly cross-validated
by independent reproductions} until 2027-class non-DeepSeek/Qwen
frontier MoE training runs publish their gap numbers.
\deepencite{matched-active MoE-vs-dense quality at frontier scale ---
no published independent third-party head-to-head as of 2026-Q2.}
\end{contradiction}

\subsection{Forward link}
\label{sec:ffn-forward}

Two structural decisions of this section --- \glsterm{swiglu}{SwiGLU} as the per-expert
\glsterm{ffn}{FFN} and a \glsterm{softmax}{softmax}-routed top-$k$ wrapper around it --- are now
independent of the surrounding block. The next axis
(\cref{sec:normalization-axis}) is what surrounds the \glsterm{ffn}{FFN} in the
\glsterm{residual-stream}{residual stream}: which normalization is applied, and whether it sits
before the sublayer (\glsterm{pre-ln}{Pre-LN}), after it (\glsterm{post-ln}{Post-LN}), or wraps it
(Peri-LN). The frontier-2026 stack composes a routed \glsterm{swiglu}{SwiGLU} MoE
of \cref{eq:deepseek-moe} with the \glsterm{rmsnorm}{RMSNorm} + \glsterm{pre-ln}{Pre-LN} +
$1/\sqrt{N}$-init bundle of \cref{sec:normalization-axis}; the \glsterm{ffn}{FFN}
choices made here forward-determine which normalization placements
remain stable, because the \glsterm{residual-stream}{residual stream}'s per-block variance budget
is shared between attention and \glsterm{ffn}{FFN}
sublayers~\citep{xiong2020}.
