% Paper 1, §9 — State-space models and hybrids: the alternative path
% Thesis: S4/S5 → Mamba (selective SSM) → Mamba-2 (state-space-duality
% with attention). Hybrids: Jamba, Samba, Zamba2, Hymba. Pure-SSM has
% not displaced attention at frontier; hybrids are gaining.

\section{State-space models and hybrids: the alternative path}
\label{sec:state-space-and-hybrids}
\label{sec:state-space}

The architectural axes of \S\ref{sec:positional-encoding}--\S\ref{sec:normalization-axis}
all live inside the attention-and-FFN-with-residuals scaffold inherited
from~\citet{vaswani2017}. A different lineage drops that scaffold for
the residual stream while keeping it for the per-token mixing function:
the \emph{state-space model} (SSM) replaces the attention block's
$\mathrm{softmax}(QK^\top)V$ with a \emph{recurrent state}
$\boldsymbol{h}_t \in \mathbb{R}^d$ that summarizes the prefix in fixed
size. The thesis of this section is that the SSM lineage --
S4/S5 $\to$ Mamba $\to$ Mamba-2 -- has produced (i) an architecture with
$O(N d^2)$ training cost and $O(d^2)$ per-token decode, (ii) a theorem
(\citealp{mamba2}'s State Space Duality) that recasts a class of SSMs
as structurally restricted attention, and (iii) a family of
\emph{hybrid} models that interleave SSM and attention blocks. As of
mid-2026, no pure-SSM model has been deployed at frontier scale; the
deployed long-context architectures are either attention-based or
SSM-attention hybrids, and the gap between hybrid demos and frontier
attention models has not been closed by third-party reproduction.

\subsection{The state-space recurrence form}
\label{sec:ssm-recurrence}

A linear continuous-time SSM of dimension $d$ on scalar input
$u(t) \in \mathbb{R}$ is the classical control-theory system
\begin{equation}
  \frac{d \boldsymbol{h}(t)}{d t} \;=\; A\,\boldsymbol{h}(t) + B\,u(t),
  \qquad
  y(t) \;=\; C\,\boldsymbol{h}(t) + D\,u(t),
  \label{eq:ssm-continuous}
\end{equation}
with $A \in \mathbb{R}^{d \times d}$, $B \in \mathbb{R}^{d \times 1}$,
$C \in \mathbb{R}^{1 \times d}$, $D \in
\mathbb{R}$.\deepencite{gu2023-mamba \S 2 (KB excerpt missing; verify
notation against PDF Eq.~(1a-1b))} Discretization with step $\Delta$ via
zero-order hold yields the discrete recurrence used at training and
inference time:
\begin{equation}
  \boldsymbol{h}_t \;=\; \bar{A}\,\boldsymbol{h}_{t-1} + \bar{B}\,u_t,
  \qquad
  y_t \;=\; C\,\boldsymbol{h}_t,
  \tag{Mamba--Eq.~2a}
  \label{eq:ssm-discrete}
\end{equation}
with $\bar{A} = \exp(\Delta A)$ and $\bar{B} =
(\Delta A)^{-1}(\exp(\Delta A) - I)\,(\Delta B)$.\deepencite{gu2023-mamba
\S 2.1 (KB excerpt missing; verify discretization formula)} Two
properties of \eqref{eq:ssm-discrete} make it attractive as a sequence
mixer. First, applied to a length-$N$ sequence the recurrence costs
$O(N d^2)$ flops -- \emph{linear} in $N$ where attention is quadratic.
Second, for fixed $\bar{A}, \bar{B}, C$ the output equals a 1-D
convolution of $u$ with a kernel
$\bar{K} = (C\bar{B},\, C\bar{A}\bar{B},\, C\bar{A}^2\bar{B}, \ldots)$,
so $y_{1:N}$ can be computed in $O(N \log N)$ via FFT during training
while still streamable as an RNN at decode time -- the
\emph{convolution-recurrence dual} that \citet{gu2023-mamba} inherit
from S4.

The S4 line~\citep[no KB entry; \deepencite{gu2022-s4 (Gu, Goel, R\'e
2022): HiPPO-structured $A$ matrix, Eq.~(2)--(3) of S4 paper}]{}
chooses $A$ to be a HiPPO operator that ``optimally compresses'' input
history under an orthogonal-polynomial basis; S5~\deepencite{smith2023-s5
(no KB entry): diagonal-plus-low-rank $A$ replacing S4's HiPPO-NPLR
parameterization, Eq.~(4)} simplifies the parameterization to a diagonal
$A$ with multi-input-multi-output recurrence. Both share
\eqref{eq:ssm-discrete}; both rely on the convolution-recurrence dual to
get parallel training. In modern decoder-only notation, the SSM mixer
takes a tensor of shape $\BSH$ and applies \eqref{eq:ssm-discrete}
independently along each of the $\mathsf{D}$ feature channels with a
shared (or per-channel) parameterization.

\subsection{Selective SSMs: Mamba}
\label{sec:ssm-mamba}

\citet{gu2023-mamba}'s ``Mamba'' breaks the convolution-recurrence dual
deliberately. The SSM \eqref{eq:ssm-discrete} is
\emph{linear-time-invariant}: $\bar{A}, \bar{B}, C$ do not depend on
$u_t$, so the same kernel $\bar{K}$ applies to every position. This
means the SSM cannot \emph{select} which inputs to incorporate vs.\
ignore -- a problem on associative-recall, copying, and induction-head
tasks where the model must conditionally route information based on
content. Mamba's fix: make $\bar{A}, \bar{B}, C$ \emph{input-dependent}
via per-token linear projections $s_B(\cdot), s_C(\cdot),
s_\Delta(\cdot)$ of the current input,
\begin{equation}
  \bar{B}_t \;=\; s_B(u_t),\quad
  C_t \;=\; s_C(u_t),\quad
  \Delta_t \;=\; \tau_\Delta(s_\Delta(u_t)),
  \label{eq:mamba-selection}
\end{equation}
with $A$ kept as a learned diagonal matrix and $\Delta_t$ passed through
a softplus $\tau_\Delta$.\deepencite{gu2023-mamba \S 3.1 (KB excerpt
missing; verify Eq.~(4) and the $s_B, s_C, s_\Delta$ parameterization)}
Substituting into \eqref{eq:ssm-discrete} gives
\begin{equation}
  \boldsymbol{h}_t \;=\; \bar{A}_t\,\boldsymbol{h}_{t-1} + \bar{B}_t\,u_t,
  \qquad
  y_t \;=\; C_t\,\boldsymbol{h}_t,
  \tag{Mamba--Eq.~2b}
  \label{eq:mamba-selective}
\end{equation}
where $\bar{A}_t, \bar{B}_t$ now vary with position. The
\emph{selection mechanism} is $\Delta_t$: large $\Delta_t$ means the
recurrence ``forgets'' more of $\boldsymbol{h}_{t-1}$ and absorbs more
of $u_t$, while small $\Delta_t$ preserves state. Per-token control of
$\Delta_t$ lets Mamba implement input-dependent gating analogous to the
update gate of an LSTM, but with the recurrence's matrix form intact.

The cost is that \eqref{eq:mamba-selective} is no longer a convolution
(the kernel changes per position), so the FFT-based parallel training
recipe of S4/S5 fails. Mamba recovers training-time GPU parallelism via
a hardware-aware \emph{parallel scan}~\citep[\S
3.3]{gu2023-mamba}.\deepencite{gu2023-mamba \S 3.3 (KB excerpt missing;
verify the selective-scan algorithm and the SRAM-tiled implementation
described as ``hardware-aware'')} The scan uses an associative
combination of the per-position transition tuples
$(\bar{A}_t, \bar{B}_t u_t)$ to compute all
$\boldsymbol{h}_{1:N}$ in $O(\log N)$ depth on GPU, and avoids
materializing the full state by keeping the scan in SRAM. The Mamba
block wraps the selective SSM in a gated-MLP envelope (a SiLU-gated
linear unit, with a residual addition and an RMSNorm at the
input)~\citep[\S 3.4]{gu2023-mamba} so the unit is drop-in-replaceable
for an attention~+~FFN sublayer pair.

\begin{intuition}
The state $\boldsymbol{h}_t \in \mathbb{R}^d$ is a fixed-size
\emph{compression buffer} for the prefix $u_{1:t}$. Each new $u_t$
overwrites part of the buffer; $\Delta_t$ controls how much. Attention,
by contrast, keeps every prior $u_{1:t}$ in addressable form via the
KV cache and pays $O(N)$ memory for it. The SSM's buffer is $O(d^2)$
regardless of $N$ -- cheap, but lossy: any token whose information has
been overwritten is unrecoverable. Returning to math: per-token decode
cost is $O(d^2)$ for the SSM (one application of \eqref{eq:ssm-discrete})
versus $O(N d_h)$ for attention (one query against the cache). The
crossover where SSM is cheaper is at $N \sim d^2 / d_h$, which for
$d \approx 16$ (Mamba's state width) and $d_h = 128$ is on the order of
a few tokens -- the SSM is essentially always cheaper at decode.
\end{intuition}

\subsection{State Space Duality: Mamba-2}
\label{sec:ssm-mamba2}

\citet{mamba2}'s contribution is the \emph{State Space Duality} (SSD):
a theorem-and-framework showing that a class of selective SSMs and a
restricted form of attention compute the same function. Concretely, the
authors show that the input-output map of a Mamba-2 block can be
written
\begin{equation}
  \boldsymbol{y} \;=\; M\,\boldsymbol{u},
  \qquad
  M \in \mathbb{R}^{N \times N}\ \text{lower-triangular,
  $1$-semi-separable of rank}\le d,
  \label{eq:ssd}
\end{equation}
where $\boldsymbol{u} = u_{1:N}$ is the stacked input and $M$ is a
lower-triangular \emph{semi-separable} matrix with rank at most the SSM
state width $d$.\deepencite{mamba2 \S 3 (KB excerpt missing; verify the
SSD theorem statement, the semi-separable rank bound, and the duality
between the recurrent form \eqref{eq:mamba-selective} and the matrix
form \eqref{eq:ssd})} The semi-separable structure both (i) recovers
$O(N d^2)$ flops via a structured matmul that reuses
FlashAttention-style block tiling, and (ii) frames Mamba-2 as a vanilla
attention with a structurally restricted attention matrix. The reported
2--8$\times$ wall-clock speedup over Mamba-1 comes from this
matmul-friendly form.\deepencite{mamba2 \S 6 (KB excerpt missing;
benchmarks vs Mamba-1 and FlashAttention-2)}

\begin{analogy}
Mamba-2 is ``attention with a bandwidth budget.'' Vanilla causal
attention computes a dense lower-triangular $N \times N$
matrix $A_{ij} = \mathrm{softmax}(QK^\top)_{ij}$ with no rank
constraint; FlashAttention-class kernels just compute it efficiently.
Mamba-2 constrains the analogous matrix $M$ in \eqref{eq:ssd} to be
rank-$\le d$ semi-separable -- the same recurrent state's information
bottleneck rewritten in attention notation. The continuum is: vanilla
attention (unconstrained) $\to$ sliding-window attention (banded) $\to$
Mamba-2 (rank-restricted). Returning to math: the cost of attention is
$O(N^2 d_h)$ flops to fill the unconstrained $N \times N$ matrix; the
cost of Mamba-2 is $O(N d \cdot d)$ flops to apply the rank-$d$
factorization \eqref{eq:ssd}. The bandwidth budget $d$ is exactly the
ratio of the two.
\end{analogy}

The SSD theorem matters less for raw speed (FlashAttention-3 closes
much of that gap on H100/H200; cf.\
\S\ref{sec:hardware-implementation-axis}) than for taxonomy: it tells
us that the practical distinction between attention and SSM is not
``recurrent vs.\ parallel'' but rather \emph{which structural restriction
on the attention matrix the architecture imposes}. This is the
theoretical lens under which hybrids stop being weird and start looking
like a deliberate mix of low- and high-rank attention layers.

\subsection{Hybrid architectures}
\label{sec:ssm-hybrids}

The hybrid lineage drops the all-or-nothing question -- ``SSM or
attention?'' -- and treats the decoder stack as a place to interleave
both. The layout question becomes: at what ratio, in what pattern, with
what bridge between the two block types? Four representative designs
illustrate the choices:

\paragraph{Jamba (AI21, 2024).}
\citet{lieber2024-jamba} interleave \emph{1 attention layer per 7 Mamba
layers} on a 12B active / 52B total MoE backbone, with full causal
attention on the attention layers (using GQA;
\S\ref{sec:attention-mechanism-axis}) and Mamba-1 selective SSMs on the
SSM layers.\deepencite{lieber2024-jamba \S 3 (KB excerpt missing; verify
the 1:7 ratio, the MoE configuration, and the placement of attention
layers within the stack)} The 1:7 ratio is enough attention to cover
recall-style needs (associative-recall, induction, copying) while
keeping the cheaper SSM substrate dominant; AI21 report comparable
quality to similar-active-parameter dense Transformers at lower KV-cache
cost.

\paragraph{Samba (Microsoft, 2024).}
A finer interleaving: SSM and sliding-window attention alternate at a
near-1:1 ratio, with the attention restricted to a local window rather
than full causal. The design assumes the SSM substrate handles
long-range mixing while the local-window attention handles
short-range exact-recall.\deepencite{ren2024-samba (no KB entry; verify
the alternation pattern, the window size, and the small-LM scale at
which Samba is reported)}

\paragraph{Zamba2 (Zyphra, 2024).}
A different bridge: a Mamba-2 backbone with two \emph{shared}
attention blocks applied periodically across the stack. The shared-block
design lets a single set of attention weights act at multiple depths,
amortizing the attention parameter cost; the authors report a
$\sim 6\times$ KV-cache reduction relative to Zamba1.\deepencite{zamba2-2024
(no KB entry; verify the shared-attention-block mechanism, the
backbone-layer count, and the $6\times$ KV-cache claim)}

\paragraph{Hymba (NVIDIA, 2024).}
A \emph{parallel} rather than serial hybrid: each block has SSM heads
and attention heads operating on the same input in parallel, with the
outputs concatenated before the FFN. The design avoids the serial
attention-bottleneck on long sequences (the attention layer in a Jamba
stack must wait for all 7 prior SSM outputs at decode) at the cost of
running both mixers per block.\deepencite{hymba-2024 (no KB entry;
verify the parallel-head architecture, the head-count split between
SSM and attention, and the small-LM benchmark scale)}

The interleaving ratio is the key knob. Jamba's 1:7 says attention is
\emph{rare-but-essential}; Samba's $\sim$1:1 says local attention is
worth pairing with every SSM layer; Zamba2's shared-attention says one
set of attention weights amortizes across depth; Hymba's parallel design
says compute both mixers always and let the concat-and-project step
choose. None of these are head-to-head compared at matched
parameters/compute in published work, and all four are at sub-frontier
scale (12B active is the largest deployed example,
\citet{lieber2024-jamba}).

\begin{speculation}
\textbf{Why the 1:7 hybrid ratio is roughly right.} An attention layer
once every $\sim 8$ tokens of stack-depth gives the model enough
exact-recall capacity to handle the rare positions where the SSM
substrate cannot, while keeping the cheaper SSM dominant for the
routine residual-stream work. The optimal ratio likely depends on the
task distribution: code probably benefits from more attention,
narrative summarization benefits from more SSM. The published lineage
has not run that ratio sweep at frontier scale, so this remains a
plausible heuristic rather than a derived constant.
\end{speculation}

\subsection{Why pure SSM has not won at frontier}
\label{sec:ssm-frontier}

Three obstacles keep pure SSM models from displacing attention at the
frontier as of mid-2026, all traceable to the fixed-state information
bottleneck of \eqref{eq:ssm-discrete}.

\paragraph{In-context learning and exact recall.}
A pure SSM compresses the entire prefix into $\boldsymbol{h}_t \in
\mathbb{R}^d$ ($d \approx 16$ for Mamba; $d \approx 128$ for some
follow-ups). Tasks that require \emph{exact retrieval} of a specific
prior token -- copying, hashing, multi-hop induction,
needle-in-haystack -- expose this bottleneck: by the time the relevant
token's information matters, intervening updates may have overwritten
it.\deepencite{mamba2 \S 7, jamba2024 ablations (KB excerpts missing;
both papers' ablations show pure-Mamba below Transformer baselines on
associative-recall and needle-in-haystack at matched compute)} This is
the empirical motivation for hybrids: an attention layer is the
straightforward fix to the recall bottleneck, and one or two attention
layers per stack already close most of the gap.

\paragraph{Retrieval-style task brittleness.}
Beyond pure recall, tasks that depend on attending to specific tokens
based on \emph{content} (retrieval-augmented QA, code-completion that
references a specific function name, multi-document reasoning) degrade
faster on pure SSM than on attention as context grows. The forum-signal
consensus is that this is the real ceiling, but it has not been
quantitatively reported at frontier
scale.\deepencite{tier-B benchmarks: HuggingFace open-LLM leaderboard
2025 entries for Falcon-Mamba 7B vs Llama-class 7B at matched compute,
no primary paper}

\paragraph{Training-data efficiency.}
A subtler claim made informally in the Mamba-2 follow-up
literature~\deepencite{mamba2 \S 7 (KB excerpt missing; verify whether
the paper makes a quantitative training-token-efficiency claim or only
an inference-FLOPs claim)}: pure SSMs at matched parameters require
\emph{more} pretraining tokens to reach the same loss as Transformers,
because the recall-limited mixer makes some training signal harder to
absorb. If true, this changes the cost calculus: the SSM's training
flop savings are partly offset by needing more tokens.

\begin{contradiction}
\label{con:ssm-frontier-quality}
Quality of SSM-attention hybrids at frontier scale (the 100B-active
range, where DeepSeek-V3, Llama-4, and the Claude/GPT-class systems
operate) is an open question as of 2026-Q2. Vendor demos
(\citealp{lieber2024-jamba} at 52B total / 12B active; Zamba2 at 7B;
Hymba at 1.5B) are not third-party-reproduced at frontier scale, and no
controlled head-to-head exists between (Jamba 1:7), (Samba $\sim$1:1),
(Zamba2 shared-attention), and (Hymba parallel) at matched
parameters/compute. \citet{mamba2}'s SSD theorem is \emph{evidence} that
a hybrid stack with a small attention budget should be Pareto-better
than pure SSM, but the optimal interleaving pattern, ratio, and
attention-block count remain settled by neither theory nor the
publicly-deployed model zoo.
\end{contradiction}

The frontier-2026 deployed long-context architectures
(DeepSeek-V3~\citep{deepseek-v3}, Llama-4~\citep{meta-llama3},
Qwen3~\citep{qwen3}, Mistral Large~\citep{jiang2023}) are all
attention-based; none ship an SSM block. The forum signal is that
hybrids will appear at scale eventually -- the SSD theorem and the
demonstrated long-context FLOP advantage are real -- but the public
leaderboard as of 2026-05 contains no SSM-hybrid model in the top
quality tier.

\subsection{What this section settles, and what comes next}
\label{sec:ssm-summary}

The state-space line is the most theoretically distinct alternative to
the attention scaffold of \S\ref{sec:positional-encoding}--\S\ref{sec:normalization-axis},
and it has produced a real architectural family with a real cost
advantage on long sequences. Two facts are stable: (i) selective SSMs
\eqref{eq:mamba-selective} are linear-time per-token mixers with
$O(d^2)$ decode cost regardless of $N$; (ii) the SSD theorem
\eqref{eq:ssd} unifies a class of SSMs with rank-restricted attention.
Two facts are open: (a) whether hybrid stacks scale to frontier quality,
and (b) what the optimal interleaving pattern is. The
conclusion is that as of mid-2026 the field has \emph{retained}
attention as the high-rank exact-recall mechanism, and added SSM blocks
where cheap context-mixing is the binding constraint -- but only in
secondary models, not in the deployed frontier.

The next section (\S\ref{sec:multi-token-output}) returns to the
attention scaffold and asks a different question: at the
\emph{output} side of the decoder, what changes when the model predicts
multiple tokens at once? Multi-Token Prediction
(\citealp{gloeckle2024-mtp}; productionized in
\citealp{deepseek-v3}) is an architectural choice on the output head,
orthogonal to the SSM-vs-attention question of this section, and to the
KV-sharing axis of \S\ref{sec:attention-mechanism-axis}.
