% Paper 1, §13 — Frontier-2026 snapshot: what choices the leading models share
% Pass-C synthesis section. Cross-tabulates the per-axis lineages of §3-§12 across
% the publicly-known 2026 frontier model lineup. Output is a 2-page snapshot table
% (the "most-screenshotted figure" of the paper) plus per-cluster discussion.

\section{Frontier-2026 snapshot: what choices the leading models share}
\label{sec:frontier-snapshot}

The previous ten sections walked the architectural axes one at a
time. This section composes them. We tabulate the per-axis choices
made by the 2026 frontier model lineup --- DeepSeek V3 and R1, Llama
4, \glsterm{mistral}{Mistral} Large 2, Gemma 3, Qwen 2.5 / Qwen 3, plus the partially
disclosed \glsterm{claude}{Claude} 3.7-class, GPT-class, and Gemini 2.5 systems --- on
the same eight axes used throughout this paper, and call out where
the published architectures agree and where they diverge.
\textbf{The thesis of this section: the convergence is real, narrow,
and concentrated in five of the eight axes; the remaining three
(KV-sharing strategy, dense-vs-MoE on the \glsterm{ffn}{FFN}, and the multimodal
integration pattern) are the live splits, and only one of them
(KV-sharing) cleanly partitions the disclosed lineup.}

The table that follows replaces the model-zoo prose of vendor blog
posts with a uniform ten-column row per model. Where a lab has
published a primary source we cite it; where a detail is vendor-blog
or release-card disclosure only we mark
\deepencite{citation-pending}; where the architecture is undisclosed
we record ``not disclosed'' rather than guess from inference-stack
reverse engineering, which is unreliable across point releases.

\subsection{The snapshot table}
\label{sec:snapshot-table}

\begin{table}[!h]
\centering
\caption{Frontier-2026 architectural choices, per axis. Columns: KV-sharing
(MHA / GQA-$g$ / MLA), attention implementation (FA2 / FA3 / NSA),
FFN family (dense SwiGLU / MoE / fine-grained MoE / shared-expert MoE),
normalization (RMSNorm + Pre/Peri-LN), positional encoding (RoPE plus
extension), tokenizer, multi-token output (none / MTP), multimodal
integration. ``ND'' = not publicly disclosed at architectural detail.
Sources are cited inline in
\S\ref{sec:snapshot-disclosed}--\S\ref{sec:snapshot-undisclosed}.}
\label{tab:frontier-snapshot}
\renewcommand{\arraystretch}{1.25}
\footnotesize
\begin{tabularx}{\textwidth}{@{}lXXXXXXXX@{}}
\toprule
\textbf{Model} & \textbf{KV} & \textbf{Attn impl} & \textbf{\glsterm{ffn}{FFN}} &
\textbf{Norm} & \textbf{Pos} & \textbf{Tok} & \textbf{Multi-tok} &
\textbf{MM} \\
\midrule
DeepSeek V3 & \glsterm{multi-head-latent-attention}{MLA}, $d_c{=}512$ & FA + custom \glsterm{multi-head-latent-attention}{MLA} kernel & MoE
$E{=}256$, $k{=}8$, $E_s{=}1$ & \glsterm{rmsnorm}{RMSNorm} + \glsterm{pre-ln}{Pre-LN} & decoupled-\glsterm{rope}{RoPE} &
\glsterm{bpe}{BPE} 129K & \glsterm{multi-token-prediction}{MTP}, 1 head & text-only \\
DeepSeek R1 & \glsterm{multi-head-latent-attention}{MLA} (V3 base) & FA + custom \glsterm{multi-head-latent-attention}{MLA} kernel & MoE (V3 base) &
\glsterm{rmsnorm}{RMSNorm} + \glsterm{pre-ln}{Pre-LN} & decoupled-\glsterm{rope}{RoPE} & \glsterm{bpe}{BPE} 129K & \glsterm{multi-token-prediction}{MTP}, 1 head &
text-only \\
Llama 4 (Scout / Maverick) & \glsterm{gqa}{GQA}-8 & FA3 & dense$\!+\!$MoE
alternating; $E{\in}\{16,128\}$, $E_s{=}1$, $k{=}1$ & \glsterm{rmsnorm}{RMSNorm} + \glsterm{pre-ln}{Pre-LN} &
\glsterm{irope}{iRoPE} (3 \glsterm{rope}{RoPE} : 1 \glsterm{nope}{NoPE}) + temperature & \glsterm{sentencepiece}{SentencePiece} \glsterm{bpe}{BPE} & none
disclosed & native multimodal \\
\glsterm{mistral}{Mistral} Large 2 & \glsterm{gqa}{GQA}-8 & FA2 / FA3 & dense \glsterm{swiglu}{SwiGLU} & \glsterm{rmsnorm}{RMSNorm} +
\glsterm{pre-ln}{Pre-LN} & \glsterm{rope}{RoPE} + position-IDs ext.\ & tekken (\glsterm{bpe}{BPE}) & none disclosed &
text-only (Pixtral variant: native MM) \\
Gemma 3 27B & \glsterm{gqa}{GQA} + 5 local : 1 global (window 1024) & FA2 & dense \glsterm{swiglu}{SwiGLU} &
\glsterm{rmsnorm}{RMSNorm} + Peri-LN + QK-Norm & \glsterm{rope}{RoPE} + interleaved windows & \glsterm{sentencepiece}{SentencePiece}
\glsterm{bpe}{BPE} & none disclosed & native MM (4B$+$ Vision) \\
Qwen 2.5 / Qwen 2.5-VL & \glsterm{gqa}{GQA} & FA2 / FA3 & dense \glsterm{swiglu}{SwiGLU} &
\glsterm{rmsnorm}{RMSNorm} + \glsterm{pre-ln}{Pre-LN} & \glsterm{rope}{RoPE} + \glsterm{yarn}{YaRN} & \glsterm{bpe}{BPE} 152K & none disclosed &
projector + \glsterm{llm}{LLM} (VL) \\
Qwen 3 235B-A22B & \glsterm{gqa}{GQA} & FA3 & MoE $E{=}128$, $k{=}8$, $E_s{=}0$,
global-batch balancing & \glsterm{rmsnorm}{RMSNorm} + \glsterm{pre-ln}{Pre-LN} & \glsterm{rope}{RoPE} & \glsterm{bpe}{BPE} 152K &
none disclosed & text-only (separate MM line) \\
\midrule
\glsterm{claude}{Claude} 3.7-class & ND & ND & ND & ND & ND & proprietary \glsterm{bpe}{BPE} & ND &
native MM \\
GPT-class (o-/$\!4o$/5) & ND & ND & ND & ND & ND & tiktoken-family &
ND & native MM (4o, 5) \\
Gemini 2.5 & ND & ND & ND & ND & ND & \glsterm{sentencepiece}{SentencePiece} & ND &
native MM \\
\bottomrule
\end{tabularx}
\end{table}

\Cref{tab:frontier-snapshot} is the headline figure of this paper.
\S\ref{sec:snapshot-disclosed}--\S\ref{sec:snapshot-undisclosed} walk
the disclosed and undisclosed clusters; then
\S\ref{sec:snapshot-agreement}--\S\ref{sec:snapshot-divergence} read
the table for points of agreement and points of divergence.

\subsection{Disclosed cluster: the open-weights frontier}
\label{sec:snapshot-disclosed}

The seven rows in the upper block of \cref{tab:frontier-snapshot} are
either fully open-weights with an architectural tech report, or
disclose architectural detail in a public release card. We walk them
briefly to anchor each cell.

\paragraph{DeepSeek V3 and R1.}
DeepSeek V3 is a 671B-total / 37B-active MoE with
$E = 256$, $E_s = 1$, $k = 8$, auxiliary-loss-free
balancing~\citep[\S 2.2]{deepseek-v3}.\footnote{KB excerpt:
\texttt{kb/excerpts/\glsterm{precision-pinning}{deepseek-v3}.md\#sec-2}.} The attention sublayer
uses \glsterm{multi-head-latent-attention}{MLA} with joint-compression rank $d_c = 512$ and the
decoupled-\glsterm{rope}{RoPE} construction of \citet[\S 2.1.2 Eq.~9]{deepseek-v2}.
\glsterm{multi-token-prediction}{Multi-Token Prediction} with one auxiliary head is part of the
pretraining objective~\citep[\S 2.3.1]{deepseek-v3}.
\glsterm{rmsnorm}{RMSNorm} in \glsterm{pre-ln}{Pre-LN} placement is inherited from the LLaMA
template~\citep{touvron2023}. DeepSeek-R1 is a continued-RL fine-tune
of V3-base with no architectural changes other than a training-stage
thinking-mode toggle~\citep[\S 2]{deepseek-r1}; its row repeats V3's
choices.

\paragraph{Llama 4 (Scout, Maverick).}
LLaMA 4 ships two MoE configurations: Scout with $E = 16$ and
Maverick with $E = 128$, both with $E_s = 1$, $k = 1$, and the \glsterm{ffn}{FFN}
sublayer alternated between dense \glsterm{swiglu}{SwiGLU} and MoE per
layer~\citep[\S 2.1]{meta-llama4}.\deepencite{meta-llama4 §2.1 ---
Llama 4 release card / Meta release blog, citation-pending: the
\texttt{meta-llama4} cite-key currently maps only the long-context /
iRoPE detail; the MoE and native-multimodal architectural detail
needs consolidation into a separate \texttt{meta-llama4-release}
entry.} Position encoding is \glsterm{irope}{iRoPE} (3:1
\glsterm{rope}{RoPE}-to-\glsterm{nope}{NoPE}) with temperature scaling on the \glsterm{nope}{NoPE}
layers~\citep[\S 2.1, \S 2.4]{meta-llama4}, trained natively at
$256\text{K}$. KV-sharing is \glsterm{gqa}{GQA}-8, inherited from
LLaMA-3~\citep[\S 3.1]{meta-llama3}. Multimodality is native: image
and text are jointly pretrained from scratch~\citep[\S 2]{meta-llama4}.

\paragraph{Mistral Large 2.}
\glsterm{mistral}{Mistral} Large 2 is dense \glsterm{swiglu}{SwiGLU} throughout, with \glsterm{gqa}{GQA}-8 attention and
\glsterm{rmsnorm}{RMSNorm} + \glsterm{pre-ln}{Pre-LN} inherited from the \glsterm{mistral}{Mistral} 7B
template~\citep[\S 2]{jiang2023}.\deepencite{citation-pending: Mistral
Large 2 release notes / Hugging Face model card, July 2024;
\texttt{mistral-large-2} is not yet a references.bib entry.}
Position encoding is \glsterm{rope}{RoPE}; multi-token output is not disclosed beyond
single-head autoregression. The Pixtral variant adds a projector +
vision encoder (\cref{sec:multimodal}); the text-only \glsterm{mistral}{Mistral} Large
2 is not multimodal.\footnote{\textbf{Superseded 2025-12-02.}
\glsterm{mistral}{Mistral} Large 3 \citep{mistral-large3-2025} crosses the
\glsterm{mistral}{Mistral} flagship line over to MoE: 675B-total / 41B-active sparse
MoE under Apache 2.0, 256K context, native multimodal. The
\texttt{\glsterm{mistral}{Mistral} Large 2} row is retained here because its detail is
the publicly-documented dense reference point Paper 1 \S\ref{sec:ffn}
uses as a counter-example; on the dense-vs-MoE roster of
\S\ref{sec:snapshot-divergence}, the \glsterm{mistral}{Mistral} row should now be read
as MoE going forward, and the dense-vs-MoE split tilts further
toward MoE than the row alone shows.}

\paragraph{Gemma 3.}
Gemma 3 27B uses \glsterm{gqa}{GQA} with a $5{:}1$ local-to-global interleave (sliding
window $1024$ tokens) of Gemma 2's pattern~\citep[\S 2.1]{gemma3} --- five
sliding-window local-attention layers per one full global-attention layer. Normalization
is the Peri-LN placement of \citet{peri-ln2025} combined with QK-Norm
on the attention path~\citep[\S 2.1]{gemma3,olmo2}. Position encoding
is \glsterm{rope}{RoPE}; the \glsterm{ffn}{FFN} is dense \glsterm{swiglu}{SwiGLU}; multimodality is native via a
SigLIP-class vision
encoder.\deepencite{gemma3 §3: multimodal pipeline detail;
verify SigLIP variant against the Gemma 3 release card.}

\paragraph{Qwen 2.5 and Qwen 3.}
Qwen 2.5 is dense \glsterm{swiglu}{SwiGLU} with \glsterm{gqa}{GQA}, \glsterm{rmsnorm}{RMSNorm} + \glsterm{pre-ln}{Pre-LN}, \glsterm{rope}{RoPE} + \glsterm{yarn}{YaRN}, and
a $152$K-vocab \glsterm{bpe}{BPE} tokenizer.\deepencite{citation-pending: Qwen 2.5
release card; \texttt{qwen3} maps the Qwen 3 paper, not Qwen 2.5
architectural detail.} Qwen 3 (235B-A22B) replaces the dense \glsterm{ffn}{FFN} with
a fine-grained MoE at $E = 128$, $E_s = 0$, $k = 8$, with
\glsterm{global-batch-load-balancing}{global-batch load balancing} in place of auxiliary-loss
balancing~\citep[\S 2.1]{qwen3}. Crucially Qwen 3 keeps \glsterm{gqa}{GQA} rather
than \glsterm{multi-head-latent-attention}{MLA}, despite training in the same year as DeepSeek V3 --- the
lab-level KV-sharing split flagged in
\cref{sec:attention-kv-sharing}. The multimodal line (Qwen 2.5-VL)
uses a projector, not native multimodal.

\subsection{Undisclosed cluster: Claude, GPT, Gemini}
\label{sec:snapshot-undisclosed}

The lower block of \cref{tab:frontier-snapshot} is the closed-source
frontier. None of Anthropic (\glsterm{claude}{Claude} 3.5 / 3.7 / 4-class), OpenAI
(GPT-4o, o1 / o3, GPT-5), or Google DeepMind (Gemini 2.5) has
published per-axis architectural detail at the level of the open
labs. The model cards report capabilities, training data scale,
context length, and multimodal coverage; they do not report
KV-sharing strategy, attention-implementation kernel, \glsterm{ffn}{FFN} family, or
normalization placement.\deepencite{Anthropic Claude 3.7 system card
/ OpenAI GPT-4o and GPT-5 system cards / Google Gemini 2.5
technical report --- citation-pending: vendor system cards are
discovery-only sources (Tier B/C in
\texttt{theory/sources/README.md}) and cannot solely back hard
architectural claims; consolidate into separate cite-keys once
formal venues catch up.} What is publicly known is restricted to:

\begin{itemize}
\item \emph{\glsterm{tokenization}{Tokenization} family.} OpenAI's tiktoken family is
publicly documented; Anthropic's \glsterm{bpe}{BPE}-class tokenizer is documented
in the API reference but not in a primary technical report; Gemini
uses a \glsterm{sentencepiece}{SentencePiece}-class tokenizer disclosed in the Gemma 3 paper
\citep[\S 2.3]{gemma3} as a sibling to Gemini's own.
\item \emph{Multimodal pattern.} GPT-4o, \glsterm{claude}{Claude} 3.5 Sonnet onward,
and Gemini 2.5 are native-multimodal (image + audio + text in a
single pretraining); this is disclosed at capability level in each
vendor's system card and is the third integration pattern of
\cref{sec:multimodal}.
\item \emph{Reasoning-mode toggle.} OpenAI o1 / o3 and (since 3.7)
\glsterm{claude}{Claude} expose a thinking-mode generation phase, mirroring the
DeepSeek-R1 / Qwen 3 toggle~\citep[\S 2]{deepseek-r1, qwen3}.
Whether this requires architectural support beyond training-stage
changes is contested
(\cref{sec:state-space-and-hybrids,sec:multi-token-output}) and
not resolved by the closed-vendor disclosures.
\end{itemize}

\begin{forumsignal}
Inference-stack reverse-engineering of closed-vendor models (latency
patterns; per-token cost differentials; rough head-count estimation
from KV-cache utilization scaling) circulates in tier-C community
sources but is unreliable across point releases and unverifiable
without vendor cooperation. We treat all three undisclosed cells as
``ND'' rather than risk laundering forum speculation as architectural
fact. The convergence claim of this section is therefore restricted
to the disclosed seven rows; whether the closed labs sit inside or
outside the disclosed convergence cluster is a falsifiable prediction
of \cref{sec:contradictions}, not a claim of this section.
\end{forumsignal}

\paragraph{Caveat: closed-vendor row labels age faster than the
architectural picture.} The closed-row labels in
\cref{tab:frontier-snapshot} (``\glsterm{claude}{Claude} 3.7-class,'' ``GPT-class
(o-/$\!4o$/5),'' ``Gemini 2.5'') are accurate as a 2025-Q1 vintage of
the closed lineup. By 2026-Q2 each lab has shipped multiple point
releases beyond those labels --- Anthropic's \glsterm{claude}{Claude} 4-family
(May 2025) plus subsequent Opus 4.x and Sonnet 4.x updates; OpenAI's
GPT-5 (August 2025) folding the o-series into a unified
fast-vs-thinking tier, with subsequent GPT-5.x point releases; and
Gemini 2.5 Deep Think (August 2025) with its parallel-thinking RL
disclosure. The architectural picture this section's thesis depends on
\emph{does not move} across those point releases at any disclosed
level: the closed rows remain ``ND'' on every architectural axis the
table tracks, and the productionised exposure of inference-time
compute (extended-thinking budgets, reasoning-effort levels,
parallel-thinking modes) converges on the same API surface as the
disclosed labs rather than diverging from it.\deepencite{closed-vendor system
cards (Anthropic Claude 4-family, OpenAI GPT-5.x, Google Gemini 2.5
Deep Think) are tier-B per \texttt{theory/sources/README.md} and back
the version-lineage / API-surface claims, not the architectural cells;
those cells remain ``ND.''} We retain the 2025-Q1 row labels with
this date stamp rather than extend them per release: the column-by-
column architectural cells would still be ``ND'' for every newer
label, and the row's load-bearing function in the table is the
disclosure-ceiling, not the version stamp. The convergence claim of
this section is therefore stable across the closed-row lineage shift
between 2025-Q1 and 2026-Q2; the lineage shift itself is documented
in \cref{sec:contradictions} as a separate question about whether the
disclosure ceiling is moving.

\subsection{Where the disclosed frontier agrees}
\label{sec:snapshot-agreement}

Five of the eight axes show near-universal agreement across the
seven disclosed rows of \cref{tab:frontier-snapshot}.

\begin{enumerate}
\item \textbf{Normalization: \glsterm{rmsnorm}{RMSNorm} + \glsterm{pre-ln}{Pre-LN}-class placement.} All
seven disclosed models use \glsterm{rmsnorm}{RMSNorm} with the norm placed before each
sublayer rather than after the residual add. Gemma 3 and OLMo 2
extend this to Peri-LN with QK-Norm~\citep{gemma3,olmo2}; the base
choice (\glsterm{rmsnorm}{RMSNorm} + \glsterm{pre-ln}{Pre-LN} + $1/\sqrt{N}$ residual init) is universal
(\cref{sec:normalization-axis}).

\item \textbf{Positional encoding: \glsterm{rope}{RoPE} plus a long-context
extension.} All seven disclosed rows apply \glsterm{rope}{RoPE} to $Q, K$ at every
attention layer (\cref{sec:positional-encoding}). The extension varies ---
vanilla, \glsterm{yarn}{YaRN}, \glsterm{irope}{iRoPE}, decoupled-\glsterm{rope}{RoPE} for \glsterm{multi-head-latent-attention}{MLA}, interleaved
local / global --- but the base mechanism (multiplicative rotation
of $Q, K$ parameterised by $\theta_i = \theta^{-2(i-1)/d_h}$) is
shared.

\item \textbf{\glsterm{ffn}{FFN} gating: \glsterm{swiglu}{SwiGLU}.} Every disclosed model uses
\glsterm{swiglu}{SwiGLU} as the per-position \glsterm{ffn}{FFN} nonlinearity, whether the sublayer
is dense or MoE. The 3-matrix
$\mathrm{FFN}(x) = (W_3 (\mathrm{Swish}(W_1 x) \odot W_2 x))$
form of \citet[Eq.~6]{shazeer2020} is universal; the dense-vs-MoE
split is on the multiplicity around it, not on the activation.

\item \textbf{KV-sharing: head-count reduction.} No disclosed
2026-frontier model uses full \glsterm{mha}{MHA}. Every row sits at \glsterm{gqa}{GQA}-8,
\glsterm{gqa}{GQA}-with-windowing, or \glsterm{multi-head-latent-attention}{MLA} --- all reducing $H_{\mathrm{kv}}$ below
$H_q$. The lab-aligned split between \glsterm{gqa}{GQA} and \glsterm{multi-head-latent-attention}{MLA} persists, but
\emph{some} form of head-count reduction is universal, and every
major inference framework (\glsterm{flashattention}{FlashAttention}, vLLM, \glsterm{compressed-fsm-2}{SGLang}) now ships
first-class \glsterm{gqa}{GQA} and increasingly first-class
\glsterm{multi-head-latent-attention}{MLA}~\citep{kwon2023,dao2023}.

\item \textbf{Attention implementation: \glsterm{flashattention}{FlashAttention}-family exact
attention.} Every disclosed model uses a \glsterm{flashattention}{FlashAttention}-class
exact-attention kernel: FA2 on Ampere, FA3 on Hopper, with
\glsterm{multi-head-latent-attention}{MLA}-aware variants for the DeepSeek
stack~\citep{dao2023,shah2024,deepseek-v3}. \glsterm{nsa}{NSA}'s three-branch
sparse attention~\citep{yuan2025} has not yet shipped at production
scale at any disclosed lab; the formula
$\softmax(QK^\top/\sqrt{d_h})V$ is unchanged across implementations.
\end{enumerate}

\begin{intuition}
Five-of-eight convergence is striking but consistent with the rest
of the paper: the axes that have converged are the ones where the
\emph{primary citation set} of \cref{sec:positional-encoding} through
\cref{sec:normalization-axis} contains a single dominant lineage
(\glsterm{rope}{RoPE}, \glsterm{swiglu}{SwiGLU}, \glsterm{rmsnorm}{RMSNorm} + \glsterm{pre-ln}{Pre-LN}, head-count-reduced KV-sharing,
\glsterm{flashattention}{FlashAttention}-class implementation). The axes that remain split
are the ones where the primary set still contains parallel
lineages (MoE-vs-dense; \glsterm{cross-attention}{cross-attention}-vs-projector-vs-native
multimodal; \glsterm{gqa}{GQA}-vs-\glsterm{multi-head-latent-attention}{MLA} at the lab level). Returning to canonical
math: the residual-stream forward pass on a 2026 frontier model
collapses to the same five-line equation (\glsterm{rmsnorm}{RMSNorm} $\to$ rotated-$Q,K$
attention with reduced $H_{\mathrm{kv}}$ $\to$ residual-add $\to$
\glsterm{rmsnorm}{RMSNorm} $\to$ \glsterm{swiglu}{SwiGLU} \glsterm{ffn}{FFN} or routed \glsterm{swiglu}{SwiGLU} MoE $\to$ residual-add)
across every disclosed row. The convergence is not a stylistic
preference; it is a fixed point of the per-axis arguments in
\cref{sec:tokenization-embedding,sec:positional-encoding,sec:attention-kv-sharing,sec:attention-hardware-axis,sec:ffn,sec:normalization-axis}.
\end{intuition}

\subsection{Where the disclosed frontier diverges}
\label{sec:snapshot-divergence}

Three axes carry live splits in \cref{tab:frontier-snapshot}.

\paragraph{KV-sharing strategy: GQA versus MLA.}
The split is lab-aligned, not capability-aligned. DeepSeek V2 / V3 /
R1 use \glsterm{multi-head-latent-attention}{MLA} throughout~\citep{deepseek-v2,deepseek-v3,deepseek-r1};
every other disclosed lab (Meta, \glsterm{mistral}{Mistral}, Google, Alibaba) uses \glsterm{gqa}{GQA}.
The contradiction in \cref{sec:attention-kv-sharing} is that \glsterm{multi-head-latent-attention}{MLA}'s
quality story --- matched \glsterm{mha}{MHA} quality at $4\times$ smaller KV-cache
--- is a single-source primary claim
(\citealp{deepseek-v2}) and has not been independently reproduced
at frontier scale by a non-DeepSeek lab. The serving-stack maturity
gap (\glsterm{flashattention}{FlashAttention}'s \glsterm{gqa}{GQA} path being older than its \glsterm{multi-head-latent-attention}{MLA} path) and
the recipe risk of \glsterm{multi-head-latent-attention}{MLA}'s decoupled-\glsterm{rope}{RoPE} construction
(\cref{sec:attention-kv-sharing}) are plausible explanations for the
non-crossover. \cref{sec:contradictions} returns to this.

\paragraph{Dense versus MoE on the FFN.}
The dense-versus-MoE split is genuine but tilting further toward MoE
through 2025--26. LLaMA-3 8B and 70B are dense~\citep{meta-llama3};
Qwen 2.5 base is dense; \glsterm{mistral}{Mistral} Large 2 was dense, but the lab's
flagship line crossed to MoE with \glsterm{mistral}{Mistral} Large 3 (Dec 2025;
\citealp{mistral-large3-2025}: 675B-total / 41B-active sparse MoE).
DeepSeek V3 is MoE with $E = 256$, $E_s = 1$, $k = 8$; Qwen 3 is MoE
with $E = 128$, $E_s = 0$, $k = 8$; Llama 4 is hybrid (alternating
dense and MoE layers); Mixtral 8$\times$7B remains the canonical
8-expert classical MoE point. As discussed in
\cref{sec:ffn-frontier}, this is no longer a question of whether
to do MoE at scale --- the answer is yes for the largest models in
this list --- but of which MoE design point: $E$, $E_s$, $k$, and
load-balancing discipline. The contradiction in
\cref{sec:ffn} is the matched-active-parameter quality
claim~\citep{deepseek-v3,deepseekmoe2024}: that fine-grained MoE
matches dense quality at matched active \glsterm{parameters}{parameters}. Cross-lab
reproduction at frontier scale is sparse, and Qwen 3's
shared-expert-free choice~\citep{qwen3} explicitly disagrees with
DeepSeek V3 on whether $E_s > 0$ helps.

\paragraph{Multimodal integration pattern.}
Three patterns coexist in \cref{tab:frontier-snapshot}: native
multimodal pretraining (LLaMA 4, Gemma 3 multimodal variant,
\glsterm{claude}{Claude} 3.5+, GPT-4o / 5, Gemini 2.5), projector + \glsterm{llm}{LLM}
(Qwen 2.5-VL, Pixtral / \glsterm{mistral}{Mistral} Pixtral), and the
\glsterm{cross-attention}{cross-attention}-bridge variant (no current frontier model; Flamingo
descendants are off-frontier). The \glsterm{cross-attention}{cross-attention} bridge of
\citet{alayrac2022-flamingo} has effectively been replaced by the
two simpler patterns; the live split is between projector and
native multimodal. The vendor claim that native multimodal produces
better quality at smaller active parameter count
(\cref{sec:multimodal}) is a forum signal pending controlled
cross-lab reproduction, and the closed-vendor cluster's coverage
here is uniform but undisclosed at architectural detail.

\subsection{Closing: which axes have converged, which remain mixed}
\label{sec:snapshot-closing}

The empirical state of the disclosed 2026 frontier is summarized in
two sentences. \textbf{The convergence is a five-axis fixed point
across all disclosed labs}: \glsterm{rope}{RoPE}-with-extension, \glsterm{rmsnorm}{RMSNorm} in a
\glsterm{pre-ln}{Pre-LN}-class placement, \glsterm{swiglu}{SwiGLU} per-position \glsterm{ffn}{FFN}, head-count-reduced
KV-sharing (some form of \glsterm{gqa}{GQA} or \glsterm{multi-head-latent-attention}{MLA}), \glsterm{flashattention}{FlashAttention}-class exact
attention; \textbf{the residual splits are concentrated in three
axes}: KV-sharing strategy (\glsterm{gqa}{GQA} vs \glsterm{multi-head-latent-attention}{MLA}, lab-aligned, single-source
quality claim), dense-vs-MoE on the \glsterm{ffn}{FFN} multiplicity (genuine
quality / serving-cost trade with cross-lab disagreement on
$E_s > 0$), and multimodal integration pattern (projector vs native,
no controlled cross-lab study). The undisclosed cluster (\glsterm{claude}{Claude},
GPT, Gemini) is silent on all eight axes at architectural detail
and provides no signal on whether it sits inside or outside the
disclosed convergence neighbourhood.

The remaining contradictions surfaced through
\cref{sec:positional-encoding,sec:attention-kv-sharing,sec:attention-hardware-axis,sec:ffn,sec:state-space-and-hybrids,sec:multi-token-output,sec:multimodal,sec:inference-adjuncts}
--- \glsterm{multi-head-latent-attention}{MLA}-vs-\glsterm{mha}{MHA} at frontier scale, \glsterm{nsa}{NSA}-vs-FA3 head-to-head,
\glsterm{state-space-model}{SSM}-hybrid quality at frontier, the matched-active MoE quality gap,
long-context extrapolation method comparison, \glsterm{multi-token-prediction}{MTP} at non-DeepSeek
scale, native-multimodal quality lift --- are now visible as a
single concentrated set. \Cref{sec:contradictions} treats them as
the open frontier of the architecture story, and asks for each what
evidence would close the gap. The five-axis convergence of this
section is the substrate against which those open questions are
posed; without naming the convergence, the contradictions cannot be
localized; without localizing the contradictions, the field's
remaining architectural questions cannot be made falsifiable.
