% Paper 1, §14 — Contradictions live and open frontiers
% Pass-C synthesis section. Concentrates the \begin{contradiction} blocks
% from §3-§13 into a numbered list, gives each its strongest claims on each
% side, and identifies the experiment that would close it. ~4 pages.

\section{Contradictions live and open frontiers}
\label{sec:contradictions}

The previous twelve sections traced the convergence: along five of the
eight axes the disclosed 2026 frontier models agree
(\S\ref{sec:frontier-snapshot}), and the remaining three are split in a
small number of camps. This section consolidates the
\texttt{[CONTRADICTION]} markers raised across
\S\ref{sec:tokenization-embedding-unembedding}--\S\ref{sec:inference-adjuncts}
into an enumerated catalogue. For each open question we state the
strongest form of each side's claim, list the evidence currently on the
table, and name the experiment that would settle it. \textbf{The thesis
of this section: the live architectural disputes are localized,
falsifiable, and almost all bottlenecked by the same scarce resource ---
controlled head-to-head training runs at the $\geq 70\mathrm{B}$-active
scale on independent training stacks. We predict that 2027-class
disclosures will close \S\ref{sec:c-mla}, \S\ref{sec:c-nsa}, and
\S\ref{sec:c-mtp} along axes
\S\ref{sec:attention-mechanism-axis}--\S\ref{sec:state-space-and-hybrids}, and that
\S\ref{sec:c-moe} and \S\ref{sec:c-longctx} will remain mixed for at
least one further model generation.}

We use a uniform sub-template per contradiction: \emph{strongest claim
on each side}, \emph{evidence on the table}, \emph{experiment that
closes it}. Where the supporting source for a side is single-laboratory
or pre-print only, we mark it \deepencite{} so the next deepening pass
can trace the lineage.

\subsection{C1 --- MLA vs.\ MHA at frontier scale}
\label{sec:c-mla}

\textbf{Pro-MLA (DeepSeek):} Multi-head Latent Attention with the
joint-low-rank decomposition $W^{DKV}, W^{UK}, W^{UV}$ matches MHA
quality at $\sim 4\times$ smaller KV cache; the cache absorption
identity (\S\ref{sec:attention-mechanism-axis}, eq.\ MLA-absorb) means
no quality penalty at decode \citep[\S 2.1.4]{deepseek-v2}.
\textbf{Pro-GQA (default):} GQA with $G \in \{4, 8\}$ groups recovers
MHA quality at matched cache budget on every published independent
ablation \citep[\S 3.1]{ainslie2023}, and Western frontier labs that do
disclose attention details (\citealp{meta-llama3, meta-llama4,
gemma3, qwen3}) all stay on GQA.

\textbf{Evidence on the table.} The MLA-MHA matched-cache comparison is
single-source \citep[\S 2.1.4]{deepseek-v2}: a 16B-active prototype on
DeepSeek's training corpus. \citet{deepseek-v3} extends MLA to 671B
total / 37B active with strong downstream benchmarks but does not
publish a head-to-head MLA-vs-MHA at matched compute at that scale.
GQA's matched-cache neutrality result \citep{ainslie2023} has been
replicated by multiple labs.

\textbf{Experiment that closes it.} A $\geq 70\mathrm{B}$-active
controlled three-way ablation --- MHA / GQA-8 / MLA at matched compute,
matched data, matched optimizer --- run by an independent laboratory on
non-DeepSeek pre-training data. As of 2026-Q2 no such run is published.
\deepencite{independent MLA-vs-GQA-vs-MHA at $\geq 70$B-active scale}.

\subsection{C2 --- NSA vs.\ FA3 head-to-head}
\label{sec:c-nsa}

\textbf{Pro-NSA:} Native Sparse Attention with the three-branch
(compressed / fine / sliding) selection scheme delivers $11.6\times$
decode at 64K and improves long-context reasoning quality over the
dense baseline, because pre-training with sparse attention teaches the
model to use it \citep{yuan2025}. \textbf{Pro-FA3:} FlashAttention-3
with H100 asynchronous warp-specialization reaches $\sim 75\%$ of
H100's theoretical peak FP16 throughput on dense attention
\citep{shah2024}\deepencite{shah2024 §3 (excerpt not yet in KB; first-use anchor in \S\ref{sec:hardware-attention})}; community indications are that NSA's $11.6\times$
margin shrinks substantially when the dense baseline is FA3 rather than
FA2 and the context length is closer to 32K than 64K.

\textbf{Evidence on the table.} \citet{yuan2025} compare against
FA2-on-GQA, not FA3, leaving the FA3 baseline unexamined. NSA's
training-time recipe is also tied to the DeepSeek pre-training stack;
no third-party laboratory has published a Hopper reproduction at
matched precision and matched task suite as of 2026-05.

\textbf{Experiment that closes it.} A controlled NSA-vs-FA3 benchmark
suite on H100 / H200 with matched precision (BF16 or FP8), matched
sequence lengths $\{8K, 16K, 32K, 64K\}$, and matched downstream
evaluation (RULER, HELMET, not just NIAH). The decode-throughput claim
is the easy half; the harder claim is whether NSA's quality lift
survives when the baseline is FA3 rather than FA2 and when the
pre-training data is not DeepSeek's.
\deepencite{NSA-vs-FA3 third-party head-to-head on Hopper}.

\subsection{C3 --- SSM-attention hybrid quality at frontier}
\label{sec:c-ssm}

\textbf{Pro-hybrid:} A small attention budget interleaved with
selective SSM blocks (\citealp{lieber2024-jamba} 1:7,
Samba\,$\sim$1:1) gives Pareto-better long-context FLOP cost than
attention-only, because Mamba-2's state-space-duality theorem
\citep[\S 3]{mamba2} bounds the loss at small $r$. \textbf{Pro-attention-only:}
The publicly deployed top-quality 2026 lineup (DeepSeek-V3, Llama-4,
Qwen3, Mistral Large) is uniformly attention-based; no SSM-hybrid
appears in the top quality tier \citep{deepseek-v3, meta-llama4, qwen3,
jiang2023}, suggesting that whatever advantage the SSD theorem predicts
is not yet operationalized at the 100B-active scale.

\textbf{Evidence on the table.} Vendor demos at sub-frontier scale
(Jamba 52B total / 12B active; Zamba2 7B; Hymba 1.5B) show competitive
within-tier quality, but neither the optimal interleaving ratio nor the
attention-block count is settled by theory or by a controlled ablation.
The negative evidence is the absence of frontier deployment, which is
informative but not dispositive --- the SSM line has had less
engineering investment than attention, and that confound is not
controlled.

\textbf{Experiment that closes it.} Three matched-compute training runs
at $\geq 70\mathrm{B}$-active: pure-attention, hybrid 1:7
(Jamba-style), hybrid $\sim$1:1 (Samba-style). Evaluate on RULER /
HELMET / 1M-token reasoning tasks at deployment-scale serving cost.
\deepencite{ssm-frontier-quality, see \cref{con:ssm-frontier-quality}}.

\subsection{C4 --- MoE quality at matched active parameters}
\label{sec:c-moe}

\textbf{Pro-MoE:} Fine-grained MoE with $E = 256$, $k = 8$, $E_s = 1$
shared expert at 37B-active matches or exceeds dense LLaMA-3 405B on
benchmark suites \citep[abstract]{deepseek-v3}, with the shared-expert
absorbing always-on capability and the fine-grained routing supplying
specialization \citep{deepseekmoe2024}. \textbf{Pro-dense (or
shared-free MoE):} \citet{qwen3} reach comparable quality with $E_s =
0$ and global-batch load balancing, contradicting the shared-expert
claim. The absence of independent third-party reproductions at
frontier scale leaves the matched-active gap empirically open;
secondary aggregations (largo.dev 2026) treat the gap as closed but
the supporting controlled ablations are scarce.

\textbf{Evidence on the table.} Primary-source claims for MoE's
matched-active parity come from DeepSeek and Qwen labs; the labs
disagree among themselves on shared-expert necessity. No published
non-Chinese-lab frontier MoE training run reports its dense-MoE gap.
Anthropic's Claude family architecture is undisclosed and may not be
MoE at all
(see \texttt{kb/index/contradictions.md\#architecture/transformer-overview}).

\textbf{Experiment that closes it.} A non-DeepSeek, non-Qwen lab
publishes a $\geq 100$B-total / $\geq 30$B-active fine-grained MoE
trained with documented $E$, $k$, $E_s$, with a matched-active dense
baseline at the same compute budget on the same pre-training data, and
reports the gap on standard benchmarks. We expect this in 2027 from
either Mistral, Meta (Llama-5 generation), or an Allen-AI / OLMo
follow-up.
\deepencite{matched-active MoE-vs-dense gap, independent-lab
reproduction}.

\begin{contradiction}
The shared-experts sub-question (a sub-contradiction within C4) is
itself unresolved within the disclosed Chinese-lab MoE family.
\citet{deepseek-v3} report on-loss regressions when ablating the shared
expert; \citet{qwen3} drop $E_s$ and report comparable quality. The
reconciling claim --- that shared-experts help when the routing is
top-$k$ with very fine-grained $E$, and not when global-batch balancing
is sufficiently aggressive --- is currently an
\texttt{[INTUITION]}, not a derivation. A controlled ablation that
varies $(E, k, E_s)$ on a matched-compute grid would settle it.
\end{contradiction}

\subsection{C5 --- Long-context extrapolation method}
\label{sec:c-longctx}

\textbf{Pro-RoPE-with-NTK-scaling:} For the $\le 4\times$ extension
regime (e.g.\ LLaMA-3's bump from 8K to 128K), NTK-aware base rescaling
on RoPE \citep{su2021} is sufficient and is what most disclosed
2026 models actually do. \textbf{Pro-YaRN:} For the $\sim 10\times$
regime, per-frequency schedule (YaRN) is required to avoid
out-of-distribution rotational frequencies; the original
\citet{su2021} embedding under naive scaling collapses on
needle-in-haystack at 32K-trained-256K-evaluated.
\deepencite{peng2023-yarn primary-source numbers}. \textbf{Pro-iRoPE
/ LongRoPE2:} For the $\ge 40\times$ regime (Llama-4's 10M, LongRoPE2's
2M), full retraining with mixed RoPE/NoPE layers and per-position
search beats both NTK-aware and YaRN
\citep{longrope2-2025, meta-llama4}.

\textbf{Evidence on the table.} Each camp's benchmarks are largely
single-source, and the benchmarks themselves disagree: NIAH-style
retrieval saturates at $\sim 100\%$ for all three at $\le 1\mathrm{M}$,
while RULER and HELMET show methodologically inconsistent degradation
patterns past 32K, making ``which extension method is best'' contingent
on which long-context evaluation is treated as ground truth.

\textbf{Experiment that closes it.} A standardized long-context
benchmark (RULER + HELMET + a 1M-token reasoning suite) applied to a
single base model with three deterministically-trained extension
recipes, with full disclosure of the per-frequency schedule. We do not
expect this in 2026; the closing depends on benchmark consensus, which
the field has not yet built.

\begin{intuition}
The three positions are not actually competing for the same regime.
Read horizontally on the extension factor: NTK-scaling owns the
short-extension regime where the unmodified rotational frequencies are
still mostly in-distribution; YaRN owns the medium-extension regime
where per-frequency correction is needed but a single base-rate
adjustment still works; iRoPE / LongRoPE2 own the long-extension regime
where the rotational backbone alone is insufficient and content-derived
position channels (NoPE layers) must absorb the residual.
Returning to the canonical math: the inner-product invariance
$\langle R_m q, R_n k \rangle = \langle q, R_{n-m} k \rangle$ from
\citet[\S 3.4]{su2021} only protects relative position when $|n-m|$
stays in the trained distribution of phases. NTK-scaling, YaRN, and
iRoPE are three points on a hierarchy of how much of the rotational
identity each method preserves.
\end{intuition}

\subsection{C6 --- MTP at non-DeepSeek scale}
\label{sec:c-mtp}

\textbf{Pro-MTP-as-quality:} \citet{deepseek-v3} report a quality lift
from the multi-token prediction auxiliary objective itself, not just
from inference-time draft-token use; the auxiliary gradient pressure
makes hidden states more predictive of the next several tokens jointly
\citep[\S 2.2]{deepseek-v3}. \textbf{Pro-MTP-as-speedup-only:}
\citet{gloeckle2024-mtp} present MTP primarily as a speedup mechanism,
with a smaller per-task quality lift than DeepSeek-V3 reports; the gap
between the two papers is not yet ablated.

\textbf{Evidence on the table.} The two primary sources frame MTP
differently and report quality lifts of different magnitudes on
different benchmark suites. No $\geq 70\mathrm{B}$-active non-DeepSeek
training run has published an MTP ablation at frontier scale; the
question of whether the quality lift survives outside DeepSeek's
training recipe is open.

\textbf{Experiment that closes it.} A non-DeepSeek lab trains two
matched-compute runs at $\geq 70\mathrm{B}$-active --- with-MTP and
without-MTP --- on the same data and reports the quality gap on a
standardized benchmark suite. The cost is small relative to a frontier
training run (MTP adds $\sim$1\%--3\% training compute) and the
information value is high, so we expect this within 12 months.
\deepencite{MTP at non-DeepSeek frontier scale}.

\subsection{Subsidiary open questions raised in passing}
\label{sec:c-subsidiary}

Three further \texttt{[CONTRADICTION]} markers from
\S\ref{sec:tokenization-embedding-unembedding}--\S\ref{sec:inference-adjuncts}
are not load-bearing for Paper 1's convergence thesis but deserve
naming. \textbf{C7 --- Tied-vs-untied embeddings at scale.} The
historical \citet{press2017-tied-embed} result favored tying;
disclosed frontier-2026 models all untie
\citep{meta-llama3, olmo2, deepseek-v3, qwen3}; the reconciling
\texttt{[INTUITION]} (gradient-imbalance dominates only when $|V| d$ is
no longer a large fraction of total parameters) is not a derivation. A
controlled ablation across two or three model sizes spanning the
crossover would settle it. \textbf{C8 --- Medusa verification rule.}
Medusa with rejection sampling is distribution-preserving; Medusa with
typical acceptance is not (\S\ref{sec:multi-token-output},
\cref{eq:spec-reject}). Vendor reports of ``Medusa speedup''
sometimes elide which rule was used; a single deployment audit would
close the reporting gap, though the underlying mathematics is settled.
\textbf{C9 --- NVFP4 production quality.} Vendor materials report
NVFP4 as production-ready with quality matching FP8; third-party
reproduction at frontier scale is sparse. Independent reproductions
through 2026-Q4 should settle whether NVFP4 inference is a tier-A
production claim or a tier-B vendor claim.

\subsection{What the catalogue says about the field}
\label{sec:c-meta}

Six load-bearing contradictions and three subsidiary ones is a small
number for a field this active. Two structural facts emerge.
\textbf{First}, the live disputes are localized to specific axes ---
KV-sharing (\S\ref{sec:c-mla}), hardware implementation
(\S\ref{sec:c-nsa}), the dense--SSM frontier
(\S\ref{sec:c-ssm}, \S\ref{sec:c-moe}), long-context method
(\S\ref{sec:c-longctx}), and multi-token output
(\S\ref{sec:c-mtp}) --- not distributed across the entire architecture.
The normalization, positional-encoding-base, FFN-activation, and
tokenization axes are quietly settled within $\pm 1$ design choice.
\textbf{Second}, the closing experiment for almost every contradiction
is the same: an independent laboratory at $\geq 70\mathrm{B}$-active
runs a controlled head-to-head on a non-DeepSeek pre-training stack.
This is a small number of experiments, not a research program. The
bottleneck is not theory but compute access and a willingness to
publish negative or null results.

We therefore predict that the \emph{shape} of the architectural
convergence visible in \S\ref{sec:frontier-snapshot} will not change in
2027-class disclosures: the same five axes will be settled, three will
remain split, and the splits will narrow. We further predict that
\S\ref{sec:c-mla}, \S\ref{sec:c-nsa}, and \S\ref{sec:c-mtp} are the
contradictions most likely to close --- they are the cheapest to
adjudicate, the most pre-trainable as side-by-side ablations, and
already have at least one well-funded non-DeepSeek lab with the
incentive to publish. \S\ref{sec:c-moe} and \S\ref{sec:c-longctx} will
remain mixed for at least one further generation, the former because
the dense-MoE economics depend on serving regime
(\S\ref{sec:inference-adjuncts}) which differs across labs, the latter
because the long-context evaluation suite has not yet converged.

This closes Paper 1. Papers 2 (training and post-training), 3
(reasoning models), and 4 (interpretability) inherit this catalogue's
discipline: each has its own \texttt{[CONTRADICTION]} cluster, and
their synthesis sections will follow the same enumerate-claim-evidence-
experiment template used here.
