% Paper 4, §13 — Scaling interpretability and frontier-2026 snapshot
% Thesis: each of the five methods has a different scaling envelope. SAEs
% scale on documented L(C), L(N,K) laws; probes are graceful by construction;
% attribution graphs scale to production but expensively and only inside the
% originating lab; manual circuit hunts and full activation patching do not
% yet scale to frontier-2026 architectures (MLA / fine-grained MoE / NSA).
% This section is the compute-and-coverage map of the paper.

\section{Scaling interpretability and frontier-2026 snapshot}
\label{sec:scaling-snapshot}

The five methods of \S\S\ref{sec:logit-lens-family}--\ref{sec:circuits}
were originally exhibited on small open models: GPT-2 small (124M
\glsterm{parameters}{parameters}) for \citet{wang2022-ioi}'s indirect-object-identification
\glsterm{circuit}{circuit}, GPT-2 XL (1.5B) for \citet{meng2022-rome}'s ROME tracing,
Pythia 410M--12B for \citet{belrose2023-tuned-lens}'s tuned-lens
benchmarks. The 2024--2026 question --- and the operational question for
anyone who wants interpretability evidence on a model people actually
deploy --- is which of these methods carry to 2--27B open-weights models
(the Gemma 2 / LLaMA 3 / Qwen 2.5 family) and to production frontier
models (\glsterm{claude}{Claude} 3.5 Haiku, GPT-4 class). The answer is method-dependent,
not uniform: SAEs scale on documented laws, probes scale by construction,
attribution graphs scale to production but only inside Anthropic, and
manual \glsterm{circuit}{circuit} hunts scale not at all. This section is the
compute-and-coverage map; \S\ref{sec:scaling-snapshot-table} consolidates
it into a single snapshot table that becomes the figure-of-record for
the paper.

\subsection{What scales}
\label{sec:scaling-yes}

\paragraph{SAEs scale on documented laws.}
\citet{gao2024-topk-saes} report clean scaling laws for sparse
autoencoders trained with a TopK activation: a compute--loss law
$L(C)$ at fixed sparsity, and a width--sparsity law $L(N, K)$ at fixed
compute, where $N = M$ is the dictionary size and $K$ the per-token
$L_0$ \citep[\S 1, \S 3]{gao2024-topk-saes}\footnote{KB excerpt:
\texttt{kb/excerpts/gao2024-topk-saes.md\#sec-1-headline}.}. The headline
demonstration is a 16M-latent \glsterm{sparse-autoencoder}{SAE} trained on GPT-4 residual-stream
activations with $\sim 7\%$ dead latents under the AuxK
mitigation~\citep[\S 2.4]{gao2024-topk-saes}\footnote{KB excerpt:
\texttt{kb/excerpts/gao2024-topk-saes.md\#sec-2-4}.}; the law is
predictive enough to guide budget allocation between $N$ and training
tokens at the design stage rather than empirically. \citet{gemma-scope-2024}
exercise the architectural cousin --- JumpReLU SAEs of
\citet{rajamanoharan2024-jumprelu} --- at the open-weights end of the
same scaling envelope: more than $400$ SAEs spanning every layer and
sublayer of Gemma~2 2B and 9B (\glsterm{residual-stream}{residual stream}, MLP output, attention
output) plus selected layers of Gemma~2 27B, with widths
$\{16.4\text{K}, 65.5\text{K}, 131\text{K}, 1\text{M}\}$, totaling
$>\!30$M learned features and roughly $20\%$ of GPT-3's pretraining
compute~\citep[\S 1]{gemma-scope-2024}\footnote{KB excerpt:
\texttt{kb/excerpts/gemma-scope-2024.md\#sec-1}.}. The \glsterm{common-crawl}{CC}-BY-4.0 release
on HuggingFace plus the Neuronpedia interactive demo
\deepencite{Neuronpedia integration: \texttt{neuronpedia.org/gemma-scope},
not in papers.json} converted \glsterm{sparse-autoencoder}{SAE}-based interpretability from an
Anthropic-internal capability into an externally reproducible one.

\paragraph{Probes scale gracefully.}
A \glsterm{linear-probe}{linear probe} is trained on cached activations (so the model itself is
called once to produce activations, then the \glsterm{probe-2}{probe} runs on a CPU);
\glsterm{mass-mean-probing}{mass-mean probing} (\S\ref{sec:mass-mean}) is closed-form
$\mathbf{d}_{\mathrm{MM}} = \boldsymbol{\mu}_+ - \boldsymbol{\mu}_-$ and
requires no training at all~\citep{marks-tegmark-2023-truth}. \glsterm{probe-2}{Probe}
\emph{cost} is dominated by the activation cache, not the \glsterm{probe-2}{probe} fit;
the activation cache scales linearly with dataset size and is
embarrassingly parallel across prompts. The same property carries the
direction-ablation operator
$\mathbf{h} \mapsto \mathbf{h} - (\mathbf{h}^\top \hat{\mathbf{d}}_y)\,\hat{\mathbf{d}}_y$
and the addition operator $\mathbf{h} \mapsto \mathbf{h} + \alpha\,
\hat{\mathbf{d}}_y$ of \S\ref{sec:ablate-add}: both are $O(D)$ per token
and survive whatever architectural innovations the underlying model
adds, because they act on the \glsterm{residual-stream}{residual stream} rather than on any
particular sublayer.

\paragraph{Attribution graphs scale to production --- expensively, and
inside one lab.}
\citet{lindsey2025-circuit-tracing}'s cross-layer-\glsterm{transcoder}{transcoder} framework
reports the only published end-to-end mechanistic-interpretability
result on a production model (\glsterm{claude}{Claude} 3.5 Haiku). The compute argument
is two-part: per-task cost on the replaced model is one forward pass
plus one backward pass --- cheap relative to \glsterm{activation-patching}{activation patching}'s
$O(N_{\mathrm{components}})$ forward passes per (clean, corrupted) pair
(\S\ref{sec:patching-protocol}) --- but the amortized cost is the
\glsterm{cross-layer-transcoder}{cross-layer transcoder} training, estimated at the same order as \glsterm{sparse-autoencoder}{SAE}
training in \citet{gemma-scope-2024} ($\sim\! 20\%$ of the underlying
model's pretraining compute). The Anthropic 2025-05-30 open release of
the \glsterm{circuit}{circuit}-tracing tooling and its Neuronpedia integration
\deepencite{Anthropic open-source release of circuit-tracer 2025-05-30,
not in papers.json} converted the methodology from a strictly internal
capability into one external groups can in principle exercise on open
models. As of 2026, no externally-trained attribution-graph result on a
frontier model has been published; the contradiction this raises about
reproducibility is the load-bearing one of \S\ref{sec:contradictions}.

\subsection{What does not scale (yet)}
\label{sec:scaling-no}

\paragraph{Manual circuit hunts.}
The IOI reverse-engineering of \citet{wang2022-ioi} took graduate-student
months of work to identify $26$ heads in $7$ functional classes inside a
$1.6\%$-of-(head, position)-pairs subgraph on a $124$M-parameter model.
The same effort would need to scale roughly with the number of heads
and layers --- $L \cdot H$ on a frontier-2026 model is $\sim 80$--$120$
times that of GPT-2 small --- and the exhaustive case-by-case behavioral
testing that gave the IOI claim its load-bearing status does not have an
obvious automated analog. \citet{conmy2023-acdc} automate the
\emph{discovery} step (iterative path-patching with a KL-divergence
threshold), but the automation produces larger and less interpretable
subgraphs, and the human-in-the-loop labeling that decided ``this head
class is a Name Mover'' on IOI does not transfer.

\paragraph{Full activation patching.}
The three-run protocol of \S\ref{sec:patching-protocol} requires
$O(N_{\mathrm{components}})$ forward passes per (clean, corrupted) pair
to fill a causal-trace heatmap. On GPT-2 XL, $N_{\mathrm{components}}$
is order-$10^4$ at head granularity and $10^5$ at single-(head,
position); on a frontier-2026 model with $\sim 80$ layers, $\sim 64$
heads per layer, and $\sim 8\text{K}$ context, full-coverage patching
is on the order of $10^7$ forward passes per task. \glsterm{attribution-patching}{Attribution patching}
\deepencite{nanda2023-attribution-patching blog post; not in
papers.json --- this is a tier-B forum signal that needs deepening}
linearizes the \glsterm{indirect-effect}{indirect effect} via the
first-order Taylor approximation
$\mathrm{IE}_C \approx \langle \nabla_{\mathbf{h}} m,
\Delta \mathbf{h} \rangle$
(\S\ref{sec:patching-attribution}) and reduces this to a small constant
number of passes, but the linearization fails in non-linear regions of
the loss landscape, so it is a discovery aid rather than a faithful
substitute for direct intervention.

\paragraph{Lens coverage of frontier-2026 architectures.}
\citet{belrose2023-tuned-lens}'s tuned-lens benchmarks cover Pythia
410M--12B and GPT-2 / GPT-Neo / BLOOM; \texttt{logitlens4llms-2025}
extends the framework to Qwen-2.5 and LLaMA-3.1
(\S\ref{sec:lens-trajectory}). Tuned lenses for the rest of the
frontier-2026 lineup --- LLaMA 3/4, full Qwen 2.5/3, \glsterm{precision-pinning}{DeepSeek-V3},
Gemma 3 --- are not all published as of 2026. The relevant
architectural shift for the lens framework is not raw scale (the
distillation loss
$\mathcal{L}(\ell) = \E_{\mathbf{x}}[D_{\mathrm{KL}}(\mathcal{M}_{>\ell}
(\mathbf{h}_\ell) \| \mathrm{TunedLens}_\ell(\mathbf{h}_\ell))]$ is
indifferent to model size) but the substrate: \glsterm{precision-pinning}{DeepSeek-V3} combines \glsterm{multi-head-latent-attention}{MLA}
\citep{deepseek-v2}'s low-rank-compressed KV with fine-grained MoE
\citep{deepseekmoe2024}; \glsterm{nsa}{NSA} \citep{yuan2025} bakes attention sparsity
into pretraining. The \glsterm{residual-stream}{residual stream} remains well-defined in all of
these, but the assumption that a single per-layer affine translator
$(A_\ell, \mathbf{b}_\ell)$ suffices to absorb basis drift may need
re-examination on a per-architecture basis. The empirical work that
would close this gap --- training tuned lenses across the frontier-2026
lineup and reporting the per-architecture distillation-loss curves ---
is not yet in the literature.

\begin{contradiction}
\textbf{The interpretability-coverage gap is widening, not closing.}
The frontier-2026 architecture snapshot (Paper~1, §13) catalogs \glsterm{multi-head-latent-attention}{MLA},
fine-grained MoE, \glsterm{nsa}{NSA}-sparse attention, sliding-window layers, and
hybrid attention/\glsterm{state-space-model}{SSM} stacks as the substrate of the deployed lineup.
The lens benchmarks of \citet{belrose2023-tuned-lens} predate this
substrate; the SAEs of \citet{gemma-scope-2024} and the attribution
graphs of \citet{lindsey2025-circuit-tracing} cover specific models in
the lineup but not the lineup as a class; manual circuits and full
\glsterm{activation-patching}{activation patching} are exhibited only on $\le$1.5B-parameter
GPT-family models. The gap between ``what frontier models do'' and
``what we have interpretability evidence about frontier models doing''
is not narrowing on net: it is widening, because architectural
innovation outpaces the per-architecture interpretability work needed
to catch up. The 2025-05-30 Anthropic open release of \glsterm{circuit}{circuit}-tracing
tooling and the Gemma Scope \glsterm{common-crawl}{CC}-BY-4.0 release are corrective forces;
whether they correct fast enough is a load-bearing empirical question
about the field's coverage trajectory.
\end{contradiction}

\subsection{Production interpretability pipelines, 2026}
\label{sec:scaling-pipelines}

Three pipelines are in active use as of 2026, each anchored on one of
the methods above.

\begin{forumsignal}
\textbf{Source caveat for the Anthropic pipeline.} The cross-layer-\glsterm{transcoder}{transcoder}
+ Jacobian-attribution methodology rests on
\citet{lindsey2025-circuit-tracing} as its sole canonical reference;
no external lab has published an independent reproduction at frontier
scale as of 2026-Q2. The pipeline below is described as a productionised
tool, but the load-bearing claim that this is the productionised
attribution-graph methodology rests on a single tier-B Anthropic source.
\end{forumsignal}

\paragraph{Anthropic --- circuit tracer + Neuronpedia.}
The cross-layer-\glsterm{transcoder}{transcoder} + Jacobian-attribution pipeline of
\citet{lindsey2025-circuit-tracing}, open-sourced in 2025-05-30 and
integrated with Neuronpedia for interactive \glsterm{feature}{feature} browsing
\deepencite{Anthropic 2025-05-30 open-source release; Neuronpedia
integration; not in papers.json}. The substantive findings on \glsterm{claude}{Claude}
3.5 Haiku --- forward-planning rhyme features, multi-step factual
chains (``Texas'' $\to$ ``Austin''), refusal-\glsterm{circuit}{circuit} features,
partial-cross-lingual \glsterm{feature}{feature} overlap --- are reported in
\citet{lindsey2025-biology} and remain Anthropic-internal as evidence
\deepencite{lindsey2025-biology HTML-only; verbatim transcription
pending Phase 2}.

\paragraph{OpenAI --- TopK SAEs and the $L(C), L(N, K)$ scaling
program.}
The $L(C)$ and $L(N, K)$ laws of \citet{gao2024-topk-saes} are the
ongoing methodological backbone, plus an evaluation-metric suite (
$L_0$, reconstruction MSE, downstream-loss recovery, contrastive \glsterm{probe-2}{probe}
recovery)~\citep[\S 1 contributions]{gao2024-topk-saes}\footnote{KB
excerpt: \texttt{kb/excerpts/gao2024-topk-saes.md\#sec-1-headline}.}.
The 16M-latent GPT-4 \glsterm{sparse-autoencoder}{SAE} is the existence proof for \glsterm{sparse-autoencoder}{SAE} scaling at
production-frontier scale; the per-\glsterm{feature}{feature} interpretability work
downstream of that release is not all public.

\paragraph{Google DeepMind --- Gemma Scope + JumpReLU + tuned-lens
benchmarks.}
The \glsterm{jumprelu-sae}{JumpReLU SAE} family of \citet{rajamanoharan2024-jumprelu}, the
Gemma Scope open release of \citet{gemma-scope-2024}, and the Gemma 2
2B/9B/27B coverage make this the most externally-reproducible pipeline
of the three. The tuned-lens benchmarks of
\citet{belrose2023-tuned-lens}, retargeted across the Gemma family,
provide the cross-method readout layer the lens-as-decoder framing
needs.

\subsection{Snapshot table}
\label{sec:scaling-snapshot-table}

Table~\ref{tab:method-scaling} consolidates the per-method coverage as
of 2026-05. ``Open release'' is the date the method's
canonical-scale-target artifact (\glsterm{sparse-autoencoder}{SAE} weights, lens \glsterm{parameters}{parameters},
attribution-graph tooling) became externally usable; ``2026
reproduction'' is whether at least one externally-validated result on
a frontier-2026 model exists.

\begin{table}[t]
\centering
\caption{Interpretability methods --- scale, target, and coverage as of
2026-05. \emph{Target} is the largest model the method has been
reported on. \emph{Open release} is the date the canonical
scale-target artifact became externally available; ``\,---\,'' if no
such release. \emph{2026 reproduction} marks whether the
canonical-scale finding has been independently reproduced outside the
originating lab on a frontier-2026 model.}
\label{tab:method-scaling}
\small
% Six-column layout. The first 5 were `l` (no-wrap, natural width) which
% overflowed the page right edge once the long-text rows landed —
% "no fixed artifact (decoding rule)" and similar phrases in the
% Open-release column don't fit on a letter-page width. Switch to
% explicit p{} widths that wrap; widths sum to ~16cm matching the
% \textwidth on a letterpaper 1in-margin layout.
\begin{tabularx}{\textwidth}{@{}>{\raggedright\arraybackslash}p{2.0cm}
                                >{\raggedright\arraybackslash}p{2.5cm}
                                >{\raggedright\arraybackslash}p{2.0cm}
                                >{\raggedright\arraybackslash}p{1.3cm}
                                >{\raggedright\arraybackslash}p{3.2cm}
                                >{\raggedright\arraybackslash}X@{}}
\toprule
Method & Canonical paper & Target model & Scale & Open release &
2026 reproduction \\
\midrule
\glsterm{logit-lens}{Logit lens} & \citet{nostalgebraist2020-logit-lens} \deepencite{nostalgebraist2020-logit-lens
LessWrong post; not in papers.json} & GPT-2 family & $\le$1.5B &
no fixed artifact (decoding rule) & yes (GPT-2, BLOOM, GPT-Neo);
fails on BLOOM/GPT-Neo per \citet{belrose2023-tuned-lens} \\
\glsterm{tuned-lens}{Tuned lens} & \citet{belrose2023-tuned-lens} & Pythia 12B; GPT-Neo X &
$\le$20B & training script + benchmarks (2023) &
partial: \texttt{logitlens4llms-2025} on Qwen-2.5 / LLaMA-3.1; no
Gemma 3 / \glsterm{precision-pinning}{DeepSeek-V3} / LLaMA-4 published \\
Mass-mean \glsterm{probe-2}{probe} & \citet{marks-tegmark-2023-truth} &
LLaMA-2-13B & 13B & code on GitHub (2023) & yes for
truth, \glsterm{sycophancy}{sycophancy} variants on LLaMA family; OOD generalization
unresolved (\S\ref{sec:truth-frontier}) \\
\glsterm{relu}{ReLU}+L1 \glsterm{sparse-autoencoder}{SAE} & \citet{templeton2024-scaling-monosemanticity}
\deepencite{templeton2024-scaling-monosemanticity HTML-only at
transformer-circuits.pub} & \glsterm{claude}{Claude} 3 Sonnet & frontier &
no public weights (Anthropic-internal) & no \\
\glsterm{topk-sae}{TopK SAE} & \citet{gao2024-topk-saes} & GPT-4 & 16M-latent & training
recipe + evaluation suite (2024) & yes at $\le$Gemma 2 27B scale;
no GPT-4-class external reproduction \\
\glsterm{jumprelu-sae}{JumpReLU SAE} & \citet{rajamanoharan2024-jumprelu};
\citet{gemma-scope-2024} & Gemma 2 27B & 1M-latent & Gemma Scope
weights, \glsterm{common-crawl}{CC}-BY-4.0, HF (2024-08) & yes (most reproducible pipeline) \\
\glsterm{activation-patching}{Activation patching} & \citet{meng2022-rome};
\citet{zhang2023-apatching} & GPT-2 XL; LLaMA-7B
\deepencite{zhang2023-apatching scale extension; PDF-grounded but no
frontier replication} & 1.5B--7B & toolkits (TransformerLens, etc.)
& yes for $\le$7B; full-coverage at frontier scale infeasible
without \glsterm{attribution-patching}{attribution patching} \\
Manual \glsterm{circuit}{circuit} hunt & \citet{wang2022-ioi};
\citet{conmy2023-acdc} & GPT-2 small & 124M & ACDC code
(2023) & yes on GPT-2 small; not yet on frontier-2026 models \\
\glsterm{cross-layer-transcoder}{Cross-layer transcoder} + \glsterm{attribution-graph}{attribution graph}
& \citet{lindsey2025-circuit-tracing} & \glsterm{claude}{Claude} 3.5 Haiku & frontier &
\glsterm{circuit}{circuit}-tracer tooling (2025-05-30)
\deepencite{Anthropic open-source release 2025-05-30; Neuronpedia
integration; not in papers.json} & no external reproduction as of
2026 \\
\glsterm{natural-language-autoencoder}{NLA} (\glsterm{activation-verbalizer}{AV} + \glsterm{activation-reconstructor}{AR} pair)
& \citet{anthropic2026-nla} & \glsterm{claude}{Claude} Opus 4.6 & frontier &
no public weights; method paper (2026-05-07) at
\glsterm{transformer}{transformer}-circuits.pub/2026/nla & no external reproduction as of
2026; only Anthropic-internal Opus 4.6 audit application reported \\
\bottomrule
\end{tabularx}
\end{table}

The table makes the asymmetry explicit. The methods with the most
favorable scaling --- probes and SAEs --- have the most external
reproduction; the method with the strongest substantive findings on a
production frontier model --- attribution graphs --- has none. Three
of the nine rows depend on tooling or claims that are HTML-only or
released-but-not-yet-reproduced; the remaining six form a defensible
shared methodological substrate.

\begin{intuition}
The five-method picture maps cleanly to the scaling table. Lenses give
you a cheap readout that may not generalize across architectures (no
fixed artifact, partial coverage). Probes give you cheap
correlational evidence with closed-form variants and broad reproduction
(\S\ref{sec:mass-mean}). SAEs give you a \glsterm{feature}{feature} dictionary at scale
under documented laws (the most reproducible pipeline). Activation
patching gives you causal claims at small scale and degrades to
linearized approximations beyond. \glsterm{circuit-tracing}{Circuit tracing} gives you the
strongest claims at frontier scale but only inside the originating
lab. The methodological lesson of \S\ref{sec:cross-method} ---
method-pluralism is principled because methods answer different
questions --- has a scaling-table corollary: \emph{method-pluralism
is also operationally necessary, because no single method covers every
(model, behavior, evidence-class) cell of the matrix}. Returning to
canonical form: the per-method costs are $O(D)$ for probes and lenses,
$O(M D \cdot \text{tokens})$ amortized for \glsterm{sparse-autoencoder}{SAE} training,
$O(N_{\mathrm{components}})$ per-task for \glsterm{activation-patching}{activation patching}, and
$O(\text{transcoder training} + 1\,\text{fwd} + 1\,\text{bwd})$ for
attribution graphs.
\end{intuition}

\subsection{Forward link}
\label{sec:scaling-forward}

The substrate this section reads against is the frontier-2026
architecture snapshot of \emph{Paper~1, §13}: the \glsterm{multi-head-latent-attention}{MLA} / fine-grained
MoE / \glsterm{nsa}{NSA}-sparse-attention / sliding-window / hybrid-\glsterm{state-space-model}{SSM} stack that
defines what production models look like as of 2026. The
interpretability methods catalogued here have caught up to that
substrate unevenly, and the two most pressing gaps --- per-architecture
tuned lenses for the frontier lineup, and external reproduction of
attribution-graph findings outside Anthropic --- are open empirical
questions whose closure is the work of the next two years. The
contradictions surfaced in this section --- \glsterm{circuit}{circuit}-tracing
reproducibility, lens cross-architecture coverage, the widening gap
between deployed-model architecture and per-architecture
interpretability evidence --- are taken up alongside the
\glsterm{sparse-autoencoder}{SAE}-canonicality and truth-direction contradictions in
\S\ref{sec:contradictions}, where each is paired with the experiment
that would resolve it. The thesis of the paper holds against this
scaling picture: ``\emph{partially read}'' is exactly the right
phrase, and the qualifier is load-bearing across method, scale, and
architecture.
