% Paper 2, §2 — Pre-training data
% Anchors: kb/notes/training/{pre-training-data,synthetic-data-and-distillation}.md
%          kb/excerpts/{fineweb2024,data-mixing-laws-2024,phi4,
%                       deepseek-v3-training}.md

\section{Pre-training data}
\label{sec:pre-training-data}

Pre-training data is no longer ``the corpus'' --- a single dump of
filtered web text against which architecture and optimizer choices
are evaluated. It is a pipeline with its own engineering surface,
its own ablation regime, and its own scaling law. The 2024 inflection
point is FineWeb \citep{fineweb2024}, which decomposes the dataset
into a chain of stages (web extraction, language and quality
heuristics, deduplication, custom filters), each judged by training a
$1.71$B-parameter / $\sim 28$B-token ``ablation model'' on the
candidate output and comparing downstream task scores
\citep[\S 3.1]{fineweb2024}.\footnote{KB excerpt:
\texttt{kb/excerpts/fineweb2024.md\#sec-3-1-proxy}.} The mixture
choice on top of those curated domains has its own functional form:
\citet{data-mixing-laws-2024} fit a closed-form relation
$L_i(\mathbf{r}) = c_i + k_i \exp(\sum_j t_{ij} r_j)$ between
mixture proportions and per-target-domain validation loss, predicting
loss on unseen mixtures from a small number of fitted runs
\citep[\S 1, eq.\ 1]{data-mixing-laws-2024}.\footnote{KB excerpt:
\texttt{kb/excerpts/data-mixing-laws-2024.md\#sec-1-functional}.} The
synthetic-data turn closes the loop: Phi-4 \citep{phi4} reports
$\sim 400$B synthetic tokens spanning $50$ generation pipelines,
$14$B parameters surpassing GPT-4o on STEM evaluations
\citep[\S 1, Table 1]{phi4}, and explicitly argues that synthetic
sequences are a \emph{lower-noise} next-token-prediction signal than
organic web text.

The thesis of this section is that the curate $\to$ ablate $\to$ mix
loop has displaced ``find more web text'' as the active design
surface. We walk the FineWeb pipeline end-to-end with concrete
numbers, transcribe the Data Mixing Law equation and its coefficient
interpretation, present FineWeb-Edu's distill-then-classify pattern,
and set up two contradictions that resolve in
\cref{sec:contradictions-frontier}: the global-vs-individual-snapshot
deduplication result that disproves ``more dedup is better''
\citep[\S 3.4]{fineweb2024} and the unresolved synthetic-fraction
question raised by Phi-4's $\sim\!80\%$-synthetic mix versus
Tülu-3's organic-heavier post-training data
\citep{tulu3}. The section closes with the token-budget choice that
\cref{sec:scaling-laws} immediately operates on: Chinchilla's
$D \approx 20\,N$ versus Llama-3-8B's $D / N \approx 1875$
over-train, and the precision-aware tension with \cref{sec:mixed-precision}
that the over-train objective creates.

\subsection{The two design problems and their symbols}
\label{sec:data-setup}

A pre-training dataset is a finite multiset $\mathcal{D} =
\bigsqcup_{j=1}^{M} \mathcal{D}_j$ partitioned into $M$ source domains
(web crawl, code, math, books, scientific papers, etc.). The
\emph{mixture} is a probability vector $\mathbf{r} = (r_1, \dots,
r_M)$ on the simplex $\Delta^{M-1}$, with $r_j \geq 0$ and
$\sum_{j=1}^{M} r_j = 1$, where $r_j$ is the fraction of the
per-step token budget drawn from $\mathcal{D}_j$
\citep[\S 2]{data-mixing-laws-2024}. Two design problems sit on top
of this object:

\begin{enumerate}
\item \textbf{Per-domain curation.} What counts as $\mathcal{D}_j$,
how is it filtered, deduplicated, and quality-scored, what tokens
make it into the multiset.
\item \textbf{Mixture choice.} Given curated domains, what
$\mathbf{r}$ minimizes loss / maximizes downstream metric at the
target $(N, D)$ scale.
\end{enumerate}

The two problems compose: curation determines what each
$\mathcal{D}_j$ contains; mixing determines how the per-step token
budget is partitioned across them. \Cref{tab:data-shapes} fixes
notation for the rest of the section.

\begin{table}[h]
\centering
\small
\begin{tabular}{lll}
\toprule
Symbol & Type & Meaning \\
\midrule
$\mathcal{D}$ & multiset & full pre-training corpus \\
$\mathcal{D}_j$ & multiset & source-$j$ domain partition, $j \in \{1, \dots, M\}$ \\
$M$ & $\Z_+$ & number of source domains in the mixture \\
$\mathbf{r}$ & $\Delta^{M-1}$ & mixture proportions, $r_j \geq 0$, $\sum_j r_j = 1$ \\
$L_i(\mathbf{r})$ & $\R$ & validation loss on held-out target domain $i$ \\
$c_i, k_i$ & $\R$ & per-target-domain irreducible / amplitude parameters \\
$t_{ij}$ & $\R$ & transfer coefficient from train-domain $j$ to val-domain $i$ \\
$N$ & $\Z_+$ & model parameters (\cref{sec:scaling-laws}'s $N$) \\
$D$ & $\Z_+$ & total training tokens \\
$D / N$ & $\R_+$ & tokens-per-parameter ratio (\cref{tab:frontier-NDC}) \\
\bottomrule
\end{tabular}
\caption{Symbols used in \cref{sec:pre-training-data}. Mixture
$\mathbf{r}$ lives on the simplex $\Delta^{M-1}$; the Data Mixing
Law \cref{eq:dml} fits per-target-domain $(c_i, k_i, t_{i,1\ldots M})$
from a small set of pilot runs, then predicts $L_i(\mathbf{r})$ on
unseen mixtures.}
\label{tab:data-shapes}
\end{table}

\subsection{The FineWeb pipeline, end to end}
\label{sec:data-fineweb}

\citet{fineweb2024} document the design choices behind a
$15$T-token English web corpus derived from $96$ Common Crawl
snapshots (2013-summer through 2024-04)
\citep[\S 3.4]{fineweb2024}.\footnote{KB excerpt:
\texttt{kb/excerpts/fineweb2024.md\#sec-3-4-minhash}.} Each pipeline
choice is judged by an \emph{ablation model} of $1.71$B parameters
trained on $\sim 28$B tokens (roughly the Chinchilla-optimal
$D \approx 20\,N$ point for that $N$) under fixed architecture,
optimizer, batch size, and seed --- the only thing that varies is
the data version
\citep[\S 3.1]{fineweb2024}.\footnote{KB excerpt:
\texttt{kb/excerpts/fineweb2024.md\#sec-3-1-proxy}.} ``We trained
two models for each dataset version, each using a different but
equal-sized random subset of the full data and a different
initialization seed, and then compared average scores''
\citep[\S 3.1]{fineweb2024}. The downstream-task score \emph{slope}
under this small-scale protocol is the only signal the pipeline
trusts.

\paragraph{Stage 1 --- text extraction (trafilatura over WET).}
Common Crawl ships two formats: raw WARC and pre-extracted WET.
\citet[\S 3.2]{fineweb2024}\footnote{KB excerpt:
\texttt{kb/excerpts/fineweb2024.md\#sec-3-2-text}.} report that
``WET files retained too much boilerplate and menu text,'' and that
re-extracting from WARCs with the open-source \texttt{trafilatura}
library yields $\sim 25\%$ more text and a measurable ablation-model
gain. Custom text extraction is computationally expensive; the
ablation lift justifies the cost.

\paragraph{Stage 2 --- base filter.}
The base-filter pipeline reuses RefinedWeb's setup
\citep[\S 3.3]{fineweb2024}:\footnote{KB excerpt:
\texttt{kb/excerpts/fineweb2024.md\#sec-3-3-base}.} URL block-list
for adult content, a fastText language classifier keeping only
English with score $\geq 0.65$, and the quality / repetition filters
from MassiveText at original thresholds. Applied across the $96$
snapshots, this leaves ``roughly $36$ trillion tokens of data when
tokenized with the GPT-2 tokenizer''
\citep[\S 3.3]{fineweb2024}. The $0.65$ language-ID threshold is a
load-bearing decision: pushing higher cuts low-confidence multilingual
content (which often \emph{is} good English with code-switched
proper nouns); pushing lower admits genuinely off-language documents.
At $0.65$ the ablation models lose nothing detectable on standard
English-language benchmarks.

\paragraph{Stage 3 --- MinHash deduplication, individually per snapshot.}
\citet[\S 3.4]{fineweb2024}\footnote{KB excerpt:
\texttt{kb/excerpts/fineweb2024.md\#sec-3-4-minhash}.} compute
MinHash signatures from each document's $5$-grams using $112$ hash
functions arranged as $14$ buckets of $8$ hashes each. Two documents
are flagged as duplicates if they collide in any of the $14$ buckets
on all $8$ hashes simultaneously. The Jaccard threshold this scheme
targets is given by the standard MinHash analysis: a pair of
documents with Jaccard similarity $J$ collides in a single bucket of
$8$ hashes with probability $J^{8}$, and at least one of $14$
buckets agrees with probability
\begin{equation}
P(\text{flagged duplicate} \mid J) \;=\; 1 - (1 - J^{8})^{14},
\label{eq:minhash}
\end{equation}
which crosses $0.5$ at $J \approx 0.75$
\citep[\S 3.4]{fineweb2024}, hence ``targeting documents that are
at least $75\%$ similar.'' Flagged pairs are clustered transitively
(if $A \sim B$ and $B \sim C$, then $A, B, C$ form one cluster), and
one document per cluster survives.

\begin{intuition}
\label{intuit:minhash}
MinHash trades exactness for parallelism: instead of computing
$O(|\mathcal{D}|^2)$ pairwise Jaccard similarities, each document is
hashed into $14$ small buckets (8 hashes each) and only intra-bucket
collisions are checked. The bucket-based collision rule
\cref{eq:minhash} implements a cheap probabilistic threshold on the
true Jaccard. Returning to canonical form: with $14 \times 8$
hashes, pairs at $J = 0.9$ collide $> 99\%$ of the time, pairs at
$J = 0.5$ collide $\sim 5\%$ of the time, and the curve crosses $0.5$
at $J \approx 0.75$. The $14 \times 8$ choice is the
\emph{precision/recall} dial on this curve --- more buckets sharpen
recall at fixed Jaccard, more hashes per bucket sharpen precision.
\end{intuition}

\paragraph{The global-vs-individual deduplication finding.}
The naive prior holds that more aggressive dedup is monotonically
better. \citet[\S 3.4]{fineweb2024}\footnote{KB excerpt:
\texttt{kb/excerpts/fineweb2024.md\#sec-3-4-global}.} directly
disprove it. Their first attempt was global MinHash dedup across all
$96$ snapshots simultaneously. ``When applied to the oldest
snapshots, this process removed as much as $90\%$ of the original
base filtered data,'' and the ablation models trained on the survivors
were \emph{worse}, not better, than models trained on individually-
deduplicated snapshots. The interpretation: documents that appear in
multiple snapshots are over-represented because they are useful ---
they are linked, mirrored, archived, and reposted. Global dedup
removes these, leaving a residue dominated by content that appeared
only \emph{once}, which on average is lower-quality (ad copy, SEO
spam, transient pages). FineWeb's final pipeline therefore
deduplicates each snapshot independently, leaving cross-snapshot
duplicates intact, producing $\sim 20$T tokens before custom filters
\citep[\S 3.4]{fineweb2024}.

\begin{contradiction}
\label{contra:global-vs-indiv-dedup}
``More deduplication is monotonically better'' is empirically false
on FineWeb-scale data \citep[\S 3.4]{fineweb2024}. Global MinHash
across $96$ Common Crawl snapshots makes the trained model
\emph{worse} than per-snapshot MinHash, because what is
duplicated-across-snapshots is signal (linkage, archival, mirroring)
and what is unique-per-snapshot tilts toward noise. The result is
single-source on FineWeb's specific corpus and ablation protocol;
whether it generalizes to non-web corpora (curated books, code) or
to other dedup metrics (suffix-array exact, semantic-embedding
fuzzy) is open. We return to this in
\cref{sec:contradictions-frontier} alongside the
\cref{sec:scaling-laws} over-training contradiction; both turn on
the question of what ``more'' means once a pipeline is composed.
\end{contradiction}

\paragraph{Stage 4 --- C4 filters minus the curly-brace rule.}
\citet[\S 3.7]{fineweb2024} apply ``a selection of C4 filters'' on
top of the deduplicated text. The C4 paper's terminal-punctuation
and short-line rules largely transfer; FineWeb \emph{omits} C4's
``curly-brace'' filter, which would remove most code-bearing
documents and which their ablation showed to hurt downstream
benchmarks (presumably because code documents help English
benchmarks via cross-domain transfer \citep[\S 3.2,
$t_{ij}$]{data-mixing-laws-2024}). The decision is local --- one
ablation comparison --- but illustrates the protocol: every C4
filter was tested individually, and only those whose removal hurt
downstream were retained.

\paragraph{Stage 5 --- custom filters from histogram inspection.}
The largest single filter-design lift comes from
\citet[\S 3.6]{fineweb2024}.\footnote{KB excerpt:
\texttt{kb/excerpts/fineweb2024.md\#sec-3-6-custom}.} They collect
$50$ document-level statistics (line length distribution, terminal
punctuation rate, character-level repetition, word-level repetition,
\dots) on a high-quality corpus (individually deduplicated 2013-48
snapshot) and a low-quality corpus (the same snapshot
\emph{globally} deduplicated, the residue from the previous
finding). For each statistic, they identify regions of the histogram
where the low-quality corpus has higher density than the high-quality
corpus, and convert those regions into filter thresholds. The three
filters that survived ablation are
\begin{align}
&\text{filter A:} \quad &&\frac{\#\{\text{lines ending with terminal punctuation}\}}{\#\{\text{lines}\}} \;\leq\; 0.12, \nonumber \\
&\text{filter B:} \quad &&\frac{\#\{\text{characters in duplicated lines}\}}{\#\{\text{characters}\}} \;\geq\; 0.10, \nonumber \\
&\text{filter C:} \quad &&\frac{\#\{\text{lines $\leq 30$ chars}\}}{\#\{\text{lines}\}} \;\geq\; 0.67. \nonumber
\end{align}
Documents satisfying \emph{any} of A, B, or C are removed. Together
the three remove $\sim 22\%$ of remaining tokens and lift the
aggregate ablation score by $\sim 1$ percentage point at the $28$B-
token scale \citep[\S 3.6]{fineweb2024}. The $0.12$ terminal-punctuation
threshold is notably looser than C4's original $0.30$ (which removed
$\sim 30\%$ of tokens vs.\ FineWeb's $10.14\%$); the gentler cut
preserves more code, technical text, and structured content while
still removing the worst SEO-spam densities.

\paragraph{Final pipeline: $15$T tokens.}
Combining the choices --- WARC + trafilatura, base filter,
per-snapshot MinHash, selected C4 filters, custom filters A--C ---
yields the $15$T-token FineWeb release
\citep[\S 3.7]{fineweb2024}. This number is the empirical asymptote
that several frontier labs have independently converged on: Llama 3
trained on $\sim 15$T tokens \citep{meta-llama3}, DeepSeek-V3 on
$14.8$T \citep[abstract]{deepseek-v3}.\footnote{KB excerpt:
\texttt{kb/excerpts/deepseek-v3-training.md\#abstract}.}

\begin{intuition}
\label{intuit:ablation-meter}
The $1.71$B / $28$B-token ablation model is too small to score
meaningfully against any production benchmark in absolute terms ---
its MMLU accuracy is below random-with-formatting-noise. The design
relies instead on the \emph{signed direction} of
$\Delta L(\text{candidate}_A, \text{candidate}_B)$ being stable
across scale: if a filter helps the ablation by $+0.5$pp on
HellaSwag, it should help the production model on the same task.
This holds for most heuristic decisions \citep[\S 3.1]{fineweb2024}.
The global-vs-individual dedup case in
\cref{contra:global-vs-indiv-dedup} is interesting precisely because
the ablation pointed toward the counterintuitive answer and the
counterintuitive answer kept holding at production scale. Returning
to canonical form: the protocol bets on the transfer relation
$\mathrm{sgn}\bigl(\Delta L(A, B; N{=}1.71\text{B}, D{=}28\text{B})\bigr) =
\mathrm{sgn}\bigl(\Delta L(A, B; N{=}\text{prod}, D{=}\text{prod})\bigr)$
holding component-wise across the pipeline's design choices.
\end{intuition}

\subsection{Data Mixing Laws --- the functional form on $\mathbf{r}$}
\label{sec:data-mixing}

Curation determines what each $\mathcal{D}_j$ contains; mixing
determines how training-step tokens partition across them.
\citet{data-mixing-laws-2024} establish that, under fixed model size
and training-token budget, the validation loss on held-out target
domain $i$ is a closed-form function of the training-mixture
proportions. Verbatim from
\citet[\S 1, eq.\ 1; \S 3.2, eq.\ 7]{data-mixing-laws-2024}\footnote{KB
excerpt: \texttt{kb/excerpts/data-mixing-laws-2024.md\#sec-1-functional}.}:
\begin{equation}
L_i(\mathbf{r}_{1\ldots M}) \;=\; c_i \,+\, k_i \,\exp\!\left(
   \sum_{j=1}^{M} t_{ij}\,r_j
\right),
\label{eq:dml}
\tag{1}
\end{equation}
where $L_i$ is validation loss on domain $i$, $\mathbf{r} \in
\Delta^{M-1}$ is the training mixture, and $(c_i, k_i, t_{i, 1\ldots
M})$ are scalar parameters fit per-target-domain. The form is the
unique exponential-of-linear shape compatible with two principles
\citep[\S 3.2]{data-mixing-laws-2024}:\footnote{KB excerpt:
\texttt{kb/excerpts/data-mixing-laws-2024.md\#sec-3-2-multi}.}

\begin{itemize}
\item \emph{Compatibility}: at $M = 2$, \cref{eq:dml} reduces to the
two-domain log-linear law $L_i(r) = c_i + k_i \exp(t_{ii} r)$
empirically verified on $70$M and $160$M models trained on
GitHub + Pile-CC at $30$B tokens
\citep[\S 3.1, eq.\ 6]{data-mixing-laws-2024}.\footnote{KB excerpt:
\texttt{kb/excerpts/data-mixing-laws-2024.md\#sec-3-1-pilot}.}
\item \emph{Symmetry}: any permutation of training-domain indices
should leave the functional form invariant; \cref{eq:dml} is
symmetric under permutations of $\{r_j, t_{ij}\}$ jointly.
\end{itemize}

\paragraph{Coefficient interpretation.}
\citet[\S 3.2]{data-mixing-laws-2024}\footnote{KB excerpt:
\texttt{kb/excerpts/data-mixing-laws-2024.md\#sec-3-2-coefficients}.}
read off the meaning of each fitted parameter:
\begin{itemize}
\item $c_i$ --- ``losses that are not reducible by adjusting the data
mixture,'' i.e., the irreducible loss for target domain $i$. No
choice of $\mathbf{r}$ on $\Delta^{M-1}$ can drop $L_i$ below $c_i$.
\item $k_i$ --- amplitude scaling the exponential term. At
$\sum_j t_{ij} r_j = 0$ (a flat-zero-sensitivity mixture), $L_i =
c_i + k_i$.
\item $t_{ij}$ --- ``the interaction'' between training domain $j$
and validation domain $i$. \emph{Negative} $t_{ij}$ means training
on more $j$ \emph{lowers} validation loss on $i$; \emph{positive}
$t_{ij}$ means training on $j$ \emph{hurts} $i$.
\end{itemize}

The off-diagonal $t_{ij}$ is the load-bearing object: it upgrades the
folklore that ``more code helps math'' from a heuristic to a
\emph{measurable per-pair coefficient}. A fitted
$t_{\text{math} \leftarrow \text{code}} < 0$ predicts how much
adding code tokens to the mix lowers held-out math validation loss,
and the prediction can be tested.

\paragraph{Nested scaling laws --- making the fit affordable.}
Fitting \cref{eq:dml} directly at the target $(N, D)$ requires
many full-scale runs and is unaffordable at frontier scale.
\citet[\S 1, \S 2]{data-mixing-laws-2024} stack a power-law of
training-step ($L = c + k\,S^{\alpha}$) inside a power-law of
model-size, inside the mixing law \cref{eq:dml}, so that small-model
short-step experiments at varied $\mathbf{r}$ predict large-model
end-of-training $L_i(\mathbf{r})$. The headline empirical
demonstration: a $1$B-parameter model trained for $100$B tokens on
RedPajama, with mixture chosen by the nested fit, reaches loss
matching the same model trained for \emph{$48\%$ more steps} on
RedPajama's default mixture
\citep[abstract]{data-mixing-laws-2024}.\footnote{KB excerpt:
\texttt{kb/excerpts/data-mixing-laws-2024.md\#abstract}.} Equivalently,
the mixing-law-tuned recipe matches the $148$B-token-budget loss at
$100$B tokens spent --- a $1.48\times$ data efficiency at fixed
target loss.

\begin{analogy}
\label{analogy:dml-as-response-surface}
\Cref{eq:dml} is a response-surface model on the simplex
$\Delta^{M-1}$: the per-target-domain loss is a fitted exponential
in $\mathbf{r}$, with the fit's coefficients carrying physical
meaning ($c_i$ as floor, $t_{ij}$ as transfer affinity). The pipeline
that uses it is straight design-of-experiments: pilot runs at a
diverse set of $\mathbf{r}$ values, fit $(c_i, k_i, t_{i, 1\ldots
M})$ per held-out $i$, then optimize the predicted $L_i$ at the
target $(N, D)$ via the nested scaling-law extrapolation. Returning
to canonical form: the design problem is
$\mathbf{r}^* = \arg\min_{\mathbf{r} \in \Delta^{M-1}}
\sum_i w_i\,\hat{L}_i(\mathbf{r})$ for target weights $w_i$ on each
held-out domain \citep[\S 2, eq.\ 5]{data-mixing-laws-2024}, with
$\hat{L}_i$ the nested-scaling-law extrapolation of \cref{eq:dml}.
This is data-mixture engineering as a constrained convex program,
not a heuristic.
\end{analogy}

\subsection{Quality classifiers --- FineWeb-Edu's distill-then-classify}
\label{sec:data-edu}

After heuristic filtering produces a $15$T-token web corpus, a
\emph{quality classifier} can extract a higher-density subset.
\citet[\S 4]{fineweb2024}\footnote{KB excerpt:
\texttt{kb/excerpts/fineweb2024.md\#sec-4-edu}.} demonstrate the
pattern at scale, producing FineWeb-Edu ($1.3$T tokens) from the full
$15$T base. The pipeline is four stages:

\begin{enumerate}
\item \textbf{Annotate.} Sample $460{,}000$ random documents from
FineWeb's CC-MAIN-2024-10 snapshot; prompt Llama-3-70B-Instruct to
score each on a $0$--$5$ ``educational quality'' rubric.
\item \textbf{Train.} Fit a small linear regressor on
\texttt{Snowflake-arctic-embed-m} embeddings of the documents,
targeting the Llama-3-70B scores. $20$ epochs, learning rate
$3 \times 10^{-4}$, embedding/encoder layers frozen
\citep[\S 4]{fineweb2024}. F1 = $82\%$ on validation at score
threshold $\geq 3$.
\item \textbf{Score.} Run the regressor over the full $15$T-token
FineWeb. Cost: $\sim 6{,}000$ H100 GPU hours.
\item \textbf{Threshold.} Keep documents with regressor score
$\geq 3$, yielding $\sim 1.3$T tokens.
\end{enumerate}

The result outperforms FineWeb on knowledge- and reasoning-intensive
benchmarks (MMLU, ARC, OpenBookQA) at matched
training-token-for-training-token comparison
\citep[abstract]{fineweb2024}. The pattern's economic structure is
\emph{amortized teacher cost}: the $\sim 6$K-H100-hour annotation
step is paid once at corpus-build time and recovered every step of
every downstream pre-training run that uses the filtered corpus.
This is the path that several frontier labs (Llama 3, Phi-3) used
internally before FineWeb-Edu published it.\deepencite{nemotron-cc-2024
\S abstract: NVIDIA's $6.3$T-token Nemotron-CC corpus uses an
analogous classifier-filtering plus rephrasing pipeline; cited as
the closest published parallel to FineWeb-Edu's distill-then-classify
recipe.} The KB note flags FineWeb-Edu, DCLM-Baseline, and
Nemotron-CC as a coherent 2024-class lineage of classifier-driven
distillation \citep[\S
1]{fineweb2024}.\footnote{KB note:
\texttt{kb/notes/training/pre-training-data.md\#variants}.}

\subsection{Synthetic-data-as-bulk --- the Phi-4 inversion}
\label{sec:data-synthetic}

FineWeb-Edu uses a teacher LLM as a \emph{labeler}: the teacher's
judgment is amortized into a frozen filter. Phi-4 \citep{phi4}
inverts the role: the teacher LLM becomes the \emph{generator} of a
substantial fraction of the pre-training corpus. The Phi-4 abstract
\citep[abstract]{phi4}\footnote{KB excerpt:
\texttt{kb/excerpts/phi4.md\#abstract}.} states the recipe explicitly:

\begin{quote}
phi-4 [\dots] strategically incorporates synthetic data throughout
the training process. While previous models in the Phi family largely
\emph{distill} the capabilities of a teacher model (specifically
GPT-4), phi-4 substantially \emph{surpasses} its teacher model on
STEM-focused QA capabilities, giving evidence that our data-generation
and post-training techniques go beyond distillation.
\end{quote}

Phi-4's pre-training and mid-training corpus contains $\sim 400$B
synthetic tokens generated through ``$50$ broad types of synthetic
datasets, each one relying on a different set of seeds and different
multi-stage prompting procedure''
\citep[\S 2.2]{phi4}.\footnote{KB excerpt:
\texttt{kb/excerpts/phi4.md\#sec-2-2-pipelines}.} Three pipeline
primitives recur:

\begin{enumerate}
\item \emph{Seed curation.} Web/code/book passages filtered for
educational and reasoning value, then segmented and re-scored.
Question banks are filtered by plurality voting --- discard
questions where teacher samples all agree (too easy) or entirely
disagree (too hard / ambiguous), keeping the middle band
\citep[\S 2.2]{phi4}.\footnote{KB excerpt:
\texttt{kb/excerpts/phi4.md\#sec-2-2-seeds}.}
\item \emph{Rewrite and augment.} Seeds transformed into exercises,
dialogues, and structured reasoning tasks via multi-step prompting.
\item \emph{Self-revision.} A model critiques and improves its own
output under reasoning / factual rubrics
\citep[\S 2.2]{phi4}.\footnote{KB excerpt:
\texttt{kb/excerpts/phi4.md\#sec-2-2-rewrite}.}
\end{enumerate}

The headline empirical claim: a $14$B-parameter Phi-4 surpasses its
teacher GPT-4o on the GPQA ($56.1$ vs.\ $50.6$) and MATH ($80.4$ vs.\
$74.6$) benchmarks \citep[\S 1, Table 1]{phi4}.\footnote{KB excerpt:
\texttt{kb/excerpts/phi4.md\#sec-1-stem}.}

\paragraph{The variance-of-next-token-loss intuition.}
\citet[\S 2.1]{phi4}\footnote{KB excerpt:
\texttt{kb/excerpts/phi4.md\#sec-2-1-purpose}.} argue that synthetic
data is \emph{lower-noise} as a next-token-prediction signal than
organic web text:

\begin{quote}
In organic datasets, the relationship between tokens is often
complex and indirect. Many reasoning steps may be required to
connect the current token to the next, making it challenging for the
model to learn effectively from next-token prediction. By contrast,
each token generated by a language model is by definition predicted
by the preceding tokens, making it easier for a model to follow the
resulting reasoning patterns. In this way, synthetic data may act as
a form of ``spoonfeeding,'' presenting challenges in a digestible
and progression-oriented manner.
\end{quote}

\begin{intuition}
\label{intuit:spoonfeeding}
The argument is a claim about the conditional variance of the
next-token loss. Let $\ell_t = -\log \pi_\theta(x_t \mid x_{<t})$ be
the per-token loss; on organic text, $\ell_t$ has high variance
because the local prefix may not contain enough information to
predict $x_t$ even at the data-generating distribution (long-range
dependencies, references to context not present in the window,
implicit world knowledge). On synthetic text generated by a related
language model under autoregressive sampling, $x_t$ is by
construction predictable from $x_{<t}$ at the teacher's
distribution, so the entropy
$H(x_t \mid x_{<t})$ at the data-generating distribution is lower.
Returning to canonical form: under the SFT-style decomposition of
\cref{sec:scaling-laws}'s parametric loss, this corresponds to a
smaller $E + B/D^{\beta}$ tail at fixed $D$ when $D$ is sourced from
$\pi_{\text{teacher}}$ rather than from organic web text --- the
\emph{irreducible-plus-finite-data} terms shrink because the
generating distribution is closer to a tractable one. The
``spoonfeeding'' analogy is exactly this: same number of training
tokens, lower per-token loss variance, faster effective learning.
\end{intuition}

\paragraph{The synthetic-fraction question.}
Phi-4 chose roughly $80\%$ synthetic tokens in its pre-training mix
for a $14$B reasoning-focused model
\citep[\S 1, \S 2.2]{phi4}.\footnote{KB note:
\texttt{kb/notes/training/synthetic-data-and-distillation.md\#sec-1-1}.}
At the post-training surface, Tülu-3 \citep{tulu3} ships an
organic-heavier mix on Llama-3 base, with synthetic generation used
selectively rather than as bulk. No published large-scale controlled
ablation isolates the optimal synthetic fraction across model scale,
domain, and training-stage (pre-train vs.\ mid-train vs.\ SFT).

\begin{contradiction}
\label{contra:synthetic-fraction}
The synthetic-fraction question splits the field. Phi-4
\citep[abstract]{phi4} argues for synthetic-as-bulk
($\sim\!80\%$ of pre-training tokens) on the
variance-of-next-token-loss grounds in
\cref{intuit:spoonfeeding}. Tülu-3 \citep{tulu3} ships an
organic-heavier post-training mix without claiming the Phi-4 ratio
generalizes. Recursive-synthetic concerns from the broader
literature \citep[\S 1]{phi4} warn that training on a model's own
outputs can degrade tail distributions
(model-collapse).\deepencite{shumailov2024-collapse \S 2: recursive
synthetic-only training accumulates loss in the data distribution's
tails; cited as the principal counter-evidence to
synthetic-as-bulk.} The synthetic-fraction at frontier scale is one
of the open questions
\cref{sec:contradictions-frontier} returns to: there is no
controlled $14$B-vs-$70$B Phi-style ablation, no
synthetic-pipeline-mixture analogue of \cref{eq:dml}, and no
published study isolating
recursive-synthetic-mode-collapse from the
quality-of-teacher-pipeline gain.
\end{contradiction}

\subsection{Token budget --- $D / N$ as the bridge to scaling}
\label{sec:data-budget}

Curation gives us $\mathcal{D}$; mixing gives us $\mathbf{r}$;
neither answers \emph{how many tokens to train on}. That choice
belongs to \cref{sec:scaling-laws}, but \cref{sec:pre-training-data}
must specify the input it provides.

\paragraph{Chinchilla's $D \approx 20\,N$.}
\citet[\S 4.1, Table 4]{hoffmann2022-chinchilla} fit the
compute-optimal frontier and report $D / N \approx 20$ at the
Gopher-scale FLOP budget, with a slow drift upward at larger scale
($D / N \approx 21.2$ at $1$T parameters)
\citep[\S 3, Table 3]{hoffmann2022-chinchilla}. The
\cref{sec:scaling-laws} parametric form
$\hat{L}(N, D) = E + A / N^{\alpha} + B / D^{\beta}$ has its
constrained minimum at $D \approx 20\,N$ for the fitted exponents.

\paragraph{The over-training inversion.}
The post-Chinchilla industry move is to substantially over-train
small dense models for inference economics. Llama-3 8B trained on
$\sim\!15$T tokens \citep{meta-llama3} sits at $D / N \approx 1875$
--- roughly $94\times$ over Chinchilla's recipe. The reasoning is
that the cost a deployed-model lab actually pays is
$C_{\text{train}} + N_{\text{queries}} \cdot C_{\text{inf-per-query}}$,
and when the second term dominates, the optimum shifts toward
smaller $N$ with proportionally more $D$
(\cref{tab:frontier-NDC}). Crucially, the parametric loss
$\hat{L}(N, D)$ keeps decreasing in $D$ at fixed $N$ --- just
suboptimally on the pre-training-FLOP margin. Over-training is not
a refutation of Chinchilla; it solves a different objective.

\paragraph{Forward link.}
Two threads of this section feed forward. \Cref{sec:scaling-laws}
takes $D$ and $N$ as the load-bearing variables of the
$L(N, D)$ functional form: the FineWeb $\sim 15$T-token web
ceiling and the Phi-4 $\sim 400$B synthetic supplement set the
\emph{numerator} of the over-training $D / N$ ratio.
\Cref{sec:mixed-precision} takes $D$ as input to the
precision-aware scaling-law tension: \citet{scaling-laws-precision-2024}
predict that over-trained models (large $D / N$) are
\emph{harder to quantize} at deployment time, partially offsetting
the inference-cost gain that motivates the over-training in the
first place \citep[\S 4]{scaling-laws-precision-2024}.\deepencite{scaling-laws-precision-2024 \S
4: effective-parameter form $N_{\text{eff}}(N, b_{\text{train}})$
and the over-training $\to$ harder-to-quantize prediction;
forward-referenced from \cref{sec:scaling-precision}.} The data
pipeline's choice of $D$ thus has a downstream cost that the
$\hat{L}(N, D)$ formula does not see.

\subsection{Variants and lineage}
\label{sec:data-lineage}

\Cref{tab:data-lineage} summarizes the open-corpus lineage from C4
through the 2024-class pipelines.

\begin{table}[h]
\centering
\small
\begin{tabular}{lrrl}
\toprule
Corpus & Tokens & Year & Distinctive choice \\
\midrule
C4                          & $0.75$T & 2020 & first open web pipeline; line-level rules \\
The Pile                    & $0.825$T & 2020 & $22$ hand-picked sources; manual mixture \\
RefinedWeb (Falcon)         & $5$T    & 2023 & web-only; aggressive global dedup \\
RedPajama-V1 / V2           & $1.2$T / $30$T & 2023/24 & open Llama-1 reproduction; V2 returns CC at scale \\
Dolma (OLMo)                & $3$T    & 2024 & $7$ open sources \\
\textbf{FineWeb}            & $\mathbf{15}$\textbf{T} & \textbf{2024} & \textbf{ablation-model-driven design} \citep{fineweb2024} \\
\textbf{FineWeb-Edu}        & $\mathbf{1.3}$\textbf{T} & \textbf{2024} & \textbf{distill-then-classify} \citep{fineweb2024} \\
DCLM-Baseline               & $4$T    & 2024 & fastText-on-instruction-data classifier-filtering \\
Nemotron-CC                 & $6.3$T  & 2024 & rephrased + classifier-filtered web (NVIDIA) \\
\midrule
DeepSeek-V3 (proprietary)   & $14.8$T & 2024 & \cite{deepseek-v3} reports diverse + high-quality \\
Llama-3 (proprietary)       & $\sim 15$T & 2024 & \cite{meta-llama3}; tokenizer 128K \\
Phi-4 (synthetic-bulk)      & $\sim 400$B syn & 2024 & $50$ pipelines \citep{phi4} \\
\bottomrule
\end{tabular}
\caption{Open-corpus lineage and selected proprietary contemporaries.
Three trends: ($i$) total tokens grew $\sim\!20\times$ in four years;
($ii$) the curation interface shifted from hand-picked sources
(Pile) to ablation-driven heuristics (FineWeb) to classifier-driven
distillation (FineWeb-Edu, DCLM, Nemotron-CC); ($iii$) multiple
labs now report $\sim 15$T as the diminishing-returns ceiling on
high-quality English web text \citep[\S
4]{deepseek-v3}.\protect\footnote{KB note:
\texttt{kb/notes/training/pre-training-data.md\#variants}.}}
\label{tab:data-lineage}
\end{table}

\subsection{Forward link}
\label{sec:data-forward}

Pre-training data is now a pipeline with three load-bearing surfaces:
curation, mixing, and synthetic-fraction. Curation is empirical
ablation: \citet{fineweb2024}'s $1.71$B / $28$B-token protocol
selects each filter on its sign-of-effect on downstream evaluation,
producing the $15$T-token FineWeb release through a chain of
WARC-trafilatura extraction, $0.65$ language-ID filtering, $14
\times 8$ MinHash deduplication \emph{per snapshot} (not globally,
\cref{contra:global-vs-indiv-dedup}), three quantile-derived custom
filters, and an optional Llama-3-70B distill-then-classify pass.
Mixing is a fittable functional form: \cref{eq:dml}'s
exponential-of-linear in $\mathbf{r}$, with per-target-domain
$(c_i, k_i, t_{ij})$ recoverable from a small set of pilot runs and
extrapolated to the target $(N, D)$ via nested scaling laws
\citep{data-mixing-laws-2024}. Synthetic-fraction is the open
contradiction (\cref{contra:synthetic-fraction}): Phi-4's
$\sim\!80\%$ for $14$B reasoning vs.\ Tülu-3's organic-heavier
post-training mix, with no controlled frontier-scale ablation.

The next section, \cref{sec:scaling-laws}, takes the token-count $D$
and parameter-count $N$ this section produces and asks how
cross-entropy loss scales in $(N, D, C)$, why Kaplan's
compute-allocation rule was overturned by Chinchilla's matched
LR-horizon re-fit, and why the frontier-2024 industry sits an order
of magnitude past Chinchilla on $D / N$ for inference economics.
\Cref{sec:mixed-precision} then takes the same $D$ and asks how the
precision $\times$ token-count interaction \citep[\S
4]{scaling-laws-precision-2024} pulls against over-training: the
larger $D / N$ a deployed model has consumed, the harder it is to
quantize at deployment time. The pipeline of this section is the
input to both.
