% Paper 2, §7 — Supervised fine-tuning
% Anchors: kb/notes/post-training/sft.md
%          kb/excerpts/{ouyang2022-instructgpt,phi4,deepseek-r1}.md

\section{Supervised fine-tuning}
\label{sec:sft}

Supervised fine-tuning (SFT) is the first stage of essentially every
modern post-training pipeline: cross-entropy maximum-likelihood on
curated prompt-response pairs, producing the instruction-following
starting point $\pi^{\mathrm{SFT}}$ that every subsequent stage
(\cref{sec:rlhf} PPO, \cref{sec:dpo} \glsterm{dpo}{DPO}, \cref{sec:rlvr-grpo} \glsterm{grpo-2}{GRPO})
takes as both initialization and KL anchor
\citep[\S 3]{ouyang2022-instructgpt}.\footnote{KB excerpt:
\texttt{kb/excerpts/ouyang2022-instructgpt.md\#sec-3}.} The thesis of
this section is the one most easily lost in the post-training
literature: \emph{the loss is plain cross-entropy with prompt-masking,
and the interesting research questions live in the data, not the
objective.} The lineage from \glsterm{flan}{FLAN}'s $\sim\!1.8$K reformatted NLP tasks
\citep{wei2021-flan,chung2022-flan-t5} through InstructGPT's $\sim\!13$K
labeler-written demonstrations \citep[\S 3]{ouyang2022-instructgpt} to
T\"ulu-3's $\sim\!939$K mixed-source set \citep{tulu3}, \glsterm{magpie}{MAGPIE}'s
self-synthesis at zero human cost \citep{magpie-2024}, R1-Distill's
$800$K reasoning-trace transfer \citep[\S 4]{deepseek-r1}, and s1's
1K-trace + budget-forcing endpoint \citep{s1-2025} is a story about
data, not loss. We walk that story, give the multi-turn loss-mask
conventions that production recipes have converged on, set up the
three open contradictions that \cref{sec:contradictions} discusses in
full, and end at the boundary of \cref{sec:rlhf}: why SFT precedes RL
in the canonical pipeline at all.

\subsection{The SFT objective}
\label{sec:sft-objective}

Given a dataset $\mathcal{D}_{\mathrm{SFT}} = \{(x^{(i)}, y^{(i)})\}_{i=1}^N$
of prompt-response pairs, SFT trains $\pi_\theta$ to minimize the
expected negative log-likelihood of the response conditional on the
prompt:
\begin{equation}
\mathcal{L}_{\mathrm{SFT}}(\pi_\theta)
  = -\E_{(x,y)\sim\mathcal{D}_{\mathrm{SFT}}}
       \!\big[\log \pi_\theta(y \mid x)\big],
\label{eq:sft-loss-canonical}
\end{equation}
verbatim from \citet[\S 3]{ouyang2022-instructgpt}.\footnote{KB excerpt:
\texttt{kb/excerpts/ouyang2022-instructgpt.md\#sec-3}.} The
implementation form, which is what training code actually computes,
expands the conditional log-probability and folds in a per-token loss
mask. Concatenate $z = (x, y)$ into one token sequence with the chat
template; let $m_t \in \{0, 1\}$ be the loss mask with $m_t = 1$ iff
$z_t$ is a response token. Then
\begin{equation}
\mathcal{L}_{\mathrm{SFT}}(\theta)
  = -\E_{z \sim \mathcal{D}_{\mathrm{SFT}}}
       \!\left[\sum_{t=1}^{|z|} m_t \, \log \pi_\theta(z_t \mid z_{<t})\right].
\label{eq:sft-loss-mask}
\end{equation}
\Cref{eq:sft-loss-mask} is \cref{eq:sft-loss-canonical} written so that
prompt-masking is explicit. The two are equivalent: the inner sum
restricted to $\{t : m_t = 1\}$ is exactly $\log \pi_\theta(y \mid x)$
under the standard left-to-right factorization
\citep[\S 3.5]{ouyang2022-instructgpt}.

\begin{table}[h]
\centering
\small
\begin{tabular}{ll}
\toprule
Symbol & Meaning \\
\midrule
$x$ & prompt tokens (instruction, system message, prior context) \\
$y = (y_1, \ldots, y_{|y|})$ & response tokens (assistant turn) \\
$z = (x, y)$ & concatenated training sequence after chat-template \\
$m_t \in \{0, 1\}$ & loss mask: $m_t = 1$ iff $z_t$ is a response token \\
$\pi_\theta$ & policy under training, initialized from a base LM \\
$\mathcal{D}_{\mathrm{SFT}}$ & SFT dataset, $|\mathcal{D}_{\mathrm{SFT}}| \in [10^3, 10^6]$ \\
$\pi^{\mathrm{SFT}}$ & resulting model after $1$--$4$ epochs of \cref{eq:sft-loss-mask} \\
\bottomrule
\end{tabular}
\caption{SFT symbol table. Scale numbers compiled from
\citet[\S 3.4]{ouyang2022-instructgpt} ($\sim\!13$K demonstrations),
\citet{tulu3} ($\sim\!939$K mixed), and \citet[\S 4]{deepseek-r1}
($800$K reasoning + non-reasoning).}
\label{tab:sft-symbols}
\end{table}

The hyperparameters that distinguish SFT from pre-training are not
algorithmic but quantitative
\citep[\S 3.5]{ouyang2022-instructgpt}: SFT runs $1$--$4$ epochs over
the curated set rather than a single pass; the learning rate is roughly
$5 \times 10^{-6}$, an order of magnitude below pre-training peak; the
mixed-precision recipe (BF16 working precision with FP32 \glsterm{master-weights}{master weights},
\cref{sec:mp-bf16}) is inherited unchanged. Architecturally there is
\emph{no} change — SFT is a continued-training stage, not a new model.
Total compute per epoch is $\sim 6 N_{\mathrm{tok}} P$ FLOPs in standard
mixed-precision training, with $N_{\mathrm{tok}} = \sum_i (|x^{(i)}| +
|y^{(i)}|)$, two to four orders of magnitude below pre-training.

\begin{intuition}
SFT is behavior cloning. The model is being shown $(x, y)$ pairs and
told ``next time you see $x$, write $y$.'' In RL terms,
$\mathcal{L}_{\mathrm{SFT}}$ is the negative log-likelihood of a
demonstrator policy $\pi^{\mathrm{demo}}$ under the model — the
labeler, the teacher LM, or the rejection-sampled own-rollout that
produced $y$. Returning to math: minimizing \cref{eq:sft-loss-canonical}
is equivalent, in the infinite-data limit, to minimizing the
\emph{forward} KL divergence
$\KL(\pi^{\mathrm{demo}} \,\|\, \pi_\theta)$
\citep[\S 3.5]{ouyang2022-instructgpt}. Forward KL is mode-covering:
the trained $\pi_\theta$ must put non-trivial mass wherever
$\pi^{\mathrm{demo}}$ does, or the loss diverges. \glsterm{rlhf}{RLHF} / \glsterm{dpo}{DPO}
(\cref{sec:rlhf}, \cref{sec:dpo}) instead minimize the \emph{reverse}
KL, $\KL(\pi_\theta \,\|\, \pi^{\mathrm{SFT}})$, which is mode-seeking:
$\pi_\theta$ may put zero mass where $\pi^{\mathrm{SFT}}$ has support
without paying a divergence cost. SFT and \glsterm{rlhf}{RLHF} are dual KL directions
around the demonstrator and the SFT-stage model respectively, which
is why the two stages compose cleanly: SFT broadens, RL sharpens.
\end{intuition}

\subsection{Why prompt-masking}
\label{sec:sft-mask}

Without the mask $m_t$ in \cref{eq:sft-loss-mask}, the SFT loss
includes $\sum_t \log \pi_\theta(x_t \mid x_{<t})$ — the likelihood of
the prompt itself. Two things go wrong. First, for a base model the
prompts are already near-distribution; the gradient through the prompt
teaches $\pi_\theta$ to \emph{generate user prompts}, which is the wrong
training signal. Second, when the SFT data is heavily synthetic with
long instruction templates (Phi-4-style pipelines run with
multi-paragraph system prompts \citep[\S 2.2]{phi4}\footnote{KB excerpt:
\texttt{kb/excerpts/phi4.md\#sec-2-2-pipelines}.}), prompt tokens
dominate the token count and dilute the response-token gradient by
sometimes an order of magnitude. Prompt-masking is equivalent to
treating $x$ as a fixed conditioning input rather than a sample to
model — the standard conditional-likelihood factorization that
\cref{eq:sft-loss-canonical} encodes implicitly and
\cref{eq:sft-loss-mask} encodes explicitly.

For multi-turn conversations $(x, y^{(1)}, x^{(2)}, y^{(2)}, \ldots,
y^{(K)})$ two mask conventions are in use \citep{tulu3}:

\paragraph{Train on all assistant turns.} $m_t = 1$ for every assistant
token across all $K$ turns; $m_t = 0$ for every user / system / tool-
result token. T\"ulu-3's default and the production-standard choice in
2026. Each assistant turn contributes loss; the model learns to
generate every assistant response in context.

\paragraph{Train on the last assistant turn only.} $m_t = 1$ only for
$y^{(K)}$. The model sees prior turns as conditioning context but does
not directly imitate them. A legacy convention that survives in some
ShareGPT-derived datasets.

The all-turns convention dominates because it gives more loss signal
per example with no obvious cost, and because the per-turn training
signal helps the model learn to maintain assistant persona across long
multi-turn dialogues. Tool-call \emph{results} are universally masked
out (the model did not generate them); tool-call \emph{invocations}
are universally included in the loss (the model did generate them, as
part of its reasoning). Conventions for system messages remain
unsettled \deepencite{tulu3 §SFT-mask, citation-pending}.

\subsection{The data lineage}
\label{sec:sft-lineage}

The interesting variation is data, not loss. Five waypoints span the
space.

\paragraph{FLAN / FLAN-T5 — instruction-tuning as terminal stage
(2021--22).} \citet{wei2021-flan} reformatted $\sim\!1.8$K NLP tasks
(classification, QA, summarization, NLI) as natural-language
instructions and fine-tuned a base LM on them with cross-entropy.
\citet{chung2022-flan-t5} extended this multi-task instruction-tuning
with multiple prompting templates and \glsterm{chain-of-thought}{chain-of-thought} traces;
\glsterm{flan}{FLAN}-\glsterm{t5}{T5} became the canonical reference for the pre-\glsterm{rlhf}{RLHF} era. SFT was
the \emph{terminal} post-training stage; no preference signal entered
the recipe.

\paragraph{InstructGPT — SFT as stage 1 of a multi-stage pipeline
(2022).} \citet[\S 3]{ouyang2022-instructgpt} re-cast SFT as the
first of three stages, with stage-2 RM training and stage-3 PPO
following \cref{eq:sft-loss-canonical}. The data is $\sim\!13$K
labeler-written prompt-plus-demonstration pairs, drawn from the
OpenAI API submission stream
\citep[\S 3.4]{ouyang2022-instructgpt}. The headline empirical fact
of the paper — that the $1.3$B aligned model is preferred over the
$175$B GPT-3 base on user prompts \citep[\S 4]{ouyang2022-instructgpt}
— is what made SFT-then-RL the dominant frame. InstructGPT also
established the convention that the small LR ($\sim\!10^{-6}$ during
PPO; $\sim\!5 \times 10^{-6}$ during SFT) and the per-token KL anchor
together hold the post-training pipeline together.

\paragraph{Llama-3 — rejection-sampling SFT, iterated.}
\citet{meta-llama3} ship a multi-round SFT loop in which each new round
is on-policy: from the current $\pi^{\mathrm{SFT}}$, sample $N \in [10,
30]$ responses per prompt, score with a \glsterm{reward-model}{reward model}, keep the top
$k \in [1, 3]$, and SFT-fine-tune on the kept responses. Iterate over
$\geq 6$ rounds, alternating with \glsterm{dpo}{DPO}. The total SFT corpus across
rounds is $\sim\!10^7$ examples — three orders of magnitude above
InstructGPT. The recipe is best read as ``on-policy SFT filtered by an
off-policy \glsterm{reward-model}{reward model}'': the model learns from its own distribution,
but only at the points where the RM judges it correct.

\paragraph{T\"ulu-3 — organic-heavy mixed sourcing.} \citet{tulu3}
publish a fully open SFT recipe at $\sim\!939$K examples, mixing four
streams: filtered organic data (Reddit Q\&A, StackExchange, \glsterm{flan}{FLAN}-style
reformatted tasks), human-written demonstrations, synthetic data from
strong teacher models, and persona-conditioned generation. The mix is
balanced explicitly across math, code, reasoning, multi-turn dialogue,
refusal, and multilingual capability, with iterative human evaluation
gating the safety-helpfulness ratio. T\"ulu-3 reports clean monotonic
gains through $\sim\!1$M samples, which sets one boundary of the
data-quantity contradiction in \cref{sec:sft-contradictions}.

\paragraph{MAGPIE — self-synthesis at zero human cost.}
\citet{magpie-2024}'s procedure is structurally trivial. Take any
aligned LM $\pi^{\mathrm{Aligned}}$ (LLaMA-3-Instruct, Qwen-Chat,
\glsterm{mistral}{Mistral}-Instruct), prompt it with \emph{only} its instruction-template
prefix (e.g.\ \texttt{<|begin\_of\_text|><|start\_header\_id|>user<|end\_header\_id|>\textbackslash n\textbackslash n})
with no question, and sample. The model fabricates a user question.
Append the instruction-template response delimiter and sample again to
get the answer. SFT on \glsterm{magpie}{MAGPIE}-generated data on Llama-3-8B-base
recovers most of the performance of Llama-3-8B-Instruct on AlpacaEval
and MT-Bench at the cost of two forward passes per example
\citep{magpie-2024}.

\begin{intuition}
What \glsterm{magpie}{MAGPIE} exposes is that an aligned model carries a
``user-question prior'' baked into its conditional distribution after
the system-prompt template: $p(\text{question} \mid
\text{template-prefix only})$ is itself a useful instruction
distribution. The base model's unconditional distribution does not
have this property — it would generate web text. Returning to math:
SFT on \glsterm{magpie}{MAGPIE} data is just \cref{eq:sft-loss-canonical} with
$\mathcal{D}_{\mathrm{SFT}}$ constructed by sampling $(x, y)$ from
$\pi^{\mathrm{Aligned}}(\cdot \mid \text{prefix})$. The aligned model is
re-used as both data source and a posteriori demonstrator
$\pi^{\mathrm{demo}}$.
\end{intuition}

\paragraph{R1-Distill and s1 — reasoning-via-SFT (2025).}
\citet[\S 4]{deepseek-r1} demonstrate that reasoning capability
acquired through RL with verifiable rewards (\cref{sec:rlvr-grpo})
transfers to small dense models through pure SFT.\footnote{KB excerpt:
\texttt{kb/excerpts/deepseek-r1.md\#sec-4}.} After training R1 by \glsterm{rlvr-3}{RLVR},
the team curated $800$K SFT samples ($600$K reasoning + $200$K
non-reasoning) by sampling from R1 and filtering for correctness;
SFT-only fine-tuning on Qwen-2.5 / Llama-3 base checkpoints (1.5B,
7B, 14B, 32B, 70B) yields the ``DeepSeek-R1-Distill'' family. The
distilled-32B matches or exceeds OpenAI's o1-mini on multiple reasoning
benchmarks, and the distilled-7B is competitive at 7B scale, all
without any RL on the small models \citep[\S 4]{deepseek-r1}.
\citet{s1-2025} push the data-scale axis to its extreme: $\sim\!1$K
hand-curated reasoning traces from Gemini 2.0 Flash Thinking, plus
inference-time \emph{budget-forcing}, reaches near-o1-mini on AIME.
The shared finding is sharp: a small set of clean, fully-worked \glsterm{chain-of-thought}{CoT}
traces is enough to elicit reasoning behavior in a strong base model,
provided the traces are correct and in the model's preferred style.
Per-example data quality dominates per-example data quantity at this
scale.

\paragraph{Phi-4 — synthetic-heavy SFT mix.} \citet[\S 1, \S
2.1]{phi4}\footnote{KB excerpts:
\texttt{kb/excerpts/phi4.md\#sec-1-pillars} and
\texttt{\#sec-2-1-purpose}.} make the strongest published case for a
synthetic-heavy mix in both pre-training and post-training. The pre-
training phase uses $\sim\!400$B tokens of synthetic data from $50$+
multi-stage prompting pipelines \citep[\S 2.2]{phi4},\footnote{KB
excerpt: \texttt{kb/excerpts/phi4.md\#sec-2-2-pipelines}.} and the
post-training SFT mix preserves the same synthetic-heavy ratio. The
underlying intuition, in Phi-4's own framing, is that synthetic data
acts as ``\glsterm{spoonfeeding}{spoonfeeding}'' \citep[\S 2.1]{phi4}: ``each token generated
by a language model is by definition predicted by the preceding
tokens, making it easier for a model to follow the resulting reasoning
patterns.'' Phi-4 surpasses its own teacher GPT-4o on GPQA ($56.1$ vs
$50.6$) and MATH ($80.4$ vs $74.6$) at $\sim\!30\times$ smaller-class
\citep[\S 1]{phi4}, evidence that the data-generation pipeline is
doing more than vanilla distillation.

\subsection{Post-SFT — rejection sampling and iteration}
\label{sec:sft-iteration}

Modern frontier recipes do not stop at one SFT pass. The Llama-3
\citep{meta-llama3} and T\"ulu-3 \citep{tulu3} loops alternate SFT
with preference-RL across multiple rounds, each round producing a new
$\pi^{\mathrm{SFT}'}$ that becomes the next round's reference. The
iteration uses \glsterm{rejection-sampling-sft}{rejection-sampling SFT} as the on-policy data refresher:

\begin{enumerate}
\item From the current $\pi^{\mathrm{SFT}}$, sample $N$ responses per
prompt ($N \in [10, 30]$).
\item Score each response with a \glsterm{reward-model}{reward model}.
\item Keep the top-$k$ ($k \in [1, 3]$) per prompt.
\item Run \cref{eq:sft-loss-mask} on the kept responses, producing
$\pi^{\mathrm{SFT}'}$.
\item Use $\pi^{\mathrm{SFT}'}$ as initialization and reference for the
next preference-RL round (\cref{sec:rlhf}, \cref{sec:dpo}).
\end{enumerate}

The iteration interleaves with the \glsterm{dpo}{DPO} step from \cref{sec:dpo},
giving the standard 2024-era recipe SFT $\to$ \glsterm{dpo}{DPO} $\to$ rejection-sample
$\to$ SFT $\to$ \glsterm{dpo}{DPO} $\to$ \ldots. The convergence criterion is
empirical (preference-win-rate plateau across rounds rather than a
fixed-point criterion on the loss).

\subsection{Why SFT before RL}
\label{sec:sft-before-rl}

\Cref{sec:rlhf,sec:dpo,sec:rlvr-grpo} all initialize the policy from
$\pi^{\mathrm{SFT}}$ and use it as the KL anchor. Why is this stage
needed at all?

The mechanical reason is that RL needs a reasonable starting policy.
A base LM produces incoherent or web-style continuations on instruction
prompts; the rollouts are mostly unscorable, the reward gradient is
uselessly sparse, and the trust-region constraint
$\KL(\pi_\theta \,\|\, \pi^{\mathrm{base}})$ is meaningless because
$\pi^{\mathrm{base}}$ does not concentrate on any ``acting like an
assistant'' manifold to anchor against. SFT cheaply produces the
distribution shift from ``language model'' to
``instruction-follower,'' after which RL refines.

The structural reason follows directly from the KL-direction duality
of the previous intuition box: SFT minimizes \emph{forward} KL to the
demonstrator, which is mode-covering and broad; \glsterm{rlhf}{RLHF} / \glsterm{dpo}{DPO} minimize
\emph{reverse} KL to $\pi^{\mathrm{SFT}}$, which is mode-seeking and
sharp. Running RL on a base LM would have $\pi^{\mathrm{SFT}}$ in
\cref{eq:rlhf-optimum} replaced by a $\pi^{\mathrm{base}}$ whose mass
is spread across web continuations rather than instruction-following
behavior; the closed-form maximizer
$\pi^*(y \mid x) \propto \pi^{\mathrm{base}}(y \mid x) \exp(r_\phi /
\beta)$ then fails to concentrate on any usable distribution at
practical $\beta$. Removing the SFT stage is feasible only when the
reward signal is strong enough on its own to substitute for the
distribution shift — DeepSeek-\glsterm{r1-zero}{R1-Zero} \citep[\S 2]{deepseek-r1}
removes the SFT cold-start in the verifiable-reward setting, but the
result has poor readability and language-mixing artifacts that
motivate the cold-start SFT stage of the production R1 pipeline
\citep[\S 3]{deepseek-r1}.

\subsection{Three live contradictions}
\label{sec:sft-contradictions}

Three open contradictions sit on the SFT data axis and propagate
through the rest of the post-training pipeline. We surface them here
and pick them up in full in \cref{sec:contradictions}.

\begin{contradiction}
\textbf{Quantity vs.\ quality.} Two empirical traditions disagree on
the SFT data scale required to elicit instruction-following.
\emph{``More is better, with quality gates''} (T\"ulu-3
\citep{tulu3}, Llama-3 \citep{meta-llama3}, Phi-4 \citep{phi4}):
scale to $10^6$+ samples with rejection-sampling and reward-model
filtering, reporting clean monotonic gains.
\emph{``Few high-quality samples suffice''}
(LIMA \deepencite{lima-2023 §results, citation-pending}; s1
\citep{s1-2025}): $1$K--$10$K carefully curated examples are enough
to elicit instruction-following from a strong base. Both traditions
are likely correct in different regimes — small-curated-sets work for
\emph{instruction style and tone}, large-sets are needed for
\emph{capability breadth} (math, code, multilingual, refusal,
safety) — but the regime boundary has not been mapped at frontier
scale.
\end{contradiction}

\begin{contradiction}
\textbf{Synthetic fraction.} What fraction of an SFT mix should be
synthetic vs.\ organic? \citet{phi4} make the strongest synthetic-
heavy case ($\sim\!80\%$ synthetic by tokens in pre-training, with
the post-training SFT mix following the same ratio). \citet{tulu3}
and \citet{meta-llama3} use organic-heavier mixes and report similar
end-task quality. No paper has run a controlled ablation at frontier
scale; the closest is the data-mixing-laws line
\citep{data-mixing-laws-2024}, which addresses pre-training mixtures,
not SFT. The question is upstream-coupled to the synthetic-vs-organic
question in \cref{sec:pretrain-data}, and the resolution of one
likely informs the other.
\end{contradiction}

\begin{contradiction}
\textbf{Reasoning-distillation ceiling.} Does pure-SFT-from-strong-
teacher saturate at the teacher's reasoning frontier, or can it
surpass through better inference-time strategies? The R1-Distill
result \citep[\S 4]{deepseek-r1} that the distilled-32B nearly matches
o1-mini on AIME is the strongest evidence yet that reasoning
capability is largely transferable through SFT, but the paper itself
notes that SFT-distilled small models do not exceed their teacher.
\citet{s1-2025} push back: $\sim\!1$K traces plus budget-forcing
appears to extract behavior that was not fully exploited at the
teacher's default inference compute, suggesting the distillation
ceiling may move with inference-time strategy. Cross-link to Paper~3
for the inference-time-compute scaling argument that organizes this
contradiction.
\end{contradiction}

\subsection{Forward link}
\label{sec:sft-forward}

\Cref{sec:sft-objective,sec:sft-mask,sec:sft-lineage,sec:sft-iteration,sec:sft-before-rl}
establish $\pi^{\mathrm{SFT}}$ as the universal starting point for the
post-training stages that follow, with the forward-KL discipline
giving a mode-covering broad policy. \Cref{sec:rlhf} now picks up the
canonical three-stage \glsterm{rlhf}{RLHF} pipeline that consumes $\pi^{\mathrm{SFT}}$
as both initialization and KL anchor: reward-model training under the
Bradley--Terry log-likelihood, PPO optimization of the KL-constrained
reward objective, and the closed-form RL optimum
$\pi^*(y \mid x) \propto \pi^{\mathrm{SFT}}(y \mid x) \exp(r_\phi /
\beta)$ that turns out — six years later — to be exactly the object
\cref{sec:dpo}'s \glsterm{dpo}{DPO} derivation inverts. The KL direction flips from
forward (SFT, mode-covering) to reverse (\glsterm{rlhf}{RLHF} / \glsterm{dpo}{DPO}, mode-seeking) at
the boundary; the mixed-precision recipe and most of the optimization
machinery (\cref{sec:mixed-precision,sec:optimization}) carry forward
unchanged. The reasoning-RL pipeline of \cref{sec:rlvr-grpo} departs
further: the learned \glsterm{reward-model}{reward model} is replaced by a programmatic
verifier, and the cold-start SFT is sometimes elided entirely
(\glsterm{r1-zero}{R1-Zero}), at the cost of the readability artifacts noted in
\cref{sec:sft-before-rl}.
