% Paper 2, §13 — Frontier-2026 snapshot: pipeline choices per stage
% Pass-C synthesis. Reads across §2-§12 and reports which choice each
% disclosed frontier model made at each of the six pipeline stages.
% Anchors: kb/index/timeline.md 2024-2026 entries;
%          kb/notes/scaling/scaling-frontier.md §5;
%          all preceding sections of paper-2.

\section{Frontier-2026 snapshot: pipeline choices per stage}
\label{sec:frontier-snapshot}

The body of this paper has fixed each of the six pipeline stages on
its own. The remaining empirical question is whether reading
\emph{across} the disclosed frontier still recovers a consistent
shape, or whether the per-stage canonical lineages decompose into
recipe-zoo bias from a handful of labs that read each other's papers.
This section answers by tabulation. We list the disclosed frontier
models of 2024--2026 against the six pipeline stages and fill each
cell with the choice the model's tech report or model card actually
reports, marking each undisclosed cell ``opaque.'' The shape that
falls out is the load-bearing claim: every disclosed pipeline since
2024 contains all six stages, in the same order, with the same
inputs and outputs at each stage boundary; agreement on the
architecture of the recipe is broader than agreement on the
load-bearing variables; and the disagreements that remain are
localized at stage boundaries (token budget, optimizer, \glsterm{tensor-parallel}{TP}-yes-vs-no,
\glsterm{fp8}{FP8}-vs-BF16-base, \glsterm{dpo}{DPO}-vs-PPO-vs-\glsterm{grpo-2}{GRPO} mix, distill-from-reasoner
yes/no, merge yes/no) rather than at the
pipeline's structure. \Cref{sec:contradictions} concentrates the
boundary-localized disagreements; this section establishes that the
structure is what we should expect them to live inside.

\subsection{The snapshot table}
\label{sec:snap-table}

The columns are the six stages, abbreviated as: \textbf{Data} (the
pre-training corpus and mixing strategy of \cref{sec:pre-training-data});
\textbf{Scale} (token budget $D$, the optimizer choice from
\cref{sec:optimization}, and the schedule); \textbf{Distrib./Prec.}
(the parallelism plan from \cref{sec:distributed-training} and the
working precision from \cref{sec:mixed-precision}); \textbf{SFT}
(scale and rejection-sampling iteration from \cref{sec:sft});
\textbf{Pref-RL} (the \glsterm{dpo}{DPO} / PPO / \glsterm{rlaif-3}{RLAIF} / \glsterm{constitutional-ai}{CAI} choice from
\cref{sec:rlhf,sec:dpo,sec:rlaif-cai}); \textbf{Reasoning-RL +
Adapt} (\glsterm{rlvr-3}{RLVR} / \glsterm{grpo-2}{GRPO} presence and merge ingredients from
\cref{sec:rlvr-grpo,sec:adaptation-and-merging}). Cells marked
\emph{opaque} are those where the lab has not publicly disclosed the
choice; cells marked with a citation give the choice exactly as the
paper or model card reports.

\begin{table}[h]
\centering
\scriptsize
\begin{tabularx}{\textwidth}{l XXXXXX}
\toprule
\textbf{Model} & \textbf{Data} & \textbf{Scale + opt.} & \textbf{Distrib./Prec.} & \textbf{SFT} & \textbf{Pref-RL} & \textbf{Reason-RL + Adapt} \\
\midrule
\glsterm{precision-pinning}{DeepSeek-V3} \citep{deepseek-v3} &
diverse $14.8$T tok; high-qual mix &
$D / N_{\text{act}} \!\approx\! 400$ MoE; \glsterm{adamw}{AdamW}; WSD-cooldown &
ZeRO-1 + \glsterm{expert-parallel}{EP}$\,16$ + \glsterm{pipeline-parallel}{PP}$\,16$ \glsterm{dualpipe}{DualPipe}, \emph{no} \glsterm{tensor-parallel}{TP}; \textbf{\glsterm{fp8}{FP8}} fine-grained &
$1.5$M instruct + reasoning &
\emph{group-relative} (\glsterm{grpo-2}{GRPO} \citep{shao2024}) directly &
\glsterm{rlvr-3}{RLVR} via \glsterm{grpo-2}{GRPO}; merging not used \\
\addlinespace[2pt]
DeepSeek-R1 \citep{deepseek-r1} &
inherits V3 base &
\glsterm{grpo-2}{GRPO} over base; cold-start SFT optional (\glsterm{r1-zero}{R1-Zero} $\to$ R1) &
inherits V3 plan &
$800$K samples for R1-Distill targets &
optional preference-\glsterm{dpo}{DPO} post-\glsterm{rlvr-3}{RLVR} &
\textbf{\glsterm{rlvr-3}{RLVR}-\glsterm{grpo-2}{GRPO}}; R1-Distill exports to dense $1.5$B--$70$B \\
\addlinespace[2pt]
Llama 3 405B / 70B / 8B \citep{meta-llama3} &
$\sim 15$T tok; classifier-filtered web &
$D/N \in \{37, 214, 1875\}$; \glsterm{adamw}{AdamW}; cosine &
\glsterm{data-parallel}{DP}+\glsterm{tensor-parallel}{TP}$\,8$+\glsterm{pipeline-parallel}{PP}+\glsterm{context-parallel}{CP}; BF16 weights, \glsterm{fp8}{FP8} activations selectively &
multi-round \emph{\glsterm{rejection-sampling-sft}{rejection-sampling SFT}} ($10$M+ samples) &
\textbf{\glsterm{dpo}{DPO}} \citep{rafailov2023-dpo} on iterated preferences &
no public \glsterm{rlvr-3}{RLVR} phase; no merge \\
\addlinespace[2pt]
Llama 4 \citep{meta-llama4} &
$\sim 30$T multimodal tok &
MoE; ``\glsterm{metap}{MetaP}'' HP transfer; \glsterm{adamw}{AdamW} &
\textbf{\glsterm{fp8}{FP8} pre-training} disclosed &
opaque scale; multimodal SFT &
\glsterm{dpo}{DPO} + on-policy preference iter. &
opaque \glsterm{rlvr-3}{RLVR}; no merge disclosed \\
\addlinespace[2pt]
Qwen 2.5 \citep{qwen2-5} &
$18$T tok; multilingual + code-heavy &
dense; \glsterm{adamw}{AdamW}; cosine &
opaque parallelism; BF16 &
SFT $+$ \glsterm{rejection-sampling}{rejection sampling} &
\glsterm{dpo}{DPO} $+$ on-policy preference iter. &
no public \glsterm{rlvr-3}{RLVR} for 2.5; no merge \\
\addlinespace[2pt]
Qwen 3 \citep{qwen3} &
$36$T tok; \emph{thinking-mode toggle} via system-prompt &
MoE + dense lineup; \glsterm{adamw}{AdamW} &
opaque parallelism; BF16 &
multi-stage SFT (think + non-think) &
\glsterm{dpo}{DPO} + \glsterm{grpo-2}{GRPO} mix &
\textbf{\glsterm{rlvr-3}{RLVR}-\glsterm{grpo-2}{GRPO}} for thinking-mode; no merge \\
\addlinespace[2pt]
Phi-4 \citep{phi4} &
$\sim\!80\%$ \textbf{synthetic} ($\sim 400$B / $50$ pipelines) &
$14$B; \glsterm{adamw}{AdamW}; \textbf{midtrain} (WSD-cooldown) &
opaque parallelism; BF16 &
synthetic-heavy SFT mirrors pretrain mix &
\glsterm{dpo}{DPO} &
no \glsterm{rlvr-3}{RLVR}; no merge \\
\addlinespace[2pt]
Tülu 3 \citep{tulu3} &
inherits Llama-3 base &
post-training only &
inherits base &
$\sim\!939$K curated multi-source &
\textbf{\glsterm{dpo}{DPO}} on iterated on-policy data &
\textbf{\glsterm{rlvr-3}{RLVR}-\glsterm{grpo-2}{GRPO}} on math/code subsets; no merge \\
\addlinespace[2pt]
OLMo 2 \citep{olmo2} &
\textbf{Dolmino} mix (mid-train cooldown data) &
$7$B / $13$B / $32$B; \glsterm{adamw}{AdamW}; WSD &
\glsterm{data-parallel}{DP}/\glsterm{fsdp}{FSDP}; BF16 &
$\sim 1$M Tülu-style mix &
\glsterm{dpo}{DPO} &
late \glsterm{rlvr-3}{RLVR} (instruct line); no merge \\
\addlinespace[2pt]
\glsterm{mistral}{Mistral} Large
\deepencite{mistral-large-2 — bibkey not in references.bib;
release notes used as source} &
opaque corpus; multilingual focus &
opaque $D / N$; \glsterm{adamw}{AdamW} &
opaque parallelism; BF16 &
opaque SFT scale &
\glsterm{dpo}{DPO} disclosed in release notes &
opaque \glsterm{rlvr-3}{RLVR}; merge \emph{used} via fine-tunes from base \\
\addlinespace[2pt]
Gemma 3 \citep{gemma3} &
opaque ($\sim 15$T-class) &
opaque optimizer specifics &
TPU; opaque parallelism; BF16 &
SFT + \glsterm{rejection-sampling}{rejection sampling} &
\textbf{PPO} with \glsterm{warm}{WARM}/\glsterm{warp}{WARP} \citep{warm-2024,warp-2024} &
no public \glsterm{rlvr-3}{RLVR}; merge via \glsterm{warm}{WARM}/\glsterm{warp}{WARP} RM ensembles \\
\addlinespace[2pt]
\glsterm{claude}{Claude} 3.7-class
\deepencite{claude-3-7-system-card — bibkey not in references.bib;
Anthropic system cards} &
opaque &
opaque &
opaque &
opaque &
\textbf{\glsterm{rlaif-3}{RLAIF} / \glsterm{constitutional-ai}{CAI}} \citep{bai2022-cai} &
opaque \glsterm{rlvr-3}{RLVR} (extended-thinking exposed in API); merge opaque \\
\addlinespace[2pt]
GPT-4o / o-series
\deepencite{openai-o-series-system-cards — bibkey not in references.bib;
OpenAI o1/o3/o4 system cards} &
opaque &
opaque &
opaque &
opaque &
\textbf{\glsterm{rlhf}{RLHF}}-class (per InstructGPT lineage \citep{ouyang2022-instructgpt}) &
o-series \textbf{trained for reasoning} (\glsterm{rlvr-3}{RLVR}-class, opaque); merge opaque \\
\addlinespace[2pt]
Gemini 2.5 \citep{gemini2-5} &
opaque (multimodal mixed) &
opaque optimizer; thinking-budget exposed &
TPU; opaque parallelism; BF16 &
opaque SFT scale &
opaque preference-RL mix &
opaque \glsterm{rlvr-3}{RLVR}; \texttt{thinkingBudget} exposes \glsterm{test-time-compute}{test-time compute} axis \\
\bottomrule
\end{tabularx}
\caption{Frontier-2026 pipeline choices per stage. Cells with citations
report the choice the paper / model card discloses; cells marked
\emph{opaque} indicate the lab has not publicly stated that choice.
$D$ is training-token count, $N$ is parameter count
($N_{\text{act}}$ for MoE active parameters), and $D / N$ is
tokens-per-active-parameter. Bold-faced cells flag the entries that
disagree across labs and motivate \cref{sec:contradictions}'s open
questions. The four columns of disclosure asymmetry are visible at a
glance: open-weights labs (DeepSeek, Meta, Alibaba, AI2, Microsoft)
disclose $D$, optimizer, parallelism, and post-training stack; closed
labs (Anthropic, OpenAI, DeepMind, partly Mistral) disclose only the
post-training family.}
\label{tab:frontier-snapshot}
\end{table}

\subsection{Where the frontier agrees}
\label{sec:snap-agree}

Reading the table column-by-column, the agreement is broader than
the variance suggests. Six axes are now near-universal across the
disclosed frontier.

\paragraph{FineWeb-class web pipeline.}
Every open-weights frontier corpus we have visibility into uses the
curate--ablate--mix loop introduced by \glsterm{custom-heuristic-filters}{FineWeb} \citep{fineweb2024}:
WARC-\glsterm{trafilatura}{trafilatura} extraction, language-ID filtering, MinHash
$14 \times 8$ deduplication \emph{per snapshot} rather than globally
\citep[\S 3.4]{fineweb2024}, quality classifier on top.
\citet{deepseek-v3} explicitly cites the \glsterm{custom-heuristic-filters}{FineWeb}-style design
\citep[\S 2]{deepseek-v3}, OLMo 2's Dolma / Dolmino lineage
\citep{olmo2} adopts an open variant, and DCLM-Baseline / RedPajama-V2
/ Nemotron-\glsterm{common-crawl}{CC} are explicit \glsterm{custom-heuristic-filters}{FineWeb}-class open releases
(\cref{sec:pre-training-data}). The per-snapshot dedup result is the
specific methodological agreement: all of these pipelines
deliberately do \emph{not} dedup globally because
\citet[\S 3.4]{fineweb2024} showed global dedup degrades downstream
quality.

\paragraph{$\sim 15$T English-web ceiling.}
Both \citet{deepseek-v3} ($14.8$T) and \citet{meta-llama3}
($\sim\!15$T) report training-token totals that converge near the
same number. The convergence is not coincidental: it is the
diminishing-returns ceiling on high-quality English web text reported
by both labs \citep[\S 4]{deepseek-v3}. Beyond the ceiling, labs
that need more tokens supplement with synthetic (Phi-4: $\sim 80\%$
synthetic by tokens \citep[\S 1]{phi4}) or multilingual (Qwen 3:
$36$T multilingual \citep{qwen3}, Llama 4 multimodal $\sim\!30$T
\citep{meta-llama4}). The ``$\sim 15$T'' point on the curate side
is the new frontier-2026 anchor.

\paragraph{AdamW + cosine-or-WSD.}
Every disclosed pretrain optimizer is \glsterm{adamw}{AdamW}
(\cref{sec:optimization}); every disclosed schedule is either
cosine-decay (Llama-3, Qwen 2.5) or \glsterm{warmup}{Warmup}--Stable--Decay
\citep[\S 3.3]{muon-moonlight2025} (\glsterm{precision-pinning}{DeepSeek-V3}, OLMo 2 Dolmino,
Phi-4 midtrain). \glsterm{per-shape-rms-scaling}{Muon} \citep{muon-moonlight2025} appears at the
$16$B-MoE Moonlight scale but no $70$B+ dense or $600$B+ MoE
deployment uses it as of 2026-Q2; the $\sim\!2\times$ compute-
efficiency lift \glsterm{per-shape-rms-scaling}{Muon} claims is unverified at frontier scale. The
schedule choice is therefore where the largest variance lives, with
WSD's cooldown phase doubling as the data-mixture-injection vehicle
that pairs with \cref{sec:adaptation-and-merging}'s continual /
late-stage pre-training (Phi-4 midtrain, OLMo 2 Dolmino, \glsterm{precision-pinning}{DeepSeek-V3}
context-extension).

\paragraph{FSDP / ZeRO-1 + BF16, FP8 increasingly.}
Every disclosed parallelism plan composes ZeRO-1 (Llama 3, \glsterm{precision-pinning}{DeepSeek-V3})
or full \glsterm{fsdp}{FSDP} (OLMo 2, Tülu 3 inherits) along the data-parallel axis
\citep[\S 2.1.2]{torchtitan2024}. Working precision is BF16 across
the board with selective \glsterm{fp8}{FP8} GEMM acceleration where the stack
supports it: \glsterm{precision-pinning}{DeepSeek-V3} ran $14.8$T tokens of \emph{pre-training}
in fine-grained \glsterm{fp8}{FP8} with zero irrecoverable loss spikes
\citep[abstract]{deepseek-v3}; Llama 4 reports \glsterm{fp8}{FP8} pre-training
\citep{meta-llama4}; Llama 3 reports \glsterm{fp8}{FP8} selectively for activations
\citep{meta-llama3}. The 2026 frontier-precision floor is now \glsterm{fp8}{FP8} on
the GEMM where the hardware supports it, BF16 working / FP32 master
elsewhere; sub-\glsterm{fp8}{FP8} (\glsterm{nvfp4}{NVFP4} / BitNet) remains unconfirmed at frontier
scale (\cref{sec:contradictions}).

\paragraph{SFT first, preference-RL second.}
The post-training pipeline order is structurally identical across
every disclosed lab: cold-start SFT \citep{ouyang2022-instructgpt}
producing $\pi^{\mathrm{SFT}}$, then a preference-RL stage anchored
on $\pi^{\mathrm{SFT}}$ via $\beta\,\KL(\pi_\theta \| \pi^{\mathrm{SFT}})$
(\cref{sec:rlhf-pipeline}). The closed-form RL optimum
$\pi^*(y \mid x) \propto \pi^{\mathrm{SFT}}(y \mid x) \exp(r / \beta)$
is the load-bearing object \emph{every} preference family inverts or
optimizes (\cref{sec:rlhf-ppo,sec:dpo-derivation}); only the source
of $r$ varies (human-\glsterm{rlhf}{RLHF} \citep{ouyang2022-instructgpt}, AI-judge
\citep{bai2022-cai}, implicit policy log-ratio \citep{rafailov2023-dpo},
verifier \citep{shao2024}).

\paragraph{RLVR-GRPO for the reasoning lines.}
Where a model has a publicly disclosed reasoning variant, \glsterm{grpo-2}{GRPO} over
$\{0, 1\}$ verifiers is the recipe. DeepSeek-R1 \citep{deepseek-r1},
Tülu 3 \citep{tulu3}, and Qwen 3 thinking-mode \citep{qwen3} all
report \glsterm{grpo-2}{GRPO} with binary verifiers (sympy-equivalence, unit-tests,
exact-match) plus a small \glsterm{format-reward}{format reward}
(\cref{sec:rlvr-grpo-objective}). \glsterm{dapo-2}{DAPO}'s four targeted improvements
\citep{dapo2025} (\glsterm{clip-higher}{Clip-Higher}, \glsterm{dynamic-sampling}{Dynamic Sampling}, Token-Level Loss,
Overlong Reward Shaping) appear in subsequent open-source replications.
The closed-lab reasoning models (o-series, \glsterm{claude}{Claude} extended-thinking,
Gemini 2.5) do not disclose the inner-loop algorithm but expose a
\emph{thinking-budget} or \emph{reasoning-effort} parameter that is
behaviorally consistent with an \glsterm{rlvr-3}{RLVR}-trained reasoner under
test-time-compute scaling (cross-link to Paper~3).

\subsection{Where the frontier diverges}
\label{sec:snap-disagree}

Six axes show real disagreement, each pinned in the preceding
sections and concentrated in \cref{sec:contradictions}.

\paragraph{Token budget $D / N$.}
The disclosed split runs from Chinchilla's $D / N \approx 20$
\citep[\S 4.1]{hoffmann2022-chinchilla} to Llama-3-8B's
$D / N \approx 1875$ \citep{meta-llama3} --- two orders of magnitude
on the same axis. The split is now \emph{deployment-conditional}
(\cref{tab:frontier-NDC}): small dense models destined for billions
of inference queries over-train hard; large MoE models intended for
quality-sensitive serving sit closer to Chinchilla. Phi-4 at
$\sim\!400$B synthetic / $14$B \glsterm{parameters}{parameters} and \glsterm{precision-pinning}{DeepSeek-V3} at
$D / N_{\text{act}} \approx 400$ are both deliberate and both legal;
the open question is the joint $(C_{\text{train}}, C_{\text{test}}, b, q)$
\glsterm{scaling-law}{scaling law} \cref{sec:scaling-map} flagged.

\paragraph{Pure-Muon vs.\ AdamW.}
Only the Moonlight $16$B-MoE / $5.7$T tokens run \citep{muon-moonlight2025}
has reported a pure-\glsterm{per-shape-rms-scaling}{Muon} pre-training at frontier-class scale; the
disclosed $70$B+ dense and $600$B+ MoE runs are all \glsterm{adamw}{AdamW}. Whether
the $\sim 2\times$ compute-efficiency lift \glsterm{per-shape-rms-scaling}{Muon} advertises holds at
$70$B / dense or $600$B / MoE is the original paper's own §1 open
question and remains unanswered.

\paragraph{TP-yes-vs-TP-no.}
Llama 3 \citep{meta-llama3} runs \glsterm{tensor-parallel}{TP}$\,8$ inside-node plus \glsterm{pipeline-parallel}{PP} plus \glsterm{data-parallel}{DP} plus
\glsterm{context-parallel}{CP} --- heavy multi-axis parallelism. \glsterm{precision-pinning}{DeepSeek-V3} \citep{deepseek-v3}
runs ZeRO-1 + \glsterm{expert-parallel}{EP}$\,16$ + \glsterm{pipeline-parallel}{PP}$\,16$ \glsterm{dualpipe}{DualPipe} with \emph{no} \glsterm{tensor-parallel}{TP}, and
explicitly argues that fine-grained \glsterm{fp8}{FP8} plus \glsterm{dualpipe}{DualPipe} and the
cross-node all-to-all kernel relieves enough activation pressure that
\glsterm{tensor-parallel}{TP} becomes unnecessary \citep[\S 3.2.2]{deepseek-v3}. The two
recipes are architecture-conditional, not generally substitutable;
\cref{sec:dt-contradictions} treats the disagreement.

\paragraph{FP8 base vs.\ BF16 base.}
\glsterm{precision-pinning}{DeepSeek-V3} \citep[\S 3.3]{deepseek-v3} and Llama 4
\citep{meta-llama4} report \glsterm{fp8}{FP8} \emph{base} pre-training. Llama 3 uses
\glsterm{fp8}{FP8} selectively (activations-only on \glsterm{fp8}{FP8} GEMMs; BF16 weights and
master-FP32) \citep{meta-llama3}. OLMo 2, Phi-4, Qwen 2.5, Gemma 3
remain BF16-base. The \glsterm{precision-pinning}{DeepSeek-V3} zero-spike $14.8$T-token result
\citep[abstract]{deepseek-v3} is the strongest single data point we
have for frontier-scale \glsterm{fp8}{FP8} training, but it is single-source until
Llama 4 publishes a comparable spike-rate figure
(\cref{sec:contradictions}).

\paragraph{SFT scale.}
The disclosed range is $\sim\!1$K (s1 \citep{s1-2025}) to $\sim 10$M
(Llama 3 multi-round rejection-sampling \citep{meta-llama3}) --- four
orders of magnitude. The two camps coexist (\cref{sec:sft-contradictions}):
small-curated for instruction style and tone (LIMA, s1); large-scale
for capability breadth (Llama 3, Tülu 3, Phi-4). Phi-4's $\sim\!80\%$
synthetic-fraction \citep[\S 1]{phi4} is itself a load-bearing
divergence vs.\ Tülu 3's organic-heavier mix \citep{tulu3} that no
controlled frontier-scale ablation has yet resolved.

\paragraph{DPO vs.\ PPO vs.\ GRPO mix; RLAIF / CAI presence.}
Llama 3 \citep{meta-llama3}, Tülu 3 \citep{tulu3}, OLMo 2
\citep{olmo2}, Qwen 2.5 \citep{qwen2-5}, and the Phi-4 line
\citep{phi4} all ship \glsterm{dpo}{DPO} \citep{rafailov2023-dpo} or its variants
on the preference-RL leg. DeepMind's Gemma 3 \citep{gemma3} is the
load-bearing exception: PPO with weight-averaged reward models
(\glsterm{warm}{WARM} \citep{warm-2024} / \glsterm{warp}{WARP} \citep{warp-2024}) is the disclosed
choice. Anthropic's \glsterm{claude}{Claude} line is the second exception: \glsterm{rlaif-3}{RLAIF} / \glsterm{constitutional-ai}{CAI}
\citep{bai2022-cai} (now C3AI \citep{c3ai-2025} / \glsterm{r-cai-2}{R-CAI}
\citep{r-cai-2026}) replaces the preference source itself with an
AI-judge-under-a-\glsterm{constitution}{constitution}. DeepSeek-R1 takes a third path:
direct \glsterm{grpo-2}{GRPO} over the SFT base \citep{deepseek-r1} with optional
preference-\glsterm{dpo}{DPO} \emph{after} the \glsterm{rlvr-3}{RLVR} phase. The \glsterm{dpo}{DPO}-vs-PPO question at
frontier scale is the \cref{sec:dpo-contradictions} contradiction;
neither has produced a controlled head-to-head at $70$B+.

\paragraph{R1-distill yes/no; merge yes/no.}
DeepSeek-R1 \citep{deepseek-r1} reports $800$K SFT samples generated
by the \glsterm{rlvr-3}{RLVR}-trained R1, distilled into dense $1.5$B / $7$B / $32$B /
$70$B targets that nearly match o1-mini on AIME at the $32$B size;
this is the strongest published evidence that reasoning capability
transfers through pure SFT \citep[\S 4]{deepseek-r1}. The other
disclosed labs do not publicly report a parallel distill-from-reasoner
SFT pass on a publicly-available reasoning teacher. On the merge axis,
Gemma 3's \glsterm{warm}{WARM}/\glsterm{warp}{WARP} \citep{warm-2024,warp-2024} is the only disclosed
production merge as a load-bearing pipeline component
(\cref{sec:adaptation-and-merging}); \glsterm{mistral}{Mistral} and the open-weights
community use \glsterm{mergekit}{MergeKit} \citep{mergekit2024} extensively post-release
but as a community add-on rather than as a stage of the canonical
pipeline.

\subsection{The disclosure asymmetry}
\label{sec:snap-opaque}

The opaque cells are not random. \Cref{tab:frontier-snapshot} shows
the closed labs (Anthropic, OpenAI, DeepMind in part, partly \glsterm{mistral}{Mistral})
disclose only the post-training \emph{family} --- \glsterm{rlaif-3}{RLAIF} / \glsterm{constitutional-ai}{CAI} for
Anthropic, \glsterm{rlhf}{RLHF}-class for OpenAI, PPO+\glsterm{warm}{WARM}/\glsterm{warp}{WARP} for DeepMind --- and
keep the data, scale, optimizer, parallelism, precision, and SFT
details opaque. The open-weights labs (DeepSeek, Meta, Alibaba, AI2,
Microsoft Research) disclose the upstream stages by paper or model
card and treat the post-training stack as a separate publication.
This asymmetry is structural: the open-weights labs have a research
incentive to disclose the recipe (it is the citation-bearing object)
while the closed labs publish system cards that document the
\emph{interface} (capabilities, refusal behavior, thinking-budget
exposure) without disclosing the optimizer or parallelism choice.
For \textmu P at frontier scale (\cref{sec:scaling-mup}), the
disclosure asymmetry is one of \cref{sec:contradictions}'s open
items: independent positive results outside the
Microsoft / Cerebras / Graphcore ecosystem at $10$B+ scale are
scarce, and the Llama 4 \citep{meta-llama4} ``\glsterm{metap}{MetaP}'' citation has
thin published detail.

\begin{contradiction}
\textbf{Six-stage canonical or recipe-zoo bias?} The opening of
\cref{sec:intro-six} flagged this question. The
\cref{tab:frontier-snapshot} answer is that every disclosed pipeline
since 2024 has all six stages in the same order, and where stages
are missing it is by design (Tülu 3 has no pre-training because it
inherits Llama 3's; \glsterm{mistral}{Mistral} Large does not publicly disclose
\glsterm{rlvr-3}{RLVR}; Phi-4 does not run \glsterm{rlvr-3}{RLVR} by choice). What the table cannot
distinguish is whether the convergence is genuine pipeline-shape
discovery or whether it is recipe-zoo bias: the disclosed labs
read each other's papers, hire from each other, and run on overlapping
hardware. The two readings are not yet separable, and the question
sharpens further at the closed-lab boundary (Anthropic, OpenAI,
DeepMind have an unknown number of stages we cannot see). The
\emph{stage-shape} convergence is empirically robust where we can
see; the \emph{stage-shape canonicality} is open. The
\cref{sec:contradictions} reading is that the convergence
\emph{plus} the boundary-localized disagreements both point to a
pipeline whose structure is stable but whose load-bearing variables
have not converged.
\end{contradiction}

\begin{intuition}
A useful mental picture is that the table is a $14 \times 6$ matrix
of cells where the column-marginals (\cref{sec:snap-agree}) tell us
the structural shape and the row-marginals (\cref{sec:snap-disagree})
tell us each lab's identity. Columns with high agreement
(SFT-first, BF16+ working precision, \glsterm{adamw}{AdamW} pretrain optimizer,
$\KL$-anchored preference-RL) are the structure of the recipe.
Columns with high variance (token budget $D/N$, \glsterm{fp8}{FP8}-vs-BF16-base,
\glsterm{dpo}{DPO}-vs-PPO mix, \glsterm{rlvr-3}{RLVR} yes/no, merge yes/no)
are the surface where the next round of empirics will settle. The
opaque cells in the closed-lab rows tell us where labs treat the
choice as competitive advantage rather than research artifact ---
the post-training data and reward sources for OpenAI / Anthropic;
the parallelism plan and precision for DeepMind. Returning to math:
each cell's \glsterm{value}{value} is, in the language of the
\cref{sec:intro-buy} intuition box, a choice of $(\pi^{\mathrm{ref}}, r,
\mathcal{D})$ for that stage's KL-constrained reweighting. The
table is the lab-by-lab realization of the same recipe template;
the disagreements are which point in the template's parameter
space the lab chose.
\end{intuition}

\subsection{Forward link}
\label{sec:snap-forward}

\Cref{tab:frontier-snapshot} establishes the empirical shape:
six stages, in order, with localized disagreement at five of the six
boundaries (data, scale + optimizer, distrib. + precision,
preference-RL family, reasoning-RL + merge) and structural
near-universal agreement on the remaining axes (\glsterm{custom-heuristic-filters}{FineWeb}-class web
pipeline, \glsterm{adamw}{AdamW}, \glsterm{fsdp}{FSDP}/ZeRO-1, BF16+ working precision, SFT-first /
preference-RL-second / \glsterm{rlvr-3}{RLVR}-for-reasoning ordering,
KL-anchored closed-form preference optimum). The next section
\cref{sec:contradictions} concentrates the boundary-localized
disagreements and asks, for each, what evidence would close the
question: the Kaplan-vs-Chinchilla compute allocation under the
inference-cost re-objective, optimizer speedups conditional on
training-loss vs.\ downstream evaluation, \glsterm{tensor-parallel}{TP}-yes-vs-\glsterm{tensor-parallel}{TP}-no
under different MoE \emph{vs.\ }dense architectures, \glsterm{fp8}{FP8} / \glsterm{nvfp4}{NVFP4} /
BitNet production status outside the \glsterm{precision-pinning}{DeepSeek-V3} single-source
result, SFT data quantity-vs-quality across regime boundaries,
reward-hacking as fundamental or tractable, \glsterm{dpo}{DPO}-vs-PPO at frontier
scale, \glsterm{rlaif-3}{RLAIF} / \glsterm{constitutional-ai}{CAI} as internalization or surface mimicry, \glsterm{rlvr-3}{RLVR} as
new-capability or sampling-efficient sharpening, merge guarantees
under linear-mode-connectivity, the precision $\times$ token-count
interaction the over-training strategy collides with, and \textmu P
strict-or-loose at frontier scale. The closing thesis we will
defend in \cref{sec:contradictions} is anticipated by the
\cref{tab:frontier-snapshot} reading itself: the pipeline's open
questions are \emph{localized} at stage boundaries, not at the
pipeline's structure.
