% Paper 2, §14 — Contradictions live and open frontiers
% Pass-C synthesis. Concentrates the [CONTRADICTION] markers from §2-§13
% into one explicit treatment. Each contradiction: what each side claims,
% what evidence would close it. Closing thesis: the pipeline's open
% questions are localized at stage boundaries, not at pipeline structure.
% Anchors: kb/index/contradictions.md (training, post-training, scaling
% sections); the §2-§13 contradiction blocks themselves.

\section{Contradictions live and open frontiers}
\label{sec:contradictions}
\label{sec:contradictions-frontier}

The previous twelve sections each ended with
\texttt{[CONTRADICTION]} blocks where the literature does \emph{not}
yet converge. This section concentrates them. For each of twelve
live disagreements we name what each side claims, why the
disagreement is real, and what evidence would close it. The closing
thesis is structural: the open questions are \emph{localized} at
stage boundaries (precision~$\times$~optimizer;
token-count~$\times$~quantization; preference~$\times$~verifier),
not at the pipeline's six-stage structure, which
\cref{sec:frontier-snapshot} shows has converged.

\subsection{Pre-training, scaling, optimization}
\label{sec:contradictions-pretrain-scale-opt}

\paragraph{(1) Kaplan-vs-Chinchilla compute allocation.}
On the methodological question this is settled.
\citet[\S 6]{kaplan2020} fit $N \propto C^{0.73}$, $D \propto C^{0.27}$
under a cosine-LR schedule decayed to a fixed token horizon, which
biased shorter-data runs upward;
\citet[\S 3, Table 2]{hoffmann2022-chinchilla} re-fit with
LR-horizon matching across three independent approaches
(training-curves, IsoFLOP, parametric) and recover the
$\approx 0.50/0.50$ split with the canonical $D \approx 20 N$ at
compute-optimal pre-training. The contradiction that remains is the
``\glsterm{chinchilla-trap}{Chinchilla trap}'' framing: frontier industry over-trains far past
the Chinchilla optimum (Llama-3-8B at $D / N \approx 1875$
\citep{meta-llama3}, two orders of magnitude past
$D / N \approx 20$). This is not a refutation. It is a different
objective --- minimize \emph{total lifetime cost} (pre-training plus
amortized inference) rather than pre-training-FLOP-optimal loss
\citep[\S \ref{sec:scaling-map}]{hoffmann2022-chinchilla}\protect\footnote{KB
excerpt: \texttt{kb/excerpts/hoffmann2022-chinchilla.md\#sec-3}.}. What
\emph{is} open is the joint
$(C_{\text{train}}, C_{\text{test}}, b, q)$ \glsterm{scaling-law}{scaling law}, where
$C_{\text{test}}$ is inference-time compute (Paper~3) and $b, q$ are
precision and mixture quality: documented empirically, not in closed
form. Closing evidence would be a multi-axis IsoFLOP sweep at
$10$B-$70$B that exposes the trade-off surface; no public lab has
reported one.

\paragraph{(2) Optimizer speedups are training-loss-conditional.}
\glsterm{lion}{Lion} (sign-momentum, Chen et al.\ 2023) and \glsterm{sophia}{Sophia} (diagonal-Hessian
preconditioning, Liu, Li et al.\ 2023) reported $\sim$1.5--$2\times$
training-loss speedups at \glsterm{llm}{LLM}-relevant scales. \glsterm{adamw}{AdamW} closes or
reverses the gap on downstream evaluation panels (\glsterm{mmlu}{MMLU}, GSM8K,
HumanEval, BBH) at matched wall-clock time
\citep[\S 1]{muon-moonlight2025}\protect\footnote{KB excerpt:
\texttt{kb/excerpts/\glsterm{muon}{muon}-moonlight2025.md\#sec-1-challenges}.}.
\glsterm{per-shape-rms-scaling}{Muon}-Moonlight \citep{muon-moonlight2025} fits
$L_{\text{Muon}} = 2.506\,C^{-0.052}$ vs.\
$L_{\text{AdamW}} = 2.608\,C^{-0.054}$ for a constant matched-loss
compute ratio $\eta \approx 0.52$, and the downstream panel at the
$16$B-MoE / $5.7$T-token Moonlight point is competitive with similarly
sized \glsterm{precision-pinning}{DeepSeek-V3} variants. The disagreement is whether \glsterm{per-shape-rms-scaling}{Muon}'s lift
generalizes beyond a single tech report. As of 2026-Q2 the result is
\emph{single-source} on a $16$B-MoE run; whether the $\sim$$2\times$
constant holds at $70$B+ dense and $600$B+ MoE is the paper's own
\S 1 open question. Closing evidence would be one independent
third-party replication at frontier dense scale plus a downstream
panel matched on tokens and wall-clock --- the structural lesson
(re-verify training-loss claims on downstream surface) is the
operational rule until then.

\paragraph{(11) Precision $\times$ token-count interaction.}
\citet{scaling-laws-precision-2024} extend the Chinchilla form with
precision as a third axis and predict, via an effective parameter
count $N_{\text{eff}}(N, b_{\text{train}})$, that
\emph{over-trained models are harder to quantize}: the more $D$ a
model has consumed relative to $N$, the larger the post-quantization
loss penalty at deployment.\deepencite{scaling-laws-precision-2024 \S
4: effective-parameter form $N_{\text{eff}}$ and over-training
$\to$ harder-to-quantize prediction; needed at frontier scale.} This
collides directly with the inference-cost-aware over-training
strategy of contradiction (1): the same Llama-3-8B that gains from
being small-and-token-heavy is, by the Kumar prediction, more
difficult to quantize at deployment, partially offsetting the
strategy's gain. The industry response has been native low-precision
training: pre-train at the precision you will deploy, not above. The
\glsterm{precision-pinning}{DeepSeek-V3} \citep[abstract]{deepseek-v3} zero-spike $14.8$T-token
\glsterm{fp8}{FP8} result is the strongest production data point. The full
$(D / N) \times b$ surface at frontier scale is incompletely mapped;
the specific Kumar prediction is \emph{fitted on small models} and
extrapolated. Closing evidence: a controlled
$(D / N \in \{20, 200, 2000\}) \times (b \in \{\text{BF16, FP8,
FP4}\})$ sweep at $7$B--$70$B with both training stability and
post-quantization downstream loss reported.

\paragraph{(12) \textmu P at frontier scale.}
\citet{yang2022-mup} prove, under Definition 4.1's strict scaling
rules, that the master HPs (LR, betas, schedule shape) transfer
losslessly across width. Microsoft (Phi-4 \citep{phi4}), Cerebras,
and Graphcore (u-\textmu P, \citealp{u-mup2024}) disclose strict-or-
near-strict use; Meta (Llama 4 \citep{meta-llama4}) cites a
``\glsterm{metap}{MetaP}'' scheme for cross-scale HP transfer with thin published
detail; Anthropic, OpenAI, Google do not disclose. Independent
positive results outside the
Microsoft/Cerebras/Graphcore ecosystem at $10$B+ scale are scarce.
The community working assumption is that frontier labs apply the
\textmu P LR scalings while keeping Vaswani's $1/\sqrt{d_k}$
attention rather than \textmu P's $1/d$, making them
``\textmu P-inspired'' rather than strict-Definition-4.1 variants.
Whether the loose variants retain the unique-\glsterm{feature}{feature}-learning
guarantee is open. Closing evidence: an open-weights frontier model
disclosing strict Definition-4.1 \textmu P with width-sweep HP
transfer evidence at $\geq 70$B \glsterm{parameters}{parameters}.

\subsection{Distributed training and precision}
\label{sec:contradictions-dist-prec}

\paragraph{(3) TP-yes-vs-TP-no.}
\citet{meta-llama3} train Llama 3 405B with heavy \glsterm{tensor-parallel}{TP}+\glsterm{pipeline-parallel}{PP}+\glsterm{data-parallel}{DP} --- the
prototypical Megatron-3D shape. \citet{deepseek-v3} train
\glsterm{precision-pinning}{DeepSeek-V3} 671B-A37B \emph{without \glsterm{tensor-parallel}{TP} at all}, claiming the
combination of ZeRO-1 \glsterm{data-parallel}{DP} $+$ \glsterm{expert-parallel}{EP} $+$ \glsterm{dualpipe}{DualPipe} $+$ \glsterm{fp8}{FP8} supplies enough
activation relief to make \glsterm{tensor-parallel}{TP}'s per-linear collective cost a net
loss. This is a real architectural-conditional disagreement, not a
wash. \glsterm{tensor-parallel}{TP} costs $2L$ inter-GPU all-reduces per step; \glsterm{expert-parallel}{EP} costs
$2 L_{\text{MoE}}$ inter-node all-to-alls per step. DeepSeek's
fine-grained MoE forces them to pay the \glsterm{expert-parallel}{EP} all-to-all already
\citep[\S 3.2]{deepseek-v3}, and pairing it with \glsterm{dualpipe}{DualPipe} overlap
plus the $4$-node-cap kernel hides the cost. Llama 3 is dense at
active-parameter level; without an \glsterm{expert-parallel}{EP} all-to-all to fold \glsterm{tensor-parallel}{TP} work
into, the Megatron-3D plan stays preferable. The general rule is
\emph{which} parallelism is best is a function of the model graph
(MoE vs.\ dense, expert count, expert size, attention head count) ---
not a universal answer. Closing evidence is a plan-from-graph
optimizer that picks $(d, t, c, p, e)$ from $(N, D, \text{cluster},
\text{graph})$; \citet[\S 1]{torchtitan2024} explicitly punt on this,
GSPMD propagates rather than searches, and AlpaDiscovery is not
dominant in 2026 open practice.

\paragraph{(4) FP8 / NVFP4 / BitNet production status.}
Three precision regimes, three different statuses. \emph{\glsterm{fp8}{FP8}}: the
\glsterm{precision-pinning}{DeepSeek-V3} zero-spike $14.8$T-token result \citep[abstract]{deepseek-v3}
is the strongest production data point on record; Llama 4
\citep{meta-llama4} likewise discloses \glsterm{fp8}{FP8} pre-training. The
contradiction is that the \glsterm{fp8}{FP8} result remains \emph{single-source} on
its specific recipe (per-token-per-128-channel activation tile,
per-128$\times$128 weight block, all-E4M3, $N_C = 128$ promotion)
and no other major lab has publicly matched the envelope. \emph{FP4
/ \glsterm{nvfp4}{NVFP4}}: Blackwell hardware ships FP4 Tensor Cores with MXScale,
several Blackwell-deploying vendors are running FP4 pilots, but as
of 2026-Q2 no major lab has publicly confirmed FP4 \emph{pre-training}
of a frontier model. The Kumar effective-parameter prediction warns
of substantial loss below $\sim$$7$ bits and is fitted on small
models. \emph{BitNet 1.58 ternary}: \citet{ma2024-bitnet} claim
ternary $\{-1, 0, 1\}$ matches FP16 quality, with the BitNet b1.58
2B4T tech report \citep{bitnet-2b4t-2025} as a follow-up, but
independent reproductions thin past $\sim$$7$B \glsterm{parameters}{parameters} and no
frontier-scale companion has appeared. Treat \glsterm{fp8}{FP8} as
production-stable contingent on recipe transfer; treat sub-\glsterm{fp8}{FP8} as
open. Closing evidence: a non-DeepSeek frontier-scale \glsterm{fp8}{FP8}
zero-spike replication; a frontier-scale FP4 pre-training
disclosure; a $\geq$$30$B BitNet replication outside the original
authors.

\subsection{Post-training: SFT, preference-RL, reasoning-RL}
\label{sec:contradictions-post}

\paragraph{(5) SFT data quantity-vs-quality.}
Two empirical traditions. \emph{``Few high-quality samples
suffice''}:
LIMA\deepencite{lima-2023 §results --- bibkey not yet in
references.bib} and s1 \citep{s1-2025} report $1$K--$10$K curated
examples eliciting instruction-following from a strong base.
\emph{``Scale with quality gates''}: T\"ulu-3 \citep{tulu3}, Llama-3
\citep{meta-llama3}, Phi-4 \citep{phi4} run $10^6$+-sample SFT mixes
with rejection-sampling and reward-model filtering, reporting clean
monotonic gains. Both are likely correct in different regimes:
small-curated for \emph{instruction style and tone}, large-set for
\emph{capability breadth} (math, code, multilingual, refusal,
safety). The regime boundary has not been mapped at frontier scale.
Closing evidence: a controlled
$(\text{data scale}) \times (\text{evaluation axis})$ sweep on a
fixed strong base, separating style/tone from capability.

\paragraph{(6) Reward-hacking is fundamental or tractable.}
\emph{Tractable view} (Coste 2023; \citealp{warm-2024,warp-2024};
Gemma 3 production \citep{gemma3}): RM ensembles, weight-averaged
RMs, KL scheduling, and on-policy preference iteration mitigate
\glsterm{reward-hacking}{reward hacking} enough for production. \emph{Fundamental view} (the
direct-alignment scaling-laws line\deepencite{rosset2024-direct-alignment-scaling
arXiv:2406.02900 \S 3 --- bibkey not yet in references.bib} and
the persistence of \glsterm{sycophancy}{sycophancy} as a localized shortcut at all scales,
\citealp{sharma2023-sycophancy}): proxy-reward optimization produces
overoptimization across \emph{all} preference-based post-training,
including \glsterm{dpo}{DPO}, since \glsterm{dpo}{DPO}'s \glsterm{implicit-reward}{implicit reward}
$\hat r_\theta = \beta \log \pi_\theta / \pi^{\mathrm{ref}}$ is itself
a learned proxy that can be hacked. Closing evidence: a controlled
RM-quality $\times$ KL-budget $\times$ pipeline (PPO/\glsterm{dpo}{DPO}/\glsterm{grpo-2}{GRPO}) sweep
with both proxy reward and gold reward measured, plus an adversarial
\glsterm{sycophancy}{sycophancy}-\glsterm{probe-2}{probe} panel.

\paragraph{(7) DPO-vs-PPO at frontier scale.}
\citet{rafailov2023-dpo} establishes that \glsterm{dpo}{DPO} and PPO-\glsterm{rlhf}{RLHF} share
the \emph{same} fixed point on the same preference data, with \glsterm{dpo}{DPO}
reaching it without sampling, RM, or RL loop. The empirical
disagreement is whether the shared fixed point is reached equally
well in practice. \emph{Llama-3 / T\"ulu-3 view}
\citep{meta-llama3,tulu3}: with iterated on-policy preference
collection, \glsterm{dpo}{DPO} matches or exceeds PPO on standard alignment
benchmarks at lower compute. \emph{DeepMind / Gemma-3 view}
\citep{gemma3,warm-2024,warp-2024}: PPO + \glsterm{warm}{WARM}/\glsterm{warp}{WARP}-averaged reward
ensembles remains the production choice for general alignment, the
ensembled RM recovering the proxy-vs-gold gap that single-RM PPO
suffered. The disagreement is partly recipe-specific (different RMs,
preference datasets, KL schedules) and partly task-specific (\glsterm{dpo}{DPO}
more competitive on instruction-following; PPO-with-ensemble retains
advantage on creative / open-ended generation per anecdotal reports).
No controlled head-to-head at frontier scale has been published. The
\glsterm{simpo}{SimPO} length-bias fix \citep{meng2024-simpo} adjusts loss form
without resolving the head-to-head. Closing evidence: a
fixed-base $\times$ fixed-preference-data sweep over
$\{\text{DPO}, \text{PPO+WARM}\}$ $\times$ KL schedule $\times$
on-policy iteration depth, with held-out alignment $+$ creative panels.

\paragraph{(8) RLAIF / CAI: internalization or surface mimicry; CAI vs.\ sycophancy.}
\citet{bai2022-cai} establish \glsterm{constitutional-ai}{CAI} as ``\glsterm{rlhf}{RLHF} with a different
labeler'': AI-judge under randomly-chosen principle $c_k$ replaces
human preferences, BT-RM trained on AI labels feeds PPO with KL
anchor to $\pi^{\text{SL-CAI}}$. \glsterm{constitutional-ai}{CAI}'s behavioral signature is
\glsterm{non-evasion}{non-evasion}. Two open splits. First, internalization vs.\ surface
mimicry: \emph{genuine principle internalization} and \emph{learned
text shaped like constitutional reasoning} are behaviorally
indistinguishable on standard evaluations; only adversarial probes
where the principle's intent and its surface verbalization disagree
can separate them. Second, \glsterm{constitutional-ai}{CAI} vs.\ \glsterm{sycophancy}{sycophancy}: the AI judge rewards
principle-\emph{aligned} responses, so the policy can learn to
\emph{appear} principle-aligned to the judge whether the underlying
reasoning is shallow or deep --- \glsterm{sycophancy}{sycophancy} with respect to the AI
judge instead of the human labeler \citep{sharma2023-sycophancy}.
\glsterm{constitutional-ai}{CAI} may not solve \glsterm{sycophancy}{sycophancy} and may amplify it, since the
algorithmic judge is more consistent than human labelers and the
policy can find its biases more reliably. Closing evidence:
held-out adversarial \glsterm{probe-2}{probe} panels (intent-vs-verbalization,
principle-conflict, \glsterm{sycophancy}{sycophancy} with counter-evidence) reported
jointly for SFT-only, \glsterm{rlhf}{RLHF}, \glsterm{rlaif-3}{RLAIF}, and \glsterm{constitutional-ai}{CAI} baselines on a fixed base.

\begin{contradiction}
\textbf{The proxy-reward thread.} Contradictions (6), (7), and (8)
are not independent. They are three readings of one underlying
question: \emph{is post-training fundamentally a proxy-reward
optimization, and if so, does the proxy-vs-gold gap close or persist
as the proxy improves?} The fundamental view answers ``persists ---
the gap is a structural property of any learned reward signal,
including \glsterm{dpo}{DPO}'s \glsterm{implicit-reward}{implicit reward} and \glsterm{constitutional-ai}{CAI}'s AI judge.'' The tractable
view answers ``closes to acceptable levels --- ensembling,
KL anchoring, on-policy iteration, and judge-under-\glsterm{constitution}{constitution}
together push the residual below the deployment threshold.'' The
empirical signature that would distinguish them is whether the gap
\emph{shrinks} or \emph{stabilizes} as proxy quality is varied along
a controlled axis (RM ensemble size, judge capability, KL budget).
We do not yet have the controlled axis-sweep at frontier scale. The
\cref{sec:contradictions-post} contradictions are individually
tractable; the underlying question --- is reward-hacking fundamental
or engineering --- is not.
\end{contradiction}

\paragraph{(9) Does RLVR create new reasoning capability?}
\citet{deepseek-r1} frame \glsterm{rlvr-3}{RLVR} as ``incentivizing reasoning
capability'' (\glsterm{r1-zero}{R1-Zero} from $15.6\%$ to $77.9\%$ on AIME 2024 over
\glsterm{grpo-2}{GRPO} training, the ``aha-moment'' \texttt{wait}-token spike), with
parallel framings in \citet{tulu3} and \citet{qwen3} treating \glsterm{rlvr-3}{RLVR}
as eliciting reasoning behaviors the base model could not produce.
\citet{rlvr-limits-2025} (Yue et al.\ NeurIPS 2025 oral)%
\deepencite{rlvr-limits-2025 §3-§5, KB-excerpt-pending}
disagree:
by carefully comparing pass@1 vs.\ pass@$K$ of base-model vs.\
\glsterm{rlvr-3}{RLVR}'d model at matched sampling budget, \glsterm{rlvr-3}{RLVR} improves pass@1
(sampling efficiency) on problems the base model can already solve
at higher $K$, but does \emph{not} expand the set of problems
solvable at any $K$. The \glsterm{capability-ceiling}{capability ceiling} is set by pre-training;
\glsterm{rlvr-3}{RLVR} is sampling-efficient sharpening. As of 2026-Q2 the open-source
consensus leans toward Yue with caveats: the result is benchmark-,
temperature-, and sample-budget-conditional, and generalization to
non-math (agentic, long-horizon) \glsterm{rlvr-3}{RLVR} is single-source. Closing
evidence is concentrated in Paper~3, which treats \glsterm{rlvr-3}{RLVR} alongside
inference-time-compute scaling and search; the
$(C_{\text{train,RLVR}}) \times (C_{\text{test}})$ joint surface is
where the framings might reconcile.

\subsection{Adaptation and merging}
\label{sec:contradictions-merge}

\paragraph{(10) Model-merging guarantees.}
The merge literature reports two patterns that do not yet reconcile.
On controlled multi-task benchmarks, merged models often exceed any
single ingredient (\citealp[\S 4]{wortsman2022-model-soups};
\citealp[\S 4]{yadav2023-ties-merging}). On real-world deployments at
the open-weights \glsterm{llm}{LLM} merging community's frontier scale, merged
models routinely under-perform the strongest specialist on its own
task, with merge wins concentrated in multi-task generalization
rather than peak-task performance. The per-paper benchmarks are
largely single-source: TIES and Soups sweep their own task suites
with no third-party head-to-head at frontier-\glsterm{llm}{LLM} scale. The
speculative theoretical underpinning is linear-mode-connectivity ---
the basin-sharing hypothesis that fine-tuned models from a common
base lie in a connected low-loss region whose center is the average
\citep{ilharco2022-task-vectors,wortsman2022-model-soups}. \glsterm{linear-mode-connectivity-2}{LMC} holds
empirically for SFT-from-shared-base, partly for \glsterm{rlhf}{RLHF}, and breaks
for from-scratch different-init runs; no canonical theory predicts
when merging produces interference vs.\ basin-centering. The
\emph{merged-beats-best-ingredient} claim should be read as
\emph{task-set-conditional}: it holds on multi-task evaluations of
specialists, and is not yet established at the single-task ceiling.
Closing evidence: a third-party head-to-head at $\geq$$30$B with
matched training data, isolating sign-conflict resolution
(TIES \citep{yadav2023-ties-merging}) from drop-and-rescale (\glsterm{dare}{DARE})
from greedy-soup ingredient selection.

\subsection{Closing thesis: structure stable, boundaries open}
\label{sec:contradictions-thesis}

The twelve contradictions above admit a structural reading. None of
them disputes the \emph{shape} of the pipeline. Every disclosed
2024--2026 frontier pipeline (\cref{sec:frontier-snapshot}) executes
the same six stages in the same order: pre-training-data $\to$
scaling-and-optimization $\to$ distributed-and-precision
scaffolding $\to$ SFT $\to$ preference-RL $\to$ reasoning-RL /
adaptation. What the contradictions dispute is what each stage
should do internally and how stages should be \emph{coupled at their
boundaries}.

\begin{intuition}
The contradictions cluster at three boundary couplings:
\textbf{precision~$\times$~optimizer} (\glsterm{per-shape-rms-scaling}{Muon} update RMS exceeds BF16
high-precision range without weight decay; \glsterm{fp8}{FP8} stability requires
the per-tile scaling \glsterm{precision-pinning}{DeepSeek-V3} introduced; (2)$\cap$(4) live here),
\textbf{token-count~$\times$~quantization} (over-trained models are
harder to quantize, but inference economics rewards
over-training; (1)$\cap$(11) live here), and
\textbf{preference~$\times$~verifier} (proxy reward from a learned
judge can be hacked; verifier reward is gold but sparse and
domain-bound; (6)$\cap$(7)$\cap$(8)$\cap$(9) live here). Returning
to canonical form: the pipeline factorizes structurally as a chain of
stages $S_1 \to S_2 \to \cdots \to S_6$, but the load-bearing
\emph{interaction terms} $\partial S_i / \partial S_j$ for adjacent
stages have not been characterized. The contradictions are reading
those missing interaction terms.
\end{intuition}

This is the closing claim of Paper~2. The shape of frontier
post-2024 \glsterm{llm}{LLM} training is stable: six stages, canonical lineages,
broadly disclosed at the architecture level. The 2027-class results
that close the open contradictions will not be about pipeline shape
but about load-bearing variables \emph{inside} each stage and at the
boundaries \emph{between} adjacent stages: $D / N$ under joint
inference-cost and quantization constraints; optimizer choice under
joint downstream and frontier-stability constraints; \glsterm{tensor-parallel}{TP}-yes-vs-no
under joint model-graph and topology constraints; \glsterm{dpo}{DPO}-vs-PPO under
joint proxy-reward and on-policy constraints; \glsterm{rlvr-3}{RLVR}'s
sampling-efficiency-vs-capability split under joint pre-training
and inference-time-compute budgets. Where Paper~1 established the
architectural axes and Paper~3 will establish the
inference-time-compute axes, Paper~2's claim is that the
\emph{training} axes are now legible and the open boundaries between
them are localized, named, and falsifiable. The pipeline is a real
object; we now know which of its joints flex.
